---
title: "Ladakh_Microbe_explore"
author: "Adam Ruka"
date: '2025-02-13'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

if (!file.exists("Glacier Transects.Rproj") && file.exists("../../Glacier Transects.Rproj")) setwd("../..")
if (!file.exists("Glacier Transects.Rproj") && file.exists("../Glacier Transects.Rproj")) setwd("..")
stopifnot("Open Glacier Transects.Rproj first" = file.exists("Glacier Transects.Rproj"))
project_root <- normalizePath(getwd(), winslash = "/", mustWork = TRUE)
molecular_dir <- file.path(project_root, "Molecular")

library(NetCoMi)
library(phyloseq)
library(extrafont) # Tools for using fonts, CRAN v0.17
library(kableExtra) # Construct Complex Table with 'kable' and Pipe Syntax
library(tidyverse) # Easily Install and Load the 'Tidyverse', CRAN v1.3.0
library(scales) # Generic plot scaling methods
library(cowplot) # wrappers for ggplot
library(magrittr) # A Forward-Pipe Operator for R, CRAN v2.0.1
library(phyloseq) # Handling and analysis of high-throughput microbiome census data, Bioconductor v1.32.0
#library(speedyseq) # Faster implementations of phyloseq functions, [github::mikemc/speedyseq] v0.5.3.9018 
library(Biostrings) # Efficient manipulation of biological strings, Bioconductor v2.56.0
library(svglite) # An 'SVG' Graphics Device, CRAN v1.2.3.2 
library(vegan) # Community Ecology Package, CRAN v2.5-7
library(see) # Visualisation Toolbox for 'easystats' and Extra Geoms, Themes and Color Palettes for 'ggplot2', CRAN v0.6.1 
set.seed(1) # for reproducible stochastic processes
library(ggplot2)
library(patchwork) # for combining multiple plots
library(microViz)
library(dplyr)
library(microbiome)
library(tidyverse)
library(readxl)
library(readr)
library(dplyr)
library(writexl)
library(openxlsx)
library(ggrepel)
library(raster)
library(terra)
library(dismo)
library(sp)
library(VennDiagram)
library(ggvenn)
library(lme4)
library(car)
library(emmeans)
library(ARTool)
library(multcompView)
library(rcompanion)
library(svglite)
```

## Basic analyses of alpha and beta diversity of bacterial and fungal communities from glacier transect study.
# minimum library size of 1000 for fungi and bacteria
# prevalence threshold of 1% (3 samples)


##Load Phyloseq object

```{r echo=FALSE}
La_Gla_ITS <- readRDS(file.path(molecular_dir, "Ladakh_Glacier_ITS_tax_filt.Rds"))

#tax_fix_interactive(La_Gla_ITS)
# Calculate library size
lib_size <- sample_sums(La_Gla_ITS)

# Extract sample metadata
metadata <- sample_data(La_Gla_ITS)

# Convert to a data frame (if needed)
metadata <- as.data.frame(metadata)

# Add library size column
metadata$Library_Size <- lib_size

# Reassign updated metadata to the phyloseq object
sample_data(La_Gla_ITS) <- sample_data(metadata)

sample_data(La_Gla_ITS)

tax_table(La_Gla_ITS)

ITS_meta_var <- sample_variables(La_Gla_ITS)

La_Gla_16S <- readRDS(file.path(molecular_dir, "Ladakh_Glacier_16S_tax_filt.Rds"))

#tax_fix_interactive(La_Gla_16S)

# Calculate library size
lib_size <- sample_sums(La_Gla_16S)

# Extract sample metadata
metadata <- sample_data(La_Gla_16S)

# Convert to a data frame (if needed)
metadata <- as.data.frame(metadata)

# Add library size column
metadata$Library_Size <- lib_size

# Reassign updated metadata to the phyloseq object
sample_data(La_Gla_16S) <- sample_data(metadata)


sample_data(La_Gla_16S)

tax_table(La_Gla_16S)

Bac_meta_var <- sample_variables(La_Gla_16S)

match(ITS_meta_var, Bac_meta_var)

#include snow in ITS metadata
meta <-sample_data(La_Gla_ITS)
meta$rownames <- rownames(meta)

meta <- as.matrix(meta)
meta <- as.data.frame(meta)

plot_meta <- read_excel(file.path(project_root, "Glacier_Soil_Topo_merged.xlsx"))

snow <- plot_meta %>% 
  dplyr::select(Plot, Snow_1976, Moraine_x_ID, Moraine_num) %>%
  mutate(Snow_1976 = as.numeric(Snow_1976))

# Merge the two data frames by the "Plot" column
meta <- meta %>%
  mutate(Plot = as.character(Plot)) %>%
  left_join(snow, by = "Plot")

Meta_ITS <- meta

rownames(meta) <- meta$rownames

sample_data(La_Gla_ITS) <- meta


#include snow in 16S metadata
meta <-sample_data(La_Gla_16S)
meta$rownames <- rownames(meta)

meta <- as.matrix(meta)
meta <- as.data.frame(meta)

plot_meta <- read_excel(file.path(project_root, "Glacier_Soil_Topo_merged.xlsx"))

snow <- plot_meta %>% 
  dplyr::select(Plot, Snow_1976, Moraine_x_ID, Moraine_num) %>%
  mutate(Snow_1976 = as.numeric(Snow_1976))

# Merge the two data frames by the "Plot" column
meta <- meta %>%
  mutate(Plot = as.character(Plot)) %>%
  left_join(snow, by = "Plot")

Meta_16S <- meta

rownames(meta) <- meta$rownames

sample_data(La_Gla_16S) <- meta


```



## Removing controls and generating unique identifies for "Unclassified" taxa.

```{r, min library size, echo=FALSE }
La_Gla_clean_16S <- La_Gla_16S %>%
  ps_filter(Control == "FALSE", .keep_all_taxa = TRUE) %>%
  ps_dedupe(vars = "Sample.Name")
  
# Identify taxa that are NOT unclassified at the Phylum level
keep_taxa <- !tax_table(La_Gla_clean_16S)[, "Phylum"] %in% c("Unclassified", "NA", "")

# Subset the phyloseq object to keep only classified taxa
La_Gla_clean_16S <- prune_taxa(keep_taxa, La_Gla_clean_16S)

# Identify taxa that are NOT unclassified at the Phylum level
keep_taxa <- !tax_table(La_Gla_clean_16S)[, "Kingdom"] %in% c("Unclassified", "Archaea", "")

# Subset the phyloseq object to keep only classified taxa
La_Gla_clean_16S <- prune_taxa(keep_taxa, La_Gla_clean_16S)

# Number of taxa before filtering
ntaxa(La_Gla_16S)

# Number of taxa after filtering
ntaxa(La_Gla_clean_16S)

min_prevalence <- 0.01 # 10% of samples

n_samples <- nsamples(La_Gla_clean_16S)
prevalence_threshold <- min_prevalence * n_samples  
# Convert to sample count



# Create a logical table of presence/absence (1 = present, 0 = absent)
taxa_presence <- otu_table(La_Gla_clean_16S) > 0

# Count number of samples where each taxon is present
taxa_prevalence <- rowSums(taxa_presence)


# Identify taxa that meet the prevalence threshold
keep_taxa <- taxa_names(La_Gla_clean_16S)[taxa_prevalence >= prevalence_threshold]

# Subset the phyloseq object
La_Gla_clean_16S_prev_filt <- prune_taxa(keep_taxa, La_Gla_clean_16S)


# Number of taxa before filtering
ntaxa(La_Gla_clean_16S)

# Number of taxa after filtering
ntaxa(La_Gla_clean_16S_prev_filt)

#minimum library threshold
min_threshold <- 1000

# Identify samples that meet the threshold
keep_samples <- sample_sums(La_Gla_clean_16S_prev_filt) >= min_threshold

La_Gla_clean_16S_filt <- prune_samples(keep_samples, La_Gla_clean_16S_prev_filt)

La_Gla_16S_genus <- tax_glom(La_Gla_clean_16S_filt, taxrank = "Genus")

#tax_fix_interactive(La_Gla_16S_genus)

taxa<- as.data.frame(tax_table(La_Gla_16S))

#tax_fix_interactive(La_Gla_16S_genus)

# La_Gla_16S_sp_renamed <- renameTaxa(La_Gla_clean_16S_filt,
#                                   pat = "<name>",
#                                   substPat = "<name>_<subst_name>(<subst_R>)",
#                                   numDupli = "Species")

La_Gla_16S_genus_renamed <- renameTaxa(La_Gla_16S_genus,
                                  pat = "<name>",
                                  substPat = "<name>_<subst_name>(<subst_R>)",
                                  numDupli = "Genus")
                                  
print( "Number of samples before filtering (min lib size):")
nsamples(La_Gla_clean_16S_prev_filt)

print( "Number of samples after filtering:")
nsamples(La_Gla_clean_16S_filt)

#total number of sequences
print( "Total number of sequences after filtering")
sum(otu_table(La_Gla_clean_16S_filt))

#calculate average library size
avg_lib_size <- mean(sample_sums(La_Gla_clean_16S_filt))
print(paste("Average library size after filtering:", avg_lib_size))

saveRDS((La_Gla_clean_16S_filt), file = "La_Gla_final_16S.Rds")



#tax_fix_interactive(La_Gla_clean_16S_filt)

#ITS
La_Gla_clean_ITS <- La_Gla_ITS %>%
  ps_filter(Control == "FALSE", .keep_all_taxa = TRUE) %>%
  ps_dedupe(vars = "Sample.Name")
  
  
n_samples <- nsamples(La_Gla_clean_ITS)
prevalence_threshold <- min_prevalence * n_samples  
# Convert to sample count



# Create a logical table of presence/absence (1 = present, 0 = absent)
taxa_presence <- otu_table(La_Gla_clean_ITS) > 0

# Count number of samples where each taxon is present
taxa_prevalence <- rowSums(taxa_presence)


# Identify taxa that meet the prevalence threshold
keep_taxa <- taxa_names(La_Gla_clean_ITS)[taxa_prevalence >= prevalence_threshold]

# Subset the phyloseq object
La_Gla_clean_ITS_prev_filt <- prune_taxa(keep_taxa, La_Gla_clean_ITS)


# Number of taxa before filtering
ntaxa(La_Gla_clean_ITS)

# Number of taxa after filtering
ntaxa(La_Gla_clean_ITS_prev_filt)  
  
  
  # Identify samples that meet the threshold
keep_samples <- sample_sums(La_Gla_clean_ITS_prev_filt) >= min_threshold

La_Gla_clean_ITS_filt <- prune_samples(keep_samples, La_Gla_clean_ITS_prev_filt)


# Number of samples before filtering
nsamples(La_Gla_clean_ITS_prev_filt)

# Number of samples after filtering
nsamples(La_Gla_clean_ITS_filt)

#total number of sequences
print( "Total number of sequences after filtering")
sum(otu_table(La_Gla_clean_ITS_filt))

#calculate average library size
avg_lib_size <- mean(sample_sums(La_Gla_clean_ITS_filt))
print(paste("Average library size after filtering:", avg_lib_size))

La_Gla_ITS_genus <- tax_glom(La_Gla_clean_ITS_filt, taxrank = "Genus")

#tax_fix_interactive(La_Gla_ITS_genus)

La_Gla_ITS_genus_renamed <- renameTaxa(La_Gla_ITS_genus,
                                  pat = "<name>",
                                  substPat = "<name>_<subst_name>(<subst_R>)",
                                  numDupli = "Genus")

saveRDS((La_Gla_ITS_genus), file = "La_Gla_final_ITS.Rds")

#tax_fix_interactive(La_Gla_ITS_genus)
```


## All samples

```{r, all samples, echo=FALSE}
sample_data(La_Gla_16S_genus_renamed)$Sample.Type <- factor(
  sample_data(La_Gla_16S_genus_renamed)$Sample.Type,
  levels = c("Lichen", "Soil Crust", "Moss", "Bulk Soil", "Plant")
)

sample_data(La_Gla_ITS_genus_renamed)$Sample.Type <- factor(
  sample_data(La_Gla_ITS_genus_renamed)$Sample.Type,
  levels = c("Lichen", "Soil Crust", "Moss", "Bulk Soil", "Plant")
)

bac_meta <- as.matrix(sample_data(La_Gla_16S_genus_renamed))
bac_meta <- as.data.frame(bac_meta)

La_Gla_16S_genus_renamed_pseudo <- La_Gla_16S_genus_renamed  
  
# 1. Extract OTU table, add pseudocount
otu_with_pseudo <- otu_table(La_Gla_16S_genus_renamed) + 1e-6

# 2. Replace the OTU table in the phyloseq object
otu_table(La_Gla_16S_genus_renamed_pseudo) <- otu_with_pseudo

La_Gla_16S_genus_renamed_pseudo %>%
 tax_transform(rank = "Phylum", trans = "clr") %>%
 dist_calc(dist = "euclidean") %>%
 ord_calc(
  method = "auto"
 ) %>% 
 ord_plot(
  axes = c(1, 2),
  colour = "Sample.Type", fill = "Sample.Type",
  shape = "Moraine", alpha = 0.5,
  size = 3,
  plot_taxa = 1:10,
  )+
 scale_shape_girafe_filled()+
  ggtitle(
    "PCA of Bacterial Phyla from All Samples",
  )




  
 #  La_Gla_16S_genus_renamed %>%
 # tax_transform(rank = "Genus", trans = "clr") %>%
 # # dist_calc(dist = "euclidean") %>%
 # ord_calc(
 #  method = "auto"
 # ) %>% 
 # ord_plot(
 #  axes = c(1, 2),
 #  colour = "Sample.Type", fill = "Sample.Type",
 #  shape = "Moraine", alpha = 0.5,
 #  size = 3,
 #  plot_taxa = 1:10,
 #  )+
 # scale_shape_girafe_filled()+
 #  ggtitle(
 #    "PCA of Bacterial Genera from All Samples",
 #  )
  
  La_Gla_16S_genus_renamed_pseudo %>%
 tax_transform(rank = "Genus", trans = "clr") %>%
    dist_calc(dist = "euclidean") %>%
 ord_calc(
  method = "auto"
 ) %>% 
 ord_plot(
  axes = c(1, 2),
  colour = "Sample.Type", fill = "Sample.Type",
  shape = "Moraine", alpha = 0.5,
  size = 2
 ) + 
 scale_shape_girafe_filled() +
 ggplot2::stat_ellipse(
  ggplot2::aes(colour = Sample.Type)) + 
    ggtitle(
    "PCA of Bacterial Genera from All Samples",
 )
  
  
P <- La_Gla_16S_genus_renamed_pseudo %>%
  tax_transform(rank = "Genus", trans = "clr") %>%
  dist_calc(dist = "euclidean") %>%
  ord_calc(method = "auto") %>%
  ord_plot(
    axes = c(1, 2),
    colour = "Sample.Type",    # outline color (will be overwritten by scale_colour_manual)
    fill = "Sample.Type",      # fill color
    shape = "Moraine",
    alpha = 0.8,               # fill opacity
    size = 2
    
  ) +
  scale_shape_girafe_filled() +
  ggplot2::stat_ellipse(ggplot2::aes(colour = Sample.Type)) +
  scale_colour_manual(   # outline color
    values = c(
      "Lichen" = "#FF7F0E",
      "Bulk Soil" = "#59310E",
      "Plant" = "#1DA20E",
      "Moss" = "#A3A500",
      "Soil Crust" = "#155A8A"
    )
  ) +
  scale_fill_manual(     # fill color
    values = c(
      "Lichen" = "#FF7F0E",
      "Bulk Soil" = "#59310E",
      "Plant" = "#1DA20E",
      "Moss" = "#A3A500",
      "Soil Crust" = "#155A8A"
    )
  ) +
  ggtitle("PCoA of Bacterial Genera from All Samples_Final_pseudo") +
  theme_minimal()
 
 P
 
 # Save as SVG
ggsave("La_Gla_16S_genus_renamed_16S_PCoA.svg", P,device = "svg", width = 8, height = 8)
  

# # Extract ordination object
# ord <- La_Gla_16S_genus_renamed_pseudo %>%
#   tax_transform(rank = "Genus", trans = "clr") %>%
#   dist_calc(dist = "euclidean") %>%
#   ord_calc(method = "auto")
# 
# # Create the plot manually
# P <- ord_plot(ord, axes = c(1, 2)) +
#   geom_point(
#     aes(
#       x = Axis.1, y = Axis.2,
#       fill = Sample.Type,
#       shape = Moraine
#     ),
#     data = sample_data(ord),
#     color = "black",       # outline color
#     stroke = 0.8,          # outline thickness
#     alpha = 0.8,           # fill opacity
#     size = 3
#   ) +
#   stat_ellipse(aes(x = Axis.1, y = Axis.2, colour = Sample.Type),
#                data = sample_data(ord)) +
#   scale_shape_girafe_filled() +
#   scale_fill_manual(values = c(
#     "Lichen" = "#FF7F0E",
#     "Bulk Soil" = "#59310E",
#     "Plant" = "#1DA20E",
#     "Moss" = "#A3A500",
#     "Soil Crust" = "#155A8A"
#   )) +
#   scale_colour_manual(values = c(
#     "Lichen" = "#FF7F0E",
#     "Bulk Soil" = "#59310E",
#     "Plant" = "#1DA20E",
#     "Moss" = "#A3A500",
#     "Soil Crust" = "#155A8A"
#   )) +
#   ggtitle("PCA of Bacterial Genera from All Samples_Final_pseudo_testingOutlines") +
#   theme_minimal()
# 
# # Save as SVG
# ggsave("La_Gla_16S_genus_renamed_PCoA_testingOutlines.svg", P,device = "svg", width = 8, height = 8)




    
 # La_Gla_16S_genus_renamed %>% perm_get()

# library(car)
# vif(lm(DistanceFromGlacier ~ Sample.Type + GlacierID + Location + Moraine_num + Library_Size, data = Meta_16S))


# 3. Now proceed with your CLR + PERMANOVA
La_Gla_16S_genus_renamed_perm <- La_Gla_16S_genus_renamed %>%
  tax_transform("clr", rank = "Genus") %>%
  dist_calc("euclidean") %>%
  dist_permanova(
    variables = c("Sample.Type", "Library_Size"),
    n_perms = 999,
    #strata = bac_meta$Plot,
    n_processes = 1
  )  

print(La_Gla_16S_genus_renamed_perm)
  
  
print("      permanova with rarefaction at genus level, with aitchinson distance")
La_Gla_16S_genus_renamed_perm <- La_Gla_16S_genus_renamed_pseudo %>%
  tax_transform("clr", rank = "Genus") %>%
  dist_calc("euclidean") %>%
  dist_permanova(
    variables = c("Sample.Type", "Moraine", "Location"),
    n_perms = 999, # you should use more permutations in your real analyses!
    n_processes = 1
  )

La_Gla_16S_genus_renamed_perm



# ANOVA (if assumptions are met)
#aov_model <- aov(Shannon ~ Sample.Type * Moraine, data = your_data)

# OR Kruskal-Wallis (non-parametric)
#kruskal_model <- kruskal.test(Shannon ~ interaction(Sample.Type, Moraine), data = your_data)


#plot shannon for 16S
p <- plot_richness(La_Gla_16S_genus_renamed, x = "Sample.Type", color = "Sample.Type", measures = c("Shannon"))

# Remove default points layer
p$layers <- p$layers[-1]

# Add boxplot, faded points, and facet by Moraine
p <- p + 
  geom_boxplot(size = 1) + 
  geom_jitter(position = position_jitter(width = 0.2), size = 2, alpha = 0.5) + 
  facet_wrap(~Moraine) + 
  theme_modern()+
  xlab(NULL)+
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )+  ylab("Shannon Diversity Index - 16S")

# Save as SVG
ggsave("La_Gla_16S_genus_renamed_16S_Shannon.svg", p,device = "svg", width = 8, height = 3)

#ADD TO METADATA TABLE

# 1. Calculate richness (e.g., Shannon)
shannon_df <- estimate_richness(La_Gla_16S_genus_renamed, measures = "Shannon")

# 2. Check that rownames match sample names
head(rownames(shannon_df))
head(sample_names(La_Gla_16S_genus_renamed))

# 3. Convert sample_data to data frame if needed
sample_df <- as.matrix(sample_data(La_Gla_16S_genus_renamed))
sample_df <- as.data.frame(sample_df)

# 4. Combine richness values with sample_data
sample_df$Bac_Shannon <- shannon_df$Shannon

# 5. Update sample_data in the phyloseq object
sample_data(La_Gla_16S_genus_renamed) <- sample_data(sample_df)






#plot chao richness for 16s
p <- plot_richness(La_Gla_16S_genus_renamed, x = "Sample.Type", color = "Sample.Type", measures = c("Chao1"))

# Remove default points layer
p$layers <- p$layers[-1]

# Add boxplot, faded points, and facet by Moraine
p <- p + 
  geom_boxplot(size = 1) + 
  geom_jitter(position = position_jitter(width = 0.2), size = 2, alpha = 0.5) + 
  facet_wrap(~Moraine) + 
  theme_modern()+
  xlab(NULL)+
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )+ ylab("Chao Richness - 16S")

# Save as SVG
ggsave("La_Gla_16S_genus_renamed_16S_Chao.svg", p,device = "svg", width = 8, height = 3)

#ADD TO METADATA TABLE
# 1. Calculate richness (e.g., Chao1)
chao_df <- estimate_richness(La_Gla_16S_genus_renamed, measures = "Chao1")

# 2. Check that rownames match sample names
head(rownames(chao_df))
head(sample_names(La_Gla_16S_genus_renamed))

# 3. Convert sample_data to data frame if needed
sample_df <- as.matrix(sample_data(La_Gla_16S_genus_renamed))
sample_df <- as.data.frame(sample_df)

# 4. Combine richness values with sample_data
sample_df$Bac_Chao <- chao_df$Chao1

# 5. Update sample_data in the phyloseq object
sample_data(La_Gla_16S_genus_renamed) <- sample_data(sample_df)

# Load required libraries
library(phyloseq)
library(microbiome)   # for CLR transformation
library(ggplot2)
library(dplyr)

# Normalize to relative abundances (proportions), then add pseudocount
otu_rel <- transform_sample_counts(La_Gla_16S_genus_renamed, function(x) x / sum(x))

# Add pseudocount and CLR-transform
otu_clr <- otu_rel %>%
  transform_sample_counts(function(x) x + 0.0001) %>%  # Small pseudocount for compositional data
  microbiome::transform("clr")


# Step 2: Extract the CLR-transformed OTU table
otu_mat <- as.data.frame(otu_table(otu_clr))
if (taxa_are_rows(otu_clr)) {
  otu_mat <- t(otu_mat)
}

# Step 3: Perform PCA
pca_res <- prcomp(otu_mat, scale = TRUE)
pca_df <- as.data.frame(pca_res$x)
pca_df$SampleID <- rownames(pca_df)

# Step 4: Extract and attach sample metadata
meta_df <- as.matrix(sample_data(otu_clr))
meta_df <- as.data.frame(meta_df)
meta_df$SampleID <- rownames(meta_df)
pca_df <- left_join(pca_df, meta_df, by = "SampleID")

# Step 5: Plot PCA with 95% confidence ellipses by Sample.Type
ggplot(pca_df, aes(x = PC1, y = PC2, color = Sample.Type, shape = Moraine)) +
  #stat_ellipse(aes(group = Sample.Type), level = 0.95, linetype = "dashed") +
  geom_point(size = 3) +
  theme_minimal() +
  labs(
    title = "PCA of CLR-transformed OTU Table",
    x = paste0("PC1 (", round(summary(pca_res)$importance[2, 1] * 100, 1), "%)"),
    y = paste0("PC2 (", round(summary(pca_res)$importance[2, 2] * 100, 1), "%)")
  )


#Extract OTU table
Bac_otu_pseudo <- as.matrix(otu_table(La_Gla_16S_genus_renamed_pseudo)) 
Bac_otu_pseudo <- as.data.frame(Bac_otu_pseudo)

#CLR transformation
Bac_otu_clr <- Bac_otu_pseudo %>%
  microbiome::transform("clr")
Bac_otu_clr <- as.data.frame(Bac_otu_clr)

class(Bac_otu_clr)

Bac_meta <- as.matrix(bac_meta)
Bac_meta <- as.data.frame(Bac_meta)

class(Bac_meta)

#convert to numeric and factor
Bac_meta$Moraine <- as.factor(Bac_meta$Moraine)
Bac_meta$Sample.Type <- as.factor(Bac_meta$Sample.Type)
Bac_meta$Location <- as.factor(Bac_meta$Location)
Bac_meta$Plot <- as.factor(Bac_meta$Plot)

result <- adonis2(Bac_otu_clr ~  Moraine + Sample.Type + Location, data = Bac_meta, method="euclidean", strata = Bac_meta$Plot ,perm=999)
print(result)




#run betadisperse
# Get Sample.Type grouping variable
group <- sample_data(La_Gla_16S_genus_renamed_pseudo)$Sample.Type

# Calculate Bray-Curtis dissimilarity
euc_dist <- vegdist(Bac_otu_clr, method = "euclidean")


# Run beta dispersion
disper <- betadisper(euc_dist, group)

# Test for significant differences in dispersion
anova(disper)

plot(disper)
boxplot(disper, xlab = "Sample Type", ylab = "Distance to Centroid")



permutest(disper, permutations = 999)


#mixed effect models

Bac_meta <- as.matrix(sample_data(La_Gla_16S_genus_renamed))
Bac_meta <- as.data.frame(Bac_meta)

Bac_meta$Bac_Shannon <- as.numeric(Bac_meta$Bac_Shannon)
Bac_meta$Moraine_num <- as.numeric(Bac_meta$Moraine_num)
Bac_meta$Sample.Type <- as.factor(Bac_meta$Sample.Type)
Bac_meta$Location <- as.factor(Bac_meta$Location)
Bac_meta$Plot <- as.factor(Bac_meta$Plot)
Bac_meta$Bac_Chao <- as.numeric(Bac_meta$Bac_Chao)


mixed_effect_model <- lmer(Bac_Shannon ~ Moraine * Sample.Type * Location  + (1|Plot), data = Bac_meta)

# Perform ANOVA using the car package
anova_result <- Anova(mixed_effect_model, test="F")

# Display the ANOVA table
print(anova_result)

#subset first moraine
Bac_meta_first_moraine <- Bac_meta[Bac_meta$Moraine_num == 1, ]

# Fit the ART model (fixed effect: Treatment, random effect: Plot)
art_model <- art(Bac_Shannon ~  Sample.Type , data = Bac_meta_first_moraine)


anova_art <- anova(art_model)
print(anova_art)


# Extract the aligned model (for Treatment)
art_emm <- emmeans(artlm(art_model, "Sample.Type"), pairwise ~ Sample.Type, adjust = "fdr")

# See pairwise comparisons
summary(art_emm$contrasts)

pw_df <- summary(art_emm$contrasts)

letters_df <- cldList(p.value ~ contrast, data = pw_df, threshold = 0.05)

#now for second moraine
Bac_meta_second_moraine <- Bac_meta[Bac_meta$Moraine_num == 2, ]

# Fit the ART model (fixed effect: Treatment, random effect: Plot)
art_model_second <- art(Bac_Shannon ~  Sample.Type , data = Bac_meta_second_moraine)

anova_art_second <- anova(art_model_second)
print(anova_art_second)

# Extract the aligned model (for Treatment)
art_emm_second <- emmeans(artlm(art_model_second, "Sample.Type"), pairwise ~ Sample.Type, adjust = "fdr")

# See pairwise comparisons
summary(art_emm_second$contrasts)
pw_df_second <- summary(art_emm_second$contrasts)
letters_df_second <- cldList(p.value ~ contrast, data = pw_df_second, threshold = 0.05)

#third moraine now
Bac_meta_third_moraine <- Bac_meta[Bac_meta$Moraine_num == 3, ]

# Fit the ART model (fixed effect: Treatment, random effect: Plot)
art_model_third <- art(Bac_Shannon ~  Sample.Type , data = Bac_meta_third_moraine)
anova_art_third <- anova(art_model_third)

print(anova_art_third)

# Extract the aligned model (for Treatment)
art_emm_third <- emmeans(artlm(art_model_third, "Sample.Type"), pairwise ~ Sample.Type, adjust = "fdr")
# See pairwise comparisons
summary(art_emm_third$contrasts)
pw_df_third <- summary(art_emm_third$contrasts)
letters_df_third <- cldList(p.value ~ contrast, data = pw_df_third, threshold = 0.05)


#now for richness
mixed_effect_model <- lmer(Bac_Chao ~ Moraine * Sample.Type * Location  + (1|Plot), data = Bac_meta)

# Perform ANOVA using the car package
anova_result <- Anova(mixed_effect_model, test="F")

# Display the ANOVA table
print(anova_result)


#first moraine w chao
Bac_meta_first_moraine <- Bac_meta[Bac_meta$Moraine_num == 1, ]
# Fit the ART model (fixed effect: Treatment, random effect: Plot)
art_model_chao <- art(Bac_Chao ~  Sample.Type , data = Bac_meta_first_moraine)
anova_art_chao <- anova(art_model_chao)
print(anova_art_chao)

# Extract the aligned model (for Treatment)
art_emm_chao <- emmeans(artlm(art_model_chao, "Sample.Type"), pairwise ~ Sample.Type, adjust = "fdr")
# See pairwise comparisons
summary(art_emm_chao$contrasts)
pw_df_chao <- summary(art_emm_chao$contrasts)
letters_df_chao <- cldList(p.value ~ contrast, data = pw_df_chao, threshold = 0.05)


#second moraine w chao
Bac_meta_second_moraine <- Bac_meta[Bac_meta$Moraine_num == 2, ]
# Fit the ART model (fixed effect: Treatment, random effect: Plot)
art_model_chao_second <- art(Bac_Chao ~  Sample.Type , data = Bac_meta_second_moraine)
anova_art_chao_second <- anova(art_model_chao_second)
print(anova_art_chao_second)

# Extract the aligned model (for Treatment)
art_emm_chao_second <- emmeans(artlm(art_model_chao_second, "Sample.Type"), pairwise ~ Sample.Type, adjust = "fdr")
# See pairwise comparisons
summary(art_emm_chao_second$contrasts)
pw_df_chao_second <- summary(art_emm_chao_second$contrasts)
letters_df_chao_second <- cldList(p.value ~ contrast, data = pw_df_chao_second, threshold = 0.05)


#third moraine
Bac_meta_third_moraine <- Bac_meta[Bac_meta$Moraine_num == 3, ]
# Fit the ART model (fixed effect: Treatment, random effect: Plot)
art_model_chao_third <- art(Bac_Chao ~  Sample.Type , data = Bac_meta_third_moraine)
anova_art_chao_third <- anova(art_model_chao_third)
print(anova_art_chao_third)

# Extract the aligned model (for Treatment)
art_emm_chao_third <- emmeans(artlm(art_model_chao_third, "Sample.Type"), pairwise ~ Sample.Type, adjust = "fdr")
# See pairwise comparisons
summary(art_emm_chao_third$contrasts)
pw_df_chao_third <- summary(art_emm_chao_third$contrasts)
letters_df_chao_third <- cldList(p.value ~ contrast, data = pw_df_chao_third, threshold = 0.05)



#beta dispersion for I6S, separated by moraine

#extract betadispersion values
# Get Sample.Type grouping variable
samp_group <- sample_data(La_Gla_16S_genus_renamed_pseudo)$Sample.Type

moraine_group <- sample_data(La_Gla_16S_genus_renamed_pseudo)$Moraine

#create interaction term
group <- interaction(samp_group, moraine_group)

# Calculate Bray-Curtis dissimilarity
euc_dist <- vegdist(Bac_otu_clr, method = "euclidean")


# Run beta dispersion
disper <- betadisper(euc_dist, group)

#extract values from disper
Bac_meta$betaDist <- disper$distances

#now for richness
mixed_effect_model <- lmer(betaDist ~ Moraine * Sample.Type * Location  + (1|Plot), data = Bac_meta)

# Perform ANOVA using the car package
anova_result <- Anova(mixed_effect_model, test="F")

# Display the ANOVA table
print(anova_result)


# Create the boxplot
p_beta <- ggplot(Bac_meta, aes(x = Sample.Type, y = betaDist, color = Sample.Type)) +
  geom_boxplot(size = 1) +
  geom_jitter(position = position_jitter(width = 0.2), size = 2, alpha = 0.5) +
  facet_wrap(~Moraine) +
  theme_modern() +
  xlab(NULL) +
  ylab("Distance to centroid - 16S") +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = "right"
  )

# Print the plot
p_beta

# Save as SVG
ggsave("La_Gla_16S_genus_renamed_16S_BetaDisp.svg", p_beta,device = "svg", width = 8, height = 3)

#first moraine w betaDist
Bac_meta_first_moraine <- Bac_meta[Bac_meta$Moraine_num == 1, ]
# Fit the ART model (fixed effect: Treatment, random effect: Plot)


# Fit the ART model (fixed effect: Treatment, random effect: Plot)
art_model <- art(betaDist ~  Sample.Type , data = Bac_meta_first_moraine)


anova_art <- anova(art_model)
print(anova_art)


# Extract the aligned model (for Treatment)
art_emm <- emmeans(artlm(art_model, "Sample.Type"), pairwise ~ Sample.Type, adjust = "fdr")

# See pairwise comparisons
summary(art_emm$contrasts)

pw_df <- summary(art_emm$contrasts)

letters_df <- cldList(p.value ~ contrast, data = pw_df, threshold = 0.05)

#second moraine w betaDist
Bac_meta_second_moraine <- Bac_meta[Bac_meta$Moraine_num == 2, ]
# Fit the ART model (fixed effect: Treatment, random effect: Plot)


# Fit the ART model (fixed effect: Treatment, random effect: Plot)
art_model <- art(betaDist ~  Sample.Type , data = Bac_meta_second_moraine)


anova_art <- anova(art_model)
print(anova_art)


# Extract the aligned model (for Treatment)
art_emm <- emmeans(artlm(art_model, "Sample.Type"), pairwise ~ Sample.Type, adjust = "fdr")

# See pairwise comparisons
summary(art_emm$contrasts)

pw_df <- summary(art_emm$contrasts)

letters_df <- cldList(p.value ~ contrast, data = pw_df, threshold = 0.05)


#third moraine w betaDist
Bac_meta_third_moraine <- Bac_meta[Bac_meta$Moraine_num == 3, ]
# Fit the ART model (fixed effect: Treatment, random effect: Plot)


# Fit the ART model (fixed effect: Treatment, random effect: Plot)
art_model <- art(betaDist ~  Sample.Type , data = Bac_meta_third_moraine)


anova_art <- anova(art_model)
print(anova_art)


# Extract the aligned model (for Treatment)
art_emm <- emmeans(artlm(art_model, "Sample.Type"), pairwise ~ Sample.Type, adjust = "fdr")

# See pairwise comparisons
summary(art_emm$contrasts)

pw_df <- summary(art_emm$contrasts)

letters_df <- cldList(p.value ~ contrast, data = pw_df, threshold = 0.05)




```






```{r, all samples ITS, echo=FALSE}

La_Gla_ITS_genus_renamed_pseudo <- La_Gla_ITS_genus_renamed

# 1. Extract OTU table, add pseudocount
otu_with_pseudo <- otu_table(La_Gla_ITS_genus_renamed) + 1e-6

# 2. Replace the OTU table in the phyloseq object
otu_table(La_Gla_ITS_genus_renamed_pseudo) <- otu_with_pseudo

fun_meta <- sample_data(La_Gla_ITS_genus_renamed_pseudo)


La_Gla_ITS_genus_renamed_pseudo %>%
 tax_transform(rank = "Phylum", trans = "clr") %>%
 dist_calc(dist = "euclidean") %>%
 ord_calc(
  method = "PCoA"
 ) %>% 
 ord_plot(
  axes = c(1, 2),
  colour = "Sample.Type", fill = "Sample.Type",
  shape = "Moraine", alpha = 0.5,
  size = 3,
  plot_taxa = 1:10,
  )+
 scale_shape_girafe_filled()+
  ggtitle(
    "PCA of Fungal Phyla from All Samples",
  )
  
  
La_Gla_ITS_genus_renamed %>%
 tax_transform(rank = "Genus", trans = "clr") %>%
 dist_calc(dist = "euclidean") %>%
 ord_calc(
  method = "auto"
 ) %>% 
 ord_plot(
  axes = c(1, 2),
  colour = "Sample.Type", fill = "Sample.Type",
  shape = "Moraine", alpha = 0.5,
  size = 3,
  plot_taxa = 1:10,
  )+
 scale_shape_girafe_filled()+
  ggtitle(
    "PCA of Fungal Genera from All Samples",
  )  

P <- La_Gla_ITS_genus_renamed_pseudo %>%
  tax_transform(rank = "Genus", trans = "clr") %>%
  dist_calc(dist = "euclidean") %>%
  ord_calc(method = "auto") %>%
  ord_plot(
    axes = c(1, 2),
    colour = "Sample.Type",    # outline color (will be overwritten by scale_colour_manual)
    fill = "Sample.Type",      # fill color
    shape = "Moraine",
    alpha = 0.8,               # fill opacity
    size = 2
  ) +
  scale_shape_girafe_filled() +
  ggplot2::stat_ellipse(ggplot2::aes(colour = Sample.Type)) +
  scale_colour_manual(   # outline color
    values = c(
      "Lichen" = "#FF7F0E",
      "Bulk Soil" = "#59310E",
      "Plant" = "#1DA20E",
      "Moss" = "#A3A500",
      "Soil Crust" = "#155A8A"
    )
  ) +
  scale_fill_manual(     # fill color
    values = c(
      "Lichen" = "#FF7F0E",
      "Bulk Soil" = "#59310E",
      "Plant" = "#1DA20E",
      "Moss" = "#A3A500",
      "Soil Crust" = "#155A8A"
    )
  ) +
  ggtitle("PCoA of Fungal Genera from All Samples_Final_pseudo") +
  theme_minimal()
 
 P
 
 # Save as SVG
ggsave("La_Gla_16S_genus_renamed_ITSS_PCoA.svg", P,device = "svg", width = 8, height = 8)






# Extract ordination object
ord <- La_Gla_ITS_genus_renamed_pseudo %>%
  tax_transform(rank = "Genus", trans = "clr") %>%
  dist_calc(dist = "euclidean") %>%
  ord_calc(method = "auto")

# Create the plot manually
P <- ord_plot(ord, axes = c(1, 2)) +
  geom_point(
    aes(
      x = Axis.1, y = Axis.2,
      fill = Sample.Type,
      shape = Moraine
    ),
    data = sample_data(ord),
    color = "black",       # outline color
    stroke = 0.8,          # outline thickness
    alpha = 0.8,           # fill opacity
    size = 3
  ) +
  stat_ellipse(aes(x = Axis.1, y = Axis.2, colour = Sample.Type),
               data = sample_data(ord)) +
  scale_shape_girafe_filled() +
  scale_fill_manual(values = c(
    "Lichen" = "#FF7F0E",
    "Bulk Soil" = "#59310E",
    "Plant" = "#1DA20E",
    "Moss" = "#A3A500",
    "Soil Crust" = "#155A8A"
  )) +
  scale_colour_manual(values = c(
    "Lichen" = "#FF7F0E",
    "Bulk Soil" = "#59310E",
    "Plant" = "#1DA20E",
    "Moss" = "#A3A500",
    "Soil Crust" = "#155A8A"
  )) +
  ggtitle("PCA of Fungal Genera from All Samples_Final_pseudo_testingOutlines") +
  theme_minimal()

# Save as SVG
#ggsave("La_Gla_ITS_genus_renamed_ITSS_PCoA_testingOutlines.svg", P,device = "svg", width = 8, height = 8)




#plot shannon for ITS
p <- plot_richness(La_Gla_ITS_genus_renamed, x = "Sample.Type", color = "Sample.Type", measures = c("Shannon"))

# Remove default points layer
p$layers <- p$layers[-1]

# Add boxplot, faded points, and facet by Moraine
p <- p + 
  geom_boxplot(size = 1) + 
  geom_jitter(position = position_jitter(width = 0.2), size = 2, alpha = 0.5) + 
  facet_wrap(~Moraine) + 
  theme_modern()+
  xlab(NULL)+
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )+  ylab("Shannon Diversity Index - ITS")

p
# Save as SVG
ggsave("La_Gla_ITS_genus_renamed_ITS_Shannon.svg", p,device = "svg", width = 8, height = 3)


#ADD TO METADATA TABLE

# 1. Calculate richness (e.g., Shannon)
shannon_df <- estimate_richness(La_Gla_ITS_genus_renamed, measures = "Shannon")

# 2. Check that rownames match sample names
head(rownames(shannon_df))
head(sample_names(La_Gla_ITS_genus_renamed))

# 3. Convert sample_data to data frame if needed
sample_df <- as.matrix(sample_data(La_Gla_ITS_genus_renamed))
sample_df <- as.data.frame(sample_df)

# 4. Combine richness values with sample_data
sample_df$Fungi_Shannon <- shannon_df$Shannon

# 5. Update sample_data in the phyloseq object
sample_data(La_Gla_ITS_genus_renamed) <- sample_data(sample_df)





#plot chao richness for ITS
p <- plot_richness(La_Gla_ITS_genus_renamed, x = "Sample.Type", color = "Sample.Type", measures = c("Chao1"))

# Remove default points layer
p$layers <- p$layers[-1]

# Add boxplot, faded points, and facet by Moraine
p <- p + 
  geom_boxplot(size = 1) + 
  geom_jitter(position = position_jitter(width = 0.2), size = 2, alpha = 0.5) + 
  facet_wrap(~Moraine) + 
  theme_modern()+
  xlab(NULL)+
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )+ ylab("Chao Richness - ITS")

p
# Save as SVG
ggsave("La_Gla_ITS_genus_renamed_ITS_Chao.svg", p,device = "svg", width = 8, height = 3)

#ADD TO METADATA TABLE

# 1. Calculate richness (e.g., Shannon)
Chao_df <- estimate_richness(La_Gla_ITS_genus_renamed, measures = "Chao1")

# 2. Check that rownames match sample names
head(rownames(Chao_df))
head(sample_names(La_Gla_ITS_genus_renamed))

# 3. Convert sample_data to data frame if needed
sample_df <- as.matrix(sample_data(La_Gla_ITS_genus_renamed))
sample_df <- as.data.frame(sample_df)

# 4. Combine richness values with sample_data
sample_df$Fungi_Chao <- Chao_df$Chao1

# 5. Update sample_data in the phyloseq object
sample_data(La_Gla_ITS_genus_renamed) <- sample_data(sample_df)




p <-plot_richness(La_Gla_ITS_genus_renamed, x="Moraine", color = "Sample.Type", measures=c( "Shannon"))
p$layers <- p$layers[-1]
p$layers
p +  geom_boxplot()

p <-plot_richness(La_Gla_ITS_genus_renamed, x="Moraine", color = "Sample.Type", measures=c("Chao1"))
p$layers <- p$layers[-1]
p$layers <- p$layers[-1]
p$layers
p +  geom_boxplot()
# 
# La_Gla_ITS_genus_renamed %>%
#  tax_transform(rank = "Genus", trans = "binary") %>%
#  dist_calc(dist = "bray") %>%
#  ord_calc(
#   method = "auto"
#  ) %>% 
#  ord_plot(
#   axes = c(1, 2),
#   colour = "Sample.Type", fill = "Sample.Type",
#   shape = "Moraine", alpha = 0.5,
#   size = 2
#  ) + 
#  scale_shape_girafe_filled() +
#  ggplot2::stat_ellipse(
#   ggplot2::aes(colour = Sample.Type)
#  )+
#   ggtitle(
#     "PCA of Fungal Genera from All Samples",
#   )  

    
#  La_Gla_ITS_genus_renamed %>% perm_get()


#mixed effect models

fungi_meta <- as.matrix(sample_data(La_Gla_ITS_genus_renamed))
fungi_meta <- as.data.frame(fungi_meta)

fungi_meta$Fungi_Shannon <- as.numeric(fungi_meta$Fungi_Shannon)
fungi_meta$Moraine_num <- as.numeric(fungi_meta$Moraine_num)
fungi_meta$Sample.Type <- as.factor(fungi_meta$Sample.Type)
fungi_meta$Location <- as.factor(fungi_meta$Location)
fungi_meta$Plot <- as.factor(fungi_meta$Plot)
fungi_meta$Fungi_Chao <- as.numeric(fungi_meta$Fungi_Chao)
fungi_meta$Moraine <- as.factor(fungi_meta$Moraine)


mixed_effect_model <- lmer(Fungi_Shannon ~ Moraine * Sample.Type * Location  + (1|Plot), data = fungi_meta)

# Perform ANOVA using the car package
anova_result <- Anova(mixed_effect_model, test="F")

# Display the ANOVA table
print(anova_result)

#subset first moraine
fungi_meta_first_moraine <- fungi_meta[fungi_meta$Moraine_num == 1, ]

# Fit the ART model (fixed effect: Treatment, random effect: Plot)
art_model <- art(Fungi_Shannon ~  Sample.Type , data = fungi_meta_first_moraine)


anova_art <- anova(art_model)
print(anova_art)


# Extract the aligned model (for Treatment)
art_emm <- emmeans(artlm(art_model, "Sample.Type"), pairwise ~ Sample.Type, adjust = "fdr")

# See pairwise comparisons
summary(art_emm$contrasts)

pw_df <- summary(art_emm$contrasts)

letters_df <- cldList(p.value ~ contrast, data = pw_df, threshold = 0.05)

#now for second moraine
fungi_meta_second_moraine <- fungi_meta[fungi_meta$Moraine_num == 2, ]

# Fit the ART model (fixed effect: Treatment, random effect: Plot)
art_model_second <- art(Fungi_Shannon ~  Sample.Type , data = fungi_meta_second_moraine)

anova_art_second <- anova(art_model_second)
print(anova_art_second)

# Extract the aligned model (for Treatment)
art_emm_second <- emmeans(artlm(art_model_second, "Sample.Type"), pairwise ~ Sample.Type, adjust = "fdr")

# See pairwise comparisons
summary(art_emm_second$contrasts)
pw_df_second <- summary(art_emm_second$contrasts)
letters_df_second <- cldList(p.value ~ contrast, data = pw_df_second, threshold = 0.05)

#third moraine now
fungi_meta_third_moraine <- fungi_meta[fungi_meta$Moraine_num == 3, ]

# Fit the ART model (fixed effect: Treatment, random effect: Plot)
art_model_third <- art(Fungi_Shannon ~  Sample.Type , data = fungi_meta_third_moraine)
anova_art_third <- anova(art_model_third)

print(anova_art_third)

# Extract the aligned model (for Treatment)
art_emm_third <- emmeans(artlm(art_model_third, "Sample.Type"), pairwise ~ Sample.Type, adjust = "fdr")
# See pairwise comparisons
summary(art_emm_third$contrasts)
pw_df_third <- summary(art_emm_third$contrasts)
letters_df_third <- cldList(p.value ~ contrast, data = pw_df_third, threshold = 0.05)


#now for richness
mixed_effect_model <- lmer(Fungi_Chao ~ Moraine * Sample.Type * Location  + (1|Plot), data = fungi_meta)

# Perform ANOVA using the car package
anova_result <- Anova(mixed_effect_model, test="F")

# Display the ANOVA table
print(anova_result)


#first moraine w chao
fungi_meta_first_moraine <- fungi_meta[fungi_meta$Moraine_num == 1, ]
# Fit the ART model (fixed effect: Treatment, random effect: Plot)
art_model_chao <- art(Fungi_Chao ~  Sample.Type , data = fungi_meta_first_moraine)
anova_art_chao <- anova(art_model_chao)
print(anova_art_chao)

# Extract the aligned model (for Treatment)
art_emm_chao <- emmeans(artlm(art_model_chao, "Sample.Type"), pairwise ~ Sample.Type, adjust = "fdr")
# See pairwise comparisons
summary(art_emm_chao$contrasts)
pw_df_chao <- summary(art_emm_chao$contrasts)
letters_df_chao <- cldList(p.value ~ contrast, data = pw_df_chao, threshold = 0.05)


#second moraine w chao
fungi_meta_second_moraine <- fungi_meta[fungi_meta$Moraine_num == 2, ]
# Fit the ART model (fixed effect: Treatment, random effect: Plot)
art_model_chao_second <- art(Fungi_Chao ~  Sample.Type , data = fungi_meta_second_moraine)
anova_art_chao_second <- anova(art_model_chao_second)
print(anova_art_chao_second)

# Extract the aligned model (for Treatment)
art_emm_chao_second <- emmeans(artlm(art_model_chao_second, "Sample.Type"), pairwise ~ Sample.Type, adjust = "fdr")
# See pairwise comparisons
summary(art_emm_chao_second$contrasts)
pw_df_chao_second <- summary(art_emm_chao_second$contrasts)
letters_df_chao_second <- cldList(p.value ~ contrast, data = pw_df_chao_second, threshold = 0.05)


#third moraine
fungi_meta_third_moraine <- fungi_meta[fungi_meta$Moraine_num == 3, ]
# Fit the ART model (fixed effect: Treatment, random effect: Plot)
art_model_chao_third <- art(Fungi_Chao ~  Sample.Type , data = fungi_meta_third_moraine)
anova_art_chao_third <- anova(art_model_chao_third)
print(anova_art_chao_third)

# Extract the aligned model (for Treatment)
art_emm_chao_third <- emmeans(artlm(art_model_chao_third, "Sample.Type"), pairwise ~ Sample.Type, adjust = "fdr")
# See pairwise comparisons
summary(art_emm_chao_third$contrasts)
pw_df_chao_third <- summary(art_emm_chao_third$contrasts)
letters_df_chao_third <- cldList(p.value ~ contrast, data = pw_df_chao_third, threshold = 0.05)












print("      permanova with rarefaction at genus level, with aitchinson distance")
La_Gla_ITS_genus_renamed_perm <- La_Gla_ITS_genus_renamed_pseudo %>%
  tax_transform("clr", rank = "Genus") %>%
  dist_calc("euclidean") %>%
  dist_permanova(
    variables = c("GlacierID","Sample.Type", "Moraine", "Library_Size", "Location"),
    n_perms = 999, # you should use more permutations in your real analyses!
    strata = fun_meta$Plot,
    n_processes = 1
  )

La_Gla_ITS_genus_renamed_perm

print("      permanova with rarefaction at genus level, with aitchinson distance, including plot")
La_Gla_ITS_genus_renamed_perm <- La_Gla_ITS_genus_renamed %>%
  tax_transform("clr", rank = "Genus") %>%
  dist_calc("euclidean") %>%
  dist_permanova(
    variables = c("Sample.Type", "Moraine", "Library_Size", "Plot"),
    n_perms = 999, # you should use more permutations in your real analyses!
    n_processes = 1
  )

La_Gla_ITS_genus_renamed_perm

#ord_explore(La_Gla_ITS_genus_renamed_pseudo)

#Extract OTU table
Fun_otu_pseudo <- as.matrix(otu_table(La_Gla_ITS_genus_renamed_pseudo)) 
Fun_otu_pseudo <- as.data.frame(Fun_otu_pseudo)

#CLR transformation
Fun_otu_clr <- Fun_otu_pseudo %>%
  microbiome::transform("clr")
Fun_otu_clr <- as.data.frame(Fun_otu_clr)

class(Fun_otu_clr)


fun_meta <- as.matrix(sample_data(La_Gla_ITS_genus_renamed_pseudo))
fun_meta <- as.data.frame(fun_meta)

class(fun_meta)

#convert to numeric and factor
fun_meta$Moraine <- as.factor(fun_meta$Moraine)
fun_meta$Sample.Type <- as.factor(fun_meta$Sample.Type)
fun_meta$Location <- as.factor(fun_meta$Location)
fun_meta$Plot <- as.factor(fun_meta$Plot)

result <- adonis2(Fun_otu_clr ~  Moraine + Sample.Type + Location, data = fun_meta, method="euclidean", strata = fun_meta$Plot ,perm=999)
print(result)



#run betadisperse
# Get Sample.Type grouping variable
group <- sample_data(La_Gla_ITS_genus_renamed_pseudo)$Sample.Type

# Calculate Bray-Curtis dissimilarity
euc_dist <- vegdist(Fun_otu_clr, method = "euclidean")


# Run beta dispersion
disper <- betadisper(euc_dist, group)

# Test for significant differences in dispersion
anova(disper)

plot(disper)
boxplot(disper, xlab = "Sample Type", ylab = "Distance to Centroid")



permutest(disper, permutations = 999)

#beta dispersion for ITS, separated by moraine

#extract betadispersion values
# Get Sample.Type grouping variable
samp_group <- sample_data(La_Gla_ITS_genus_renamed_pseudo)$Sample.Type

moraine_group <- sample_data(La_Gla_ITS_genus_renamed_pseudo)$Moraine

#create interaction term
group <- interaction(samp_group, moraine_group)

# Calculate Bray-Curtis dissimilarity
euc_dist <- vegdist(Fun_otu_clr, method = "euclidean")


# Run beta dispersion
disper <- betadisper(euc_dist, group)

#extract values from disper
fungi_meta$betaDist <- disper$distances

#now for richness
mixed_effect_model <- lmer(betaDist ~ Moraine * Sample.Type * Location  + (1|Plot), data = fungi_meta)

# Perform ANOVA using the car package
anova_result <- Anova(mixed_effect_model, test="F")

# Display the ANOVA table
print(anova_result)


# Create the boxplot
p_beta <- ggplot(fungi_meta, aes(x = Sample.Type, y = betaDist, color = Sample.Type)) +
  geom_boxplot(size = 1) +
  geom_jitter(position = position_jitter(width = 0.2), size = 2, alpha = 0.5) +
  facet_wrap(~Moraine) +
  theme_modern() +
  xlab(NULL) +
  #ylab("Distance to centroid - 16S") +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = "right"
  )

# Print the plot
p_beta

# Save as SVG
ggsave("La_Gla_ITS_genus_renamed_ITS_BetaDisp.svg", p_beta,device = "svg", width = 8, height = 3)

#first moraine w betaDist
fungi_meta_first_moraine <- fungi_meta[fungi_meta$Moraine_num == 1, ]
# Fit the ART model (fixed effect: Treatment, random effect: Plot)


# Fit the ART model (fixed effect: Treatment, random effect: Plot)
art_model <- art(betaDist ~  Sample.Type , data = fungi_meta_first_moraine)


anova_art <- anova(art_model)
print(anova_art)


# Extract the aligned model (for Treatment)
art_emm <- emmeans(artlm(art_model, "Sample.Type"), pairwise ~ Sample.Type, adjust = "fdr")

# See pairwise comparisons
summary(art_emm$contrasts)

pw_df <- summary(art_emm$contrasts)

letters_df <- cldList(p.value ~ contrast, data = pw_df, threshold = 0.05)

#second moraine w betaDist
fungi_meta_second_moraine <- fungi_meta[fungi_meta$Moraine_num == 2, ]
# Fit the ART model (fixed effect: Treatment, random effect: Plot)


# Fit the ART model (fixed effect: Treatment, random effect: Plot)
art_model <- art(betaDist ~  Sample.Type , data = fungi_meta_second_moraine)


anova_art <- anova(art_model)
print(anova_art)


# Extract the aligned model (for Treatment)
art_emm <- emmeans(artlm(art_model, "Sample.Type"), pairwise ~ Sample.Type, adjust = "fdr")

# See pairwise comparisons
summary(art_emm$contrasts)

pw_df <- summary(art_emm$contrasts)

letters_df <- cldList(p.value ~ contrast, data = pw_df, threshold = 0.05)


#third moraine w betaDist
fungi_meta_third_moraine <- fungi_meta[fungi_meta$Moraine_num == 3, ]
# Fit the ART model (fixed effect: Treatment, random effect: Plot)


# Fit the ART model (fixed effect: Treatment, random effect: Plot)
art_model <- art(betaDist ~  Sample.Type , data = fungi_meta_third_moraine)


anova_art <- anova(art_model)
print(anova_art)


# Extract the aligned model (for Treatment)
art_emm <- emmeans(artlm(art_model, "Sample.Type"), pairwise ~ Sample.Type, adjust = "fdr")

# See pairwise comparisons
summary(art_emm$contrasts)

pw_df <- summary(art_emm$contrasts)

letters_df <- cldList(p.value ~ contrast, data = pw_df, threshold = 0.05)






```


#Explore Venn diagrams at the genera level
```{r, echo=FALSE}

#La_Gla_ITS_genus_renamed

# Agglomerate by Genus
La_Gla_ITS_genus <- tax_glom(La_Gla_ITS_genus_renamed, taxrank = "Genus")

# Convert ASV/OTU names to genus names
taxa_names(La_Gla_ITS_genus) <- as.character(tax_table(La_Gla_ITS_genus)[, "Genus"])

ITS_BS <- subset_samples(La_Gla_ITS_genus, Sample.Type == "Bulk Soil")
ITS_moss <- subset_samples(La_Gla_ITS_genus, Sample.Type == "Moss")
ITS_SC <- subset_samples(La_Gla_ITS_genus, Sample.Type == "Soil Crust")
ITS_Pla <- subset_samples(La_Gla_ITS_genus, Sample.Type == "Plant")
ITS_Lic <- subset_samples(La_Gla_ITS_genus, Sample.Type == "Lichen")

Bac_BS <- subset_samples(La_Gla_16S_genus_renamed, Sample.Type == "Bulk Soil")
Bac_moss <- subset_samples(La_Gla_16S_genus_renamed, Sample.Type == "Moss")
Bac_SC <- subset_samples(La_Gla_16S_genus_renamed, Sample.Type == "Soil Crust")
Bac_Pla <- subset_samples(La_Gla_16S_genus_renamed, Sample.Type == "Plant")
Bac_Lic <- subset_samples(La_Gla_16S_genus_renamed, Sample.Type == "Lichen")

library(phyloseq)
library(dplyr)
library(tibble)

get_top5_genera <- function(physeq_object) {
  # Transform to relative abundance
  physeq_rel <- transform_sample_counts(physeq_object, function(x) x / sum(x))
  
  # Agglomerate to Genus level
  physeq_genus <- tax_glom(physeq_rel, taxrank = "Genus", NArm = TRUE)

  # Extract the abundance matrix (ensure taxa are rows)
  abund <- t(as(otu_table(physeq_rel), "matrix"))

  
  # Extract the tax table from the genus-level phyloseq object
  taxa <- as(tax_table(physeq_rel), "matrix")
  
  # Calculate mean and max abundance
  genus_stats <- data.frame(
    Genus = taxa[, "Genus"],
    MeanAbundance = rowMeans(abund),
    MaxAbundance = apply(abund, 1, max),
    stringsAsFactors = FALSE
  )
  
  # Combine with full taxonomy
  taxonomy_df <- as.data.frame(taxa)
  genus_stats <- cbind(genus_stats, taxonomy_df)
  
  
  
  # Return top 5 by mean abundance
  
  
  return(genus_stats)
}

# Run for each sample type
top_BS <- get_top5_genera(ITS_BS)
top_Moss <- get_top5_genera(ITS_moss)
top_SC <- get_top5_genera(ITS_SC)
top_Pla <- get_top5_genera(ITS_Pla)
top_Lic <- get_top5_genera(ITS_Lic)

# View result
Bac_top_BS <- get_top5_genera(Bac_BS)
Bac_top_Moss <- get_top5_genera(Bac_moss)
Bac_top_SC <- get_top5_genera(Bac_SC)
Bac_top_Pla <- get_top5_genera(Bac_Pla)
Bac_top_Lic <- get_top5_genera(Bac_Lic)



otus_BS <- taxa_names(prune_taxa(taxa_sums(ITS_BS) > 0, ITS_BS))
otus_Moss <- taxa_names(prune_taxa(taxa_sums(ITS_moss) > 0, ITS_moss))
otus_SC <- taxa_names(prune_taxa(taxa_sums(ITS_SC) > 0, ITS_SC))
otus_Pla <- taxa_names(prune_taxa(taxa_sums(ITS_Pla) > 0, ITS_Pla))
otus_Lic <- taxa_names(prune_taxa(taxa_sums(ITS_Lic) > 0, ITS_Lic))

venn <- venn.diagram(
  x = list(BS = otus_BS, Moss = otus_Moss, Soil_Crust = otus_SC, Plants = otus_Pla, Lichen = otus_Lic),
  category.names = c("Bulk Soil", "Moss", "Soil Crust", "Plant", "Lichen"),
  filename = "VennDiagram_all_ITS_genus.png",
  output = TRUE,
    main = "Venn Diagram of Fungal Genera across all Moraines",  # Add title here
  main.cex = 1  # Adjust title size (optional)
)

# Convert to presence-absence
ITS_unique <- transform_sample_counts(La_Gla_ITS_genus, function(x) ifelse(x > 0, 1, 0))

# Convert to a matrix for easier processing
otu_matrix <- as(otu_table(ITS_unique), "matrix")

# Convert to data frame for manipulation
otu_df <- as.data.frame(otu_matrix)

# Sum presence across samples per taxon
otu_sums <- rowSums(otu_df)

# Identify unique taxa for each group
unique_taxa <- list()

for (group in colnames(otu_df)) {
  unique_taxa[[group]] <- rownames(otu_df)[otu_df[, group] == 1 & otu_sums == 1]
}

# Print unique taxa per group
unique_taxa


```




```{r, venn diagrams for separate moraines - fungi}
#Fungi First Moraine Venn Diagrams
# Example: Subset phyloseq object based on sample groups
ITS_morA <- subset_samples(La_Gla_ITS_genus, Moraine == "A")
Meta_ITS_A <- sample_data(ITS_morA)

ITS_morA <- prune_taxa(taxa_sums(ITS_morA) > 0, ITS_morA)

ntaxa(ITS_morA)

ITS_morA_BS <- subset_samples(ITS_morA, Sample.Type == "Bulk Soil")
ITS_morA_moss <- subset_samples(ITS_morA, Sample.Type == "Moss")
ITS_morA_SC <- subset_samples(ITS_morA, Sample.Type == "Soil Crust")
ITS_morA_Pla <- subset_samples(ITS_morA, Sample.Type == "Plant")
ITS_morA_Lic <- subset_samples(ITS_morA, Sample.Type == "Lichen")

otus_BS <- taxa_names(prune_taxa(taxa_sums(ITS_morA_BS) > 0, ITS_morA_BS))
otus_Moss <- taxa_names(prune_taxa(taxa_sums(ITS_morA_moss) > 0, ITS_morA_moss))
otus_SC <- taxa_names(prune_taxa(taxa_sums(ITS_morA_SC) > 0, ITS_morA_SC))
otus_Pla <- taxa_names(prune_taxa(taxa_sums(ITS_morA_Pla) > 0, ITS_morA_Pla))
otus_Lic <- taxa_names(prune_taxa(taxa_sums(ITS_morA_Lic) > 0, ITS_morA_Lic))

venn <- venn.diagram(
  x = list(BS = otus_BS, Moss = otus_Moss, Soil_Crust = otus_SC, Plants = otus_Pla, Lichen = otus_Lic),
  category.names = c("Bulk Soil", "Moss", "Soil Crust", "Plant", "Lichen"),
  filename = "VennDiagram_ITS_A.png",
  output = TRUE,
    main = "Venn Diagram of Fungal Genera on the First Moraines",  # Add title here
  main.cex = 1  # Adjust title size (optional)
)


#Fungi First Second Moraine Venn Diagrams
# Example: Subset phyloseq object based on sample groups
ITS_morB <- subset_samples(La_Gla_ITS_genus, Moraine == "B")
Meta_ITS_B <- sample_data(ITS_morB)

ITS_morB <- prune_taxa(taxa_sums(ITS_morB) > 0, ITS_morB)

ntaxa(ITS_morB)

ITS_morB_BS <- subset_samples(ITS_morB, Sample.Type == "Bulk Soil")
ITS_morB_moss <- subset_samples(ITS_morB, Sample.Type == "Moss")
ITS_morB_SC <- subset_samples(ITS_morB, Sample.Type == "Soil Crust")
ITS_morB_Pla <- subset_samples(ITS_morB, Sample.Type == "Plant")
ITS_morB_Lic <- subset_samples(ITS_morB, Sample.Type == "Lichen")

otus_BS <- taxa_names(prune_taxa(taxa_sums(ITS_morB_BS) > 0, ITS_morB_BS))
otus_Moss <- taxa_names(prune_taxa(taxa_sums(ITS_morB_moss) > 0, ITS_morB_moss))
otus_SC <- taxa_names(prune_taxa(taxa_sums(ITS_morB_SC) > 0, ITS_morB_SC))
otus_Pla <- taxa_names(prune_taxa(taxa_sums(ITS_morB_Pla) > 0, ITS_morB_Pla))
otus_Lic <- taxa_names(prune_taxa(taxa_sums(ITS_morB_Lic) > 0, ITS_morB_Lic))

venn <- venn.diagram(
  x = list(BS = otus_BS, Moss = otus_Moss, Soil_Crust = otus_SC, Plants = otus_Pla, Lichen = otus_Lic),
  category.names = c("Bulk Soil", "Moss", "Soil Crust", "Plant", "Lichen"),
  filename = "VennDiagram_ITS_B.png",
  output = TRUE,
    main = "Venn Diagram of Fungal Genera on the Second Moraines",  # Add title here
  main.cex = 1  # Adjust title size (optional)
)



#Fungi  Third Moraine Venn Diagrams
# Example: Subset phyloseq object based on sample groups
ITS_morC <- subset_samples(La_Gla_ITS_genus, Moraine == "C")
Meta_ITS_C <- sample_data(ITS_morC)

ITS_morC <- prune_taxa(taxa_sums(ITS_morC) > 0, ITS_morC)

ntaxa(ITS_morC)

ITS_morC_BS <- subset_samples(ITS_morC, Sample.Type == "Bulk Soil")
ITS_morC_moss <- subset_samples(ITS_morC, Sample.Type == "Moss")
ITS_morC_SC <- subset_samples(ITS_morC, Sample.Type == "Soil Crust")
ITS_morC_Pla <- subset_samples(ITS_morC, Sample.Type == "Plant")
ITS_morC_Lic <- subset_samples(ITS_morC, Sample.Type == "Lichen")

otus_BS <- taxa_names(prune_taxa(taxa_sums(ITS_morC_BS) > 0, ITS_morC_BS))
otus_Moss <- taxa_names(prune_taxa(taxa_sums(ITS_morC_moss) > 0, ITS_morC_moss))
otus_SC <- taxa_names(prune_taxa(taxa_sums(ITS_morC_SC) > 0, ITS_morC_SC))
otus_Pla <- taxa_names(prune_taxa(taxa_sums(ITS_morC_Pla) > 0, ITS_morC_Pla))
otus_Lic <- taxa_names(prune_taxa(taxa_sums(ITS_morC_Lic) > 0, ITS_morC_Lic))


venn <- venn.diagram(
  x = list(BS = otus_BS, Moss = otus_Moss, Soil_Crust = otus_SC, Plants = otus_Pla, Lichen = otus_Lic),
  category.names = c("Bulk Soil", "Moss", "Soil Crust", "Plant", "Lichen"),
  filename = "VennDiagram_ITS_C.png",
  output = TRUE,
    main = "Venn Diagram of Fungal Genera on the Third Moraines",  # Add title here
  main.cex = 1  # Adjust title size (optional)
)




#Bacterial Venn Diagrams

# Agglomerate by Genus
La_Gla_16S_genus <- tax_glom(La_Gla_16S_genus_renamed, taxrank = "Genus")

# Convert ASV/OTU names to genus names
taxa_names(La_Gla_16S_genus) <- as.character(tax_table(La_Gla_16S_genus)[, "Genus"])

Bac_BS <- subset_samples(La_Gla_16S_genus, Sample.Type == "Bulk Soil")
Bac_moss <- subset_samples(La_Gla_16S_genus, Sample.Type == "Moss")
Bac_SC <- subset_samples(La_Gla_16S_genus, Sample.Type == "Soil Crust")
Bac_Pla <- subset_samples(La_Gla_16S_genus, Sample.Type == "Plant")
Bac_Lic <- subset_samples(La_Gla_16S_genus, Sample.Type == "Lichen")

otus_BS <- taxa_names(prune_taxa(taxa_sums(Bac_BS) > 0, Bac_BS))
otus_Moss <- taxa_names(prune_taxa(taxa_sums(Bac_moss) > 0, Bac_moss))
otus_SC <- taxa_names(prune_taxa(taxa_sums(Bac_SC) > 0, Bac_SC))
otus_Pla <- taxa_names(prune_taxa(taxa_sums(Bac_Pla) > 0, Bac_Pla))
otus_Lic <- taxa_names(prune_taxa(taxa_sums(Bac_Lic) > 0, Bac_Lic))

venn <- venn.diagram(
  x = list(BS = otus_BS, Moss = otus_Moss, Soil_Crust = otus_SC, Plants = otus_Pla, Lichen = otus_Lic),
  category.names = c("Bulk Soil", "Moss", "Soil Crust", "Plant", "Lichen"),
  filename = "VennDiagram_all_Bac_genus.png",
  output = TRUE,
    main = "Venn Diagram of Bacterial Genera across all Moraines",  # Add title here
  main.cex = 1  # Adjust title size (optional)
)

#Bacteria First Moraine Venn Diagrams
# Example: Subset phyloseq object based on sample groups
Bac_morA <- subset_samples(La_Gla_16S_genus, Moraine == "A")
Meta_Bac_A <- sample_data(Bac_morA)

Bac_morA <- prune_taxa(taxa_sums(Bac_morA) > 0, Bac_morA)

ntaxa(Bac_morA)

Bac_morA_BS <- subset_samples(Bac_morA, Sample.Type == "Bulk Soil")
Bac_morA_moss <- subset_samples(Bac_morA, Sample.Type == "Moss")
Bac_morA_SC <- subset_samples(Bac_morA, Sample.Type == "Soil Crust")
Bac_morA_Pla <- subset_samples(Bac_morA, Sample.Type == "Plant")
Bac_morA_Lic <- subset_samples(Bac_morA, Sample.Type == "Lichen")

otus_BS <- taxa_names(prune_taxa(taxa_sums(Bac_morA_BS) > 0, Bac_morA_BS))
otus_Moss <- taxa_names(prune_taxa(taxa_sums(Bac_morA_moss) > 0, Bac_morA_moss))
otus_SC <- taxa_names(prune_taxa(taxa_sums(Bac_morA_SC) > 0, Bac_morA_SC))
otus_Pla <- taxa_names(prune_taxa(taxa_sums(Bac_morA_Pla) > 0, Bac_morA_Pla))
otus_Lic <- taxa_names(prune_taxa(taxa_sums(Bac_morA_Lic) > 0, Bac_morA_Lic))

venn <- venn.diagram(
  x = list(BS = otus_BS, Moss = otus_Moss, Soil_Crust = otus_SC, Plants = otus_Pla, Lichen = otus_Lic),
  category.names = c("Bulk Soil", "Moss", "Soil Crust", "Plant", "Lichen"),
  filename = "VennDiagram_16S_A.png",
  output = TRUE,
    main = "Venn Diagram of Bacterial Genera on the First Moraines",  # Add title here
  main.cex = 1  # Adjust title size (optional)
)

#Bacteria Second Moraine Venn Diagrams
# Example: Subset phyloseq object based on sample groups
Bac_morB <- subset_samples(La_Gla_16S_genus, Moraine == "B")
Meta_Bac_B <- sample_data(Bac_morB)

Bac_morB <- prune_taxa(taxa_sums(Bac_morB) > 0, Bac_morB)

ntaxa(Bac_morB)

Bac_morB_BS <- subset_samples(Bac_morB, Sample.Type == "Bulk Soil")
Bac_morB_moss <- subset_samples(Bac_morB, Sample.Type == "Moss")
Bac_morB_SC <- subset_samples(Bac_morB, Sample.Type == "Soil Crust")
Bac_morB_Pla <- subset_samples(Bac_morB, Sample.Type == "Plant")
Bac_morB_Lic <- subset_samples(Bac_morB, Sample.Type == "Lichen")

otus_BS <- taxa_names(prune_taxa(taxa_sums(Bac_morB_BS) > 0, Bac_morB_BS))
otus_Moss <- taxa_names(prune_taxa(taxa_sums(Bac_morB_moss) > 0, Bac_morB_moss))
otus_SC <- taxa_names(prune_taxa(taxa_sums(Bac_morB_SC) > 0, Bac_morB_SC))
otus_Pla <- taxa_names(prune_taxa(taxa_sums(Bac_morB_Pla) > 0, Bac_morB_Pla))
otus_Lic <- taxa_names(prune_taxa(taxa_sums(Bac_morB_Lic) > 0, Bac_morB_Lic))

venn <- venn.diagram(
  x = list(BS = otus_BS, Moss = otus_Moss, Soil_Crust = otus_SC, Plants = otus_Pla, Lichen = otus_Lic),
  category.names = c("Bulk Soil", "Moss", "Soil Crust", "Plant", "Lichen"),
  filename = "VennDiagram_16S_B.png",
  output = TRUE,
    main = "Venn Diagram of Bacterial Genera on the Second Moraines",  # Add title here
  main.cex = 1  # Adjust title size (optional)
)


#Bacteria Third Moraine Venn Diagrams
# Example: Subset phyloseq object based on sample groups
Bac_morC <- subset_samples(La_Gla_16S_genus, Moraine == "C")
Meta_Bac_C <- sample_data(Bac_morC)

Bac_morC <- prune_taxa(taxa_sums(Bac_morC) > 0, Bac_morC)

ntaxa(Bac_morC)

Bac_morC_BS <- subset_samples(Bac_morC, Sample.Type == "Bulk Soil")
Bac_morC_moss <- subset_samples(Bac_morC, Sample.Type == "Moss")
Bac_morC_SC <- subset_samples(Bac_morC, Sample.Type == "Soil Crust")
Bac_morC_Pla <- subset_samples(Bac_morC, Sample.Type == "Plant")
Bac_morC_Lic <- subset_samples(Bac_morC, Sample.Type == "Lichen")

otus_BS <- taxa_names(prune_taxa(taxa_sums(Bac_morC_BS) > 0, Bac_morC_BS))
otus_Moss <- taxa_names(prune_taxa(taxa_sums(Bac_morC_moss) > 0, Bac_morC_moss))
otus_SC <- taxa_names(prune_taxa(taxa_sums(Bac_morC_SC) > 0, Bac_morC_SC))
otus_Pla <- taxa_names(prune_taxa(taxa_sums(Bac_morC_Pla) > 0, Bac_morC_Pla))
otus_Lic <- taxa_names(prune_taxa(taxa_sums(Bac_morC_Lic) > 0, Bac_morC_Lic))

venn <- venn.diagram(
  x = list(BS = otus_BS, Moss = otus_Moss, Soil_Crust = otus_SC, Plants = otus_Pla, Lichen = otus_Lic),
  category.names = c("Bulk Soil", "Moss", "Soil Crust", "Plant", "Lichen"),
  filename = "VennDiagram_16S_C.png",
  output = TRUE,
    main = "Venn Diagram of Bacterial Genera on the Third Moraines",  # Add title here
  main.cex = 1  # Adjust title size (optional)
)


```


```{r, echo=FALSE}
#La_Gla_ITS_family_renamed

# Agglomerate by Family
La_Gla_ITS_family <- tax_glom(La_Gla_ITS_genus_renamed, taxrank = "Family")



# Convert ASV/OTU names to family names
taxa_names(La_Gla_ITS_family) <- as.character(tax_table(La_Gla_ITS_family)[, "Family"])

ITS_BS <- subset_samples(La_Gla_ITS_family, Sample.Type == "Bulk Soil")
ITS_moss <- subset_samples(La_Gla_ITS_family, Sample.Type == "Moss")
ITS_SC <- subset_samples(La_Gla_ITS_family, Sample.Type == "Soil Crust")
ITS_Pla <- subset_samples(La_Gla_ITS_family, Sample.Type == "Plant")
ITS_Lic <- subset_samples(La_Gla_ITS_family, Sample.Type == "Lichen")

otus_BS <- taxa_names(prune_taxa(taxa_sums(ITS_BS) > 0, ITS_BS))
otus_Moss <- taxa_names(prune_taxa(taxa_sums(ITS_moss) > 0, ITS_moss))
otus_SC <- taxa_names(prune_taxa(taxa_sums(ITS_SC) > 0, ITS_SC))
otus_Pla <- taxa_names(prune_taxa(taxa_sums(ITS_Pla) > 0, ITS_Pla))
otus_Lic <- taxa_names(prune_taxa(taxa_sums(ITS_Lic) > 0, ITS_Lic))

venn <- venn.diagram(
  x = list(BS = otus_BS, Moss = otus_Moss, Soil_Crust = otus_SC, Plants = otus_Pla, Lichen = otus_Lic),
  category.names = c("Bulk Soil", "Moss", "Soil Crust", "Plant", "Lichen"),
  filename = "VennDiagram_all_fam.png",
  output = TRUE,
    main = "Venn Diagram of Fungal Families across all Moraines",  # Add title here
  main.cex = 1  # Adjust title size (optional)
)

#Fungi First Moraine Venn Diagrams
# Example: Subset phyloseq object based on sample groups
ITS_morA <- subset_samples(La_Gla_ITS_family, Moraine == "A")
ITS_morA <- prune_taxa(taxa_sums(ITS_morA) > 0, ITS_morA)

ntaxa(ITS_morA)

venn <- venn.diagram(
  x = list(BS = otus_BS, Moss = otus_Moss, Soil_Crust = otus_SC, Plants = otus_Pla, Lichen = otus_Lic),
  category.names = c("Bulk Soil", "Moss", "Soil Crust", "Plant", "Lichen"),
  filename = "VennDiagram_ITS_B_fam.png",
  output = TRUE,
    main = "Venn Diagram of Fungal Families on the Second Moraines",  # Add title here
  main.cex = 1  # Adjust title size (optional)
)

Meta_ITS_A <- sample_data(ITS_morA)

ITS_morA_BS <- subset_samples(ITS_morA, Sample.Type == "Bulk Soil")
ITS_morA_moss <- subset_samples(ITS_morA, Sample.Type == "Moss")
ITS_morA_SC <- subset_samples(ITS_morA, Sample.Type == "Soil Crust")
ITS_morA_Pla <- subset_samples(ITS_morA, Sample.Type == "Plant")
ITS_morA_Lic <- subset_samples(ITS_morA, Sample.Type == "Lichen")

otus_BS <- taxa_names(prune_taxa(taxa_sums(ITS_morA_BS) > 0, ITS_morA_BS))
otus_Moss <- taxa_names(prune_taxa(taxa_sums(ITS_morA_moss) > 0, ITS_morA_moss))
otus_SC <- taxa_names(prune_taxa(taxa_sums(ITS_morA_SC) > 0, ITS_morA_SC))
otus_Pla <- taxa_names(prune_taxa(taxa_sums(ITS_morA_Pla) > 0, ITS_morA_Pla))
otus_Lic <- taxa_names(prune_taxa(taxa_sums(ITS_morA_Lic) > 0, ITS_morA_Lic))

venn <- venn.diagram(
  x = list(BS = otus_BS, Moss = otus_Moss, Soil_Crust = otus_SC, Plants = otus_Pla, Lichen = otus_Lic),
  category.names = c("Bulk Soil", "Moss", "Soil Crust", "Plant", "Lichen"),
  filename = "VennDiagram_ITS_A_fam.png",
  output = TRUE,
    main = "Venn Diagram of Fungal Families on the First Moraines",  # Add title here
  main.cex = 1  # Adjust title size (optional)
)


#Fungi First Second Moraine Venn Diagrams
# Example: Subset phyloseq object based on sample groups
ITS_morB <- subset_samples(La_Gla_ITS_family, Moraine == "B")
Meta_ITS_B <- sample_data(ITS_morB)

ITS_morB <- prune_taxa(taxa_sums(ITS_morB) > 0, ITS_morB)

ntaxa(ITS_morB)

ITS_morB_BS <- subset_samples(ITS_morB, Sample.Type == "Bulk Soil")
ITS_morB_moss <- subset_samples(ITS_morB, Sample.Type == "Moss")
ITS_morB_SC <- subset_samples(ITS_morB, Sample.Type == "Soil Crust")
ITS_morB_Pla <- subset_samples(ITS_morB, Sample.Type == "Plant")
ITS_morB_Lic <- subset_samples(ITS_morB, Sample.Type == "Lichen")

otus_BS <- taxa_names(prune_taxa(taxa_sums(ITS_morB_BS) > 0, ITS_morB_BS))
otus_Moss <- taxa_names(prune_taxa(taxa_sums(ITS_morB_moss) > 0, ITS_morB_moss))
otus_SC <- taxa_names(prune_taxa(taxa_sums(ITS_morB_SC) > 0, ITS_morB_SC))
otus_Pla <- taxa_names(prune_taxa(taxa_sums(ITS_morB_Pla) > 0, ITS_morB_Pla))
otus_Lic <- taxa_names(prune_taxa(taxa_sums(ITS_morB_Lic) > 0, ITS_morB_Lic))

venn <- venn.diagram(
  x = list(BS = otus_BS, Moss = otus_Moss, Soil_Crust = otus_SC, Plants = otus_Pla, Lichen = otus_Lic),
  category.names = c("Bulk Soil", "Moss", "Soil Crust", "Plant", "Lichen"),
  filename = "VennDiagram_ITS_B_fam.png",
  output = TRUE,
    main = "Venn Diagram of Fungal Families on the Second Moraines",  # Add title here
  main.cex = 1  # Adjust title size (optional)
)



#Fungi  Third Moraine Venn Diagrams
# Example: Subset phyloseq object based on sample groups
ITS_morC <- subset_samples(La_Gla_ITS_family, Moraine == "C")
Meta_ITS_C <- sample_data(ITS_morC)

ITS_morC <- prune_taxa(taxa_sums(ITS_morC) > 0, ITS_morC)

ntaxa(ITS_morC)

ITS_morC_BS <- subset_samples(ITS_morC, Sample.Type == "Bulk Soil")
ITS_morC_moss <- subset_samples(ITS_morC, Sample.Type == "Moss")
ITS_morC_SC <- subset_samples(ITS_morC, Sample.Type == "Soil Crust")
ITS_morC_Pla <- subset_samples(ITS_morC, Sample.Type == "Plant")
ITS_morC_Lic <- subset_samples(ITS_morC, Sample.Type == "Lichen")

otus_BS <- taxa_names(prune_taxa(taxa_sums(ITS_morC_BS) > 0, ITS_morC_BS))
otus_Moss <- taxa_names(prune_taxa(taxa_sums(ITS_morC_moss) > 0, ITS_morC_moss))
otus_SC <- taxa_names(prune_taxa(taxa_sums(ITS_morC_SC) > 0, ITS_morC_SC))
otus_Pla <- taxa_names(prune_taxa(taxa_sums(ITS_morC_Pla) > 0, ITS_morC_Pla))
otus_Lic <- taxa_names(prune_taxa(taxa_sums(ITS_morC_Lic) > 0, ITS_morC_Lic))


venn <- venn.diagram(
  x = list(BS = otus_BS, Moss = otus_Moss, Soil_Crust = otus_SC, Plants = otus_Pla, Lichen = otus_Lic),
  category.names = c("Bulk Soil", "Moss", "Soil Crust", "Plant", "Lichen"),
  filename = "VennDiagram_ITS_C_fam.png",
  output = TRUE,
    main = "Venn Diagram of Fungal Families on the Third Moraines",  # Add title here
  main.cex = 1  # Adjust title size (optional)
)




#Bacterial Venn Diagrams

# # Agglomerate by order
# La_Gla_16S_order <- tax_glom(La_Gla_16S_genus_renamed, taxrank = "Order")
# #tax_fix_interactive(La_Gla_16S_order)
# 
# # Convert ASV/OTU names to order names
# taxa_names(La_Gla_16S_order) <- as.character(tax_table(La_Gla_16S_order)[, "Order"])
# 
# Bac_BS <- subset_samples(La_Gla_16S_family, Sample.Type == "Bulk Soil")
# Bac_moss <- subset_samples(La_Gla_16S_family, Sample.Type == "Moss")
# Bac_SC <- subset_samples(La_Gla_16S_family, Sample.Type == "Soil Crust")
# Bac_Pla <- subset_samples(La_Gla_16S_family, Sample.Type == "Plant")
# Bac_Lic <- subset_samples(La_Gla_16S_family, Sample.Type == "Lichen")
# 
# otus_BS <- taxa_names(prune_taxa(taxa_sums(Bac_BS) > 0, Bac_BS))
# otus_Moss <- taxa_names(prune_taxa(taxa_sums(Bac_moss) > 0, Bac_moss))
# otus_SC <- taxa_names(prune_taxa(taxa_sums(Bac_SC) > 0, Bac_SC))
# otus_Pla <- taxa_names(prune_taxa(taxa_sums(Bac_Pla) > 0, Bac_Pla))
# otus_Lic <- taxa_names(prune_taxa(taxa_sums(Bac_Lic) > 0, Bac_Lic))
# 
# venn <- venn.diagram(
#   x = list(BS = otus_BS, Moss = otus_Moss, Soil_Crust = otus_SC, Plants = otus_Pla, Lichen = otus_Lic),
#   category.names = c("Bulk Soil", "Moss", "Soil Crust", "Plant", "Lichen"),
#   filename = "VennDiagram_all_Bac_order.png",
#   output = TRUE,
#     main = "Venn Diagram of Bacterial Orders across all Moraines",  # Add title here
#   main.cex = 1  # Adjust title size (optional)
# )
# 
# #Bacteria First Moraine Venn Diagrams
# # Example: Subset phyloseq object based on sample groups
# Bac_morA <- subset_samples(La_Gla_16S_order, Moraine == "A")
# Meta_Bac_A <- sample_data(Bac_morA)
# 
# Bac_morA <- prune_taxa(taxa_sums(Bac_morA) > 0, Bac_morA)
# 
# ntaxa(Bac_morA)
# 
# Bac_morA_BS <- subset_samples(Bac_morA, Sample.Type == "Bulk Soil")
# Bac_morA_moss <- subset_samples(Bac_morA, Sample.Type == "Moss")
# Bac_morA_SC <- subset_samples(Bac_morA, Sample.Type == "Soil Crust")
# Bac_morA_Pla <- subset_samples(Bac_morA, Sample.Type == "Plant")
# Bac_morA_Lic <- subset_samples(Bac_morA, Sample.Type == "Lichen")
# 
# otus_BS <- taxa_names(prune_taxa(taxa_sums(Bac_morA_BS) > 0, Bac_morA_BS))
# otus_Moss <- taxa_names(prune_taxa(taxa_sums(Bac_morA_moss) > 0, Bac_morA_moss))
# otus_SC <- taxa_names(prune_taxa(taxa_sums(Bac_morA_SC) > 0, Bac_morA_SC))
# otus_Pla <- taxa_names(prune_taxa(taxa_sums(Bac_morA_Pla) > 0, Bac_morA_Pla))
# otus_Lic <- taxa_names(prune_taxa(taxa_sums(Bac_morA_Lic) > 0, Bac_morA_Lic))
# 
# venn <- venn.diagram(
#   x = list(BS = otus_BS, Moss = otus_Moss, Soil_Crust = otus_SC, Plants = otus_Pla, Lichen = otus_Lic),
#   category.names = c("Bulk Soil", "Moss", "Soil Crust", "Plant", "Lichen"),
#   filename = "VennDiagram_16S_A_order.png",
#   output = TRUE,
#     main = "Venn Diagram of Bacterial Orders on the First Moraines",  # Add title here
#   main.cex = 1  # Adjust title size (optional)
# )
# 
# #Bacteria Second Moraine Venn Diagrams
# # Example: Subset phyloseq object based on sample groups
# Bac_morB <- subset_samples(La_Gla_16S_order, Moraine == "B")
# Meta_Bac_B <- sample_data(Bac_morB)
# 
# Bac_morB <- prune_taxa(taxa_sums(Bac_morB) > 0, Bac_morB)
# 
# ntaxa(Bac_morB)
# 
# Bac_morB_BS <- subset_samples(Bac_morB, Sample.Type == "Bulk Soil")
# Bac_morB_moss <- subset_samples(Bac_morB, Sample.Type == "Moss")
# Bac_morB_SC <- subset_samples(Bac_morB, Sample.Type == "Soil Crust")
# Bac_morB_Pla <- subset_samples(Bac_morB, Sample.Type == "Plant")
# Bac_morB_Lic <- subset_samples(Bac_morB, Sample.Type == "Lichen")
# 
# otus_BS <- taxa_names(prune_taxa(taxa_sums(Bac_morB_BS) > 0, Bac_morB_BS))
# otus_Moss <- taxa_names(prune_taxa(taxa_sums(Bac_morB_moss) > 0, Bac_morB_moss))
# otus_SC <- taxa_names(prune_taxa(taxa_sums(Bac_morB_SC) > 0, Bac_morB_SC))
# otus_Pla <- taxa_names(prune_taxa(taxa_sums(Bac_morB_Pla) > 0, Bac_morB_Pla))
# otus_Lic <- taxa_names(prune_taxa(taxa_sums(Bac_morB_Lic) > 0, Bac_morB_Lic))
# 
# venn <- venn.diagram(
#   x = list(BS = otus_BS, Moss = otus_Moss, Soil_Crust = otus_SC, Plants = otus_Pla, Lichen = otus_Lic),
#   category.names = c("Bulk Soil", "Moss", "Soil Crust", "Plant", "Lichen"),
#   filename = "VennDiagram_16S_B_order.png",
#   output = TRUE,
#     main = "Venn Diagram of Bacterial Orders on the Second Moraines",  # Add title here
#   main.cex = 1  # Adjust title size (optional)
# )
# 
# 
# #Bacteria Third Moraine Venn Diagrams
# # Example: Subset phyloseq object based on sample groups
# Bac_morC <- subset_samples(La_Gla_16S_order, Moraine == "C")
# Meta_Bac_C <- sample_data(Bac_morC)
# 
# Bac_morC <- prune_taxa(taxa_sums(Bac_morC) > 0, Bac_morC)
# 
# ntaxa(Bac_morC)
# 
# Bac_morC_BS <- subset_samples(Bac_morC, Sample.Type == "Bulk Soil")
# Bac_morC_moss <- subset_samples(Bac_morC, Sample.Type == "Moss")
# Bac_morC_SC <- subset_samples(Bac_morC, Sample.Type == "Soil Crust")
# Bac_morC_Pla <- subset_samples(Bac_morC, Sample.Type == "Plant")
# Bac_morC_Lic <- subset_samples(Bac_morC, Sample.Type == "Lichen")
# 
# otus_BS <- taxa_names(prune_taxa(taxa_sums(Bac_morC_BS) > 0, Bac_morC_BS))
# otus_Moss <- taxa_names(prune_taxa(taxa_sums(Bac_morC_moss) > 0, Bac_morC_moss))
# otus_SC <- taxa_names(prune_taxa(taxa_sums(Bac_morC_SC) > 0, Bac_morC_SC))
# otus_Pla <- taxa_names(prune_taxa(taxa_sums(Bac_morC_Pla) > 0, Bac_morC_Pla))
# otus_Lic <- taxa_names(prune_taxa(taxa_sums(Bac_morC_Lic) > 0, Bac_morC_Lic))
# 
# venn <- venn.diagram(
#   x = list(BS = otus_BS, Moss = otus_Moss, Soil_Crust = otus_SC, Plants = otus_Pla, Lichen = otus_Lic),
#   category.names = c("Bulk Soil", "Moss", "Soil Crust", "Plant", "Lichen"),
#   filename = "VennDiagram_16S_C_order.png",
#   output = TRUE,
#     main = "Venn Diagram of Bacterial Orders on the Third Moraines",  # Add title here
#   main.cex = 1  # Adjust title size (optional)
# )
# 

```

## Lichens


```{r, echo=FALSE}
La_Gla_lic_ITS <- La_Gla_ITS_genus_renamed %>%
  ps_filter(Sample.Type == "Lichen", .keep_all_taxa = FALSE) 

La_Gla_lic_16S <- La_Gla_16S_genus_renamed %>%
  ps_filter(Sample.Type == "Lichen", .keep_all_taxa = FALSE) 
  
La_Gla_lic_ITS %>%
  comp_barplot("Family", n_taxa = 15, merge_other = FALSE, label = "Sample.Name") +
  facet_wrap(vars(Moraine), scales = "free") + # scales = "free" is IMPORTANT!
  coord_flip() +
  ggtitle(
    "Fungal Composition of Lichen Samples",
    "Separated by moraine"
  ) +
  theme(axis.ticks.y = element_blank(), strip.text = element_text(face = "bold"))

plot_richness(La_Gla_lic_ITS, x="Moraine", color = "Species", measures=c("Chao1", "Shannon"))

La_Gla_lic_16S %>%
  comp_barplot("Phylum", n_taxa = 15, merge_other = FALSE, label = "Sample.Name") +
  facet_wrap(vars(Moraine), scales = "free") + # scales = "free" is IMPORTANT!
  coord_flip() +
  ggtitle(
    "Bacterial Composition of Lichen Samples",
    "Separated by moraine"
  ) +
  theme(axis.ticks.y = element_blank(), strip.text = element_text(face = "bold"))

plot_richness(La_Gla_lic_16S, x="Moraine", color = "Species", measures=c("Chao1", "Shannon"))

#ord_explore(La_Gla_lic_16S)

La_Gla_lic_16S %>%
 tax_transform(rank = "Phylum", trans = "clr") %>%
 # dist_calc(dist = "euclidean") %>%
 ord_calc(
  method = "auto"
 ) %>% 
 ord_plot(
  axes = c(1, 2),
  colour = "Location", fill = "Location",
  shape = "Moraine", alpha = 0.5,
  size = 3,
  plot_taxa = 1:5,
  )+
 scale_shape_girafe_filled()+
  ggtitle(
    "PCA of Bacterial Phyla from Lichens",
  )


#ord_explore(La_Gla_lic_ITS)

La_Gla_lic_ITS %>%
 tax_transform(rank = "Phylum", trans = "clr") %>%
 # dist_calc(dist = "euclidean") %>%
 ord_calc(
  method = "auto"
 ) %>% 
 ord_plot(
  axes = c(1, 2),
  colour = "Location", fill = "Location",
  shape = "Moraine", alpha = 0.5,
  size = 3,
  plot_taxa = 1:5,
  )+
 scale_shape_girafe_filled()+
  ggtitle(
    "PCA of Fungal Phyla from Lichens",
  )  

#Extract OTU table

#remova all taxa with zero reads
La_Gla_lic_ITS_pseudo <- prune_taxa(taxa_sums(La_Gla_lic_ITS) > 0, La_Gla_lic_ITS)

otu_table(La_Gla_lic_ITS_pseudo) <- otu_table(La_Gla_lic_ITS_pseudo) + + 1e-6

La_Gla_lic_16S_pseudo <- prune_taxa(taxa_sums(La_Gla_lic_16S) > 0, La_Gla_lic_16S) 

otu_table(La_Gla_lic_16S_pseudo) <- otu_table(La_Gla_lic_16S_pseudo) + + 1e-6



Fun_otu_pseudo <- as.matrix(otu_table(La_Gla_lic_ITS_pseudo)) 
Fun_otu_pseudo <- as.data.frame(Fun_otu_pseudo)

Bac_otu_pseudo <- as.matrix(otu_table(La_Gla_lic_16S_pseudo)) 
Bac_otu_pseudo <- as.data.frame(Bac_otu_pseudo)

#CLR transformation
Fun_otu_clr <- Fun_otu_pseudo %>%
  microbiome::transform("clr")
Fun_otu_clr <- as.data.frame(Fun_otu_clr)

Bac_otu_clr <- Bac_otu_pseudo %>%
  microbiome::transform("clr")
Bac_otu_clr <- as.data.frame(Bac_otu_clr)

class(Fun_otu_clr)
class(Bac_otu_clr)

fun_meta <- as.matrix(sample_data(La_Gla_lic_ITS_pseudo))
fun_meta <- as.data.frame(fun_meta)

Bac_meta <- as.matrix(sample_data(La_Gla_lic_16S_pseudo))
Bac_meta <- as.data.frame(Bac_meta)

class(fun_meta)

#convert to numeric and factor
fun_meta$Moraine <- as.factor(fun_meta$Moraine)
fun_meta$Location <- as.factor(fun_meta$Location)
fun_meta$Plot <- as.factor(fun_meta$Plot)

Bac_meta$Moraine <- as.factor(Bac_meta$Moraine)
Bac_meta$Location <- as.factor(Bac_meta$Location)
Bac_meta$Plot <- as.factor(Bac_meta$Plot)

result <- adonis2(Fun_otu_clr ~  Moraine * Location + Plot, data = fun_meta, method="euclidean",  ,perm=999)
print(result)

result <- adonis2(Bac_otu_clr ~  Moraine * Location + Plot, data = Bac_meta, method="euclidean" ,perm=999)
print(result)


```

## Mosses

```{r, echo=FALSE}
La_Gla_moss_ITS <- La_Gla_ITS_genus_renamed %>%
  ps_filter(Sample.Type == "Moss", .keep_all_taxa = FALSE) 

La_Gla_moss_16S <- La_Gla_16S_genus_renamed %>%
  ps_filter(Sample.Type == "Moss", .keep_all_taxa = FALSE) 
  
La_Gla_moss_ITS %>%
  comp_barplot("Family", n_taxa = 15, merge_other = FALSE, label = "Sample.Name") +
  facet_wrap(vars(Moraine), scales = "free") + # scales = "free" is IMPORTANT!
  coord_flip() +
  ggtitle(
    "Fungal Composition of Moss Samples",
    "Separated by moraine"
  ) +
  theme(axis.ticks.y = element_blank(), strip.text = element_text(face = "bold"))

plot_richness(La_Gla_moss_ITS, x="Moraine", color = "Species", measures=c("Chao1", "Shannon"))

La_Gla_moss_16S %>%
  comp_barplot("Phylum", n_taxa = 15, merge_other = FALSE, label = "Sample.Name") +
  facet_wrap(vars(Moraine), scales = "free") + # scales = "free" is IMPORTANT!
  coord_flip() +
  ggtitle(
    "Bacterial Composition of Moss Samples",
    "Separated by moraine"
  ) +
  theme(axis.ticks.y = element_blank(), strip.text = element_text(face = "bold"))

plot_richness(La_Gla_moss_16S, x="Moraine", color = "Species", measures=c("Chao1", "Shannon"))

#ord_explore(La_Gla_moss_16S)

La_Gla_moss_16S %>%
 tax_transform(rank = "Phylum", trans = "clr") %>%
 # dist_calc(dist = "euclidean") %>%
 ord_calc(
  method = "auto"
 ) %>% 
 ord_plot(
  axes = c(1, 2),
  colour = "Location", fill = "Location",
  shape = "Moraine", alpha = 0.5,
  size = 3,
  plot_taxa = 1:5,
  )+
 scale_shape_girafe_filled()+
  ggtitle(
    "PCA of Bacterial Phyla from Moss Samples",
  )


#ord_explore(La_Gla_moss_ITS)

La_Gla_moss_ITS %>%
 tax_transform(rank = "Phylum", trans = "clr") %>%
 # dist_calc(dist = "euclidean") %>%
 ord_calc(
  method = "auto"
 ) %>% 
 ord_plot(
  axes = c(1, 2),
  colour = "Location", fill = "Location",
  shape = "Moraine", alpha = 0.5,
  size = 3,
  plot_taxa = 1:5,
  )+
 scale_shape_girafe_filled()+
  ggtitle(
    "PCA of Fungal Phyla from Moss Samples",
  )


#remova all taxa with zero reads
La_Gla_moss_ITS_pseudo <- prune_taxa(taxa_sums(La_Gla_moss_ITS) > 0, La_Gla_moss_ITS)

otu_table(La_Gla_moss_ITS_pseudo) <- otu_table(La_Gla_moss_ITS_pseudo) + + 1e-6

La_Gla_moss_16S_pseudo <- prune_taxa(taxa_sums(La_Gla_moss_16S) > 0, La_Gla_moss_16S) 

otu_table(La_Gla_moss_16S_pseudo) <- otu_table(La_Gla_moss_16S_pseudo) + + 1e-6



Fun_otu_pseudo <- as.matrix(otu_table(La_Gla_moss_ITS_pseudo)) 
Fun_otu_pseudo <- as.data.frame(Fun_otu_pseudo)

Bac_otu_pseudo <- as.matrix(otu_table(La_Gla_moss_16S_pseudo)) 
Bac_otu_pseudo <- as.data.frame(Bac_otu_pseudo)

#CLR transformation
Fun_otu_clr <- Fun_otu_pseudo %>%
  microbiome::transform("clr")
Fun_otu_clr <- as.data.frame(Fun_otu_clr)

Bac_otu_clr <- Bac_otu_pseudo %>%
  microbiome::transform("clr")
Bac_otu_clr <- as.data.frame(Bac_otu_clr)

class(Fun_otu_clr)
class(Bac_otu_clr)

fun_meta <- as.matrix(sample_data(La_Gla_moss_ITS_pseudo))
fun_meta <- as.data.frame(fun_meta)

Bac_meta <- as.matrix(sample_data(La_Gla_moss_16S_pseudo))
Bac_meta <- as.data.frame(Bac_meta)

class(fun_meta)

#convert to numeric and factor
fun_meta$Moraine <- as.factor(fun_meta$Moraine)
fun_meta$Location <- as.factor(fun_meta$Location)
fun_meta$Plot <- as.factor(fun_meta$Plot)

Bac_meta$Moraine <- as.factor(Bac_meta$Moraine)
Bac_meta$Location <- as.factor(Bac_meta$Location)
Bac_meta$Plot <- as.factor(Bac_meta$Plot)

result <- adonis2(Fun_otu_clr ~  Moraine * Location + Plot, data = fun_meta, method="euclidean", perm=999)
print(result)

result <- adonis2(Bac_otu_clr ~  Moraine * Location + Plot, data = Bac_meta, method="euclidean",,perm=999)
print(result)


```


## Plants

```{r, echo=FALSE}

La_Gla_pla_ITS <- La_Gla_ITS_genus_renamed %>%
  ps_filter(Sample.Type == "Plant", .keep_all_taxa = FALSE) 

La_Gla_pla_16S <- La_Gla_16S_genus_renamed %>%
  ps_filter(Sample.Type == "Plant", .keep_all_taxa = FALSE) 
  
La_Gla_pla_ITS %>%
  comp_barplot("Family", n_taxa = 15, merge_other = FALSE, label = "Sample.Name") +
  facet_wrap(vars(Moraine), scales = "free") + # scales = "free" is IMPORTANT!
  coord_flip() +
  ggtitle(
    "Fungal Composition of Plant Samples",
    "Separated by moraine"
  ) +
  theme(axis.ticks.y = element_blank(), strip.text = element_text(face = "bold"))

plot_richness(La_Gla_pla_ITS, x="Moraine", color = "Species", measures=c("Chao1", "Shannon"))

La_Gla_pla_16S %>%
  comp_barplot("Phylum", n_taxa = 15, merge_other = FALSE, label = "Sample.Name") +
  facet_wrap(vars(Moraine), scales = "free") + #scales = "free" is IMPORTANT!
  coord_flip() +
  ggtitle(
    "Bacterial Composition of plant Samples",
    "Separated by moraine"
  ) +
  theme(axis.ticks.y = element_blank(), strip.text = element_text(face = "bold"))

plot_richness(La_Gla_pla_16S, x="Moraine", color = "Species", measures=c("Chao1", "Shannon"))

#ord_explore(La_Gla_pla_16S)

La_Gla_pla_16S %>%
 tax_transform(rank = "Phylum", trans = "clr") %>%
 # dist_calc(dist = "euclidean") %>%
 ord_calc(
  method = "auto"
 ) %>% 
 ord_plot(
  axes = c(1, 2),
  colour = "Location", fill = "Location",
  shape = "Moraine", alpha = 0.5,
  size = 3,
  plot_taxa = 1:5,
  )+
 scale_shape_girafe_filled()+
  ggtitle(
    "PCA of Bacterial Phyla from Plant Samples",
  )


#ord_explore(La_Gla_pla_ITS)

La_Gla_pla_ITS %>%
 tax_transform(rank = "Phylum", trans = "clr") %>%
 # dist_calc(dist = "euclidean") %>%
 ord_calc(
  method = "auto"
 ) %>% 
 ord_plot(
  axes = c(1, 2),
  colour = "Location", fill = "Location",
  shape = "Moraine", alpha = 0.5,
  size = 3,
  plot_taxa = 1:5,
  )+
 scale_shape_girafe_filled()+
  ggtitle(
    "PCA of Fungal Phyla from Plant Samples",
  )



#remova all taxa with zero reads
La_Gla_pla_ITS_pseudo <- prune_taxa(taxa_sums(La_Gla_pla_ITS) > 0, La_Gla_pla_ITS)

otu_table(La_Gla_pla_ITS_pseudo) <- otu_table(La_Gla_pla_ITS_pseudo) + + 1e-6

La_Gla_pla_16S_pseudo <- prune_taxa(taxa_sums(La_Gla_pla_16S) > 0, La_Gla_pla_16S) 

otu_table(La_Gla_pla_16S_pseudo) <- otu_table(La_Gla_pla_16S_pseudo) + + 1e-6



Fun_otu_pseudo <- as.matrix(otu_table(La_Gla_pla_ITS_pseudo)) 
Fun_otu_pseudo <- as.data.frame(Fun_otu_pseudo)

Bac_otu_pseudo <- as.matrix(otu_table(La_Gla_pla_16S_pseudo)) 
Bac_otu_pseudo <- as.data.frame(Bac_otu_pseudo)

#CLR transformation
Fun_otu_clr <- Fun_otu_pseudo %>%
  microbiome::transform("clr")
Fun_otu_clr <- as.data.frame(Fun_otu_clr)

Bac_otu_clr <- Bac_otu_pseudo %>%
  microbiome::transform("clr")
Bac_otu_clr <- as.data.frame(Bac_otu_clr)

class(Fun_otu_clr)
class(Bac_otu_clr)

fun_meta <- as.matrix(sample_data(La_Gla_pla_ITS_pseudo))
fun_meta <- as.data.frame(fun_meta)

Bac_meta <- as.matrix(sample_data(La_Gla_pla_16S_pseudo))
Bac_meta <- as.data.frame(Bac_meta)

class(fun_meta)

#convert to numeric and factor
fun_meta$Moraine <- as.factor(fun_meta$Moraine)
fun_meta$Location <- as.factor(fun_meta$Location)
fun_meta$Plot <- as.factor(fun_meta$Plot)

Bac_meta$Moraine <- as.factor(Bac_meta$Moraine)
Bac_meta$Location <- as.factor(Bac_meta$Location)
Bac_meta$Plot <- as.factor(Bac_meta$Plot)

result <- adonis2(Fun_otu_clr ~  Moraine * Location + Plot, data = fun_meta, method="euclidean" ,perm=999)
print(result)

result <- adonis2(Bac_otu_clr ~  Moraine * Location + Plot, data = Bac_meta, method="euclidean" ,perm=999)
print(result)




```




## Soil Crust

```{r, echo=FALSE}

La_Gla_SC_ITS <- La_Gla_ITS_genus_renamed %>%
  ps_filter(Sample.Type == "Soil Crust", .keep_all_taxa = FALSE) 

La_Gla_SC_16S <- La_Gla_16S_genus_renamed %>%
  ps_filter(Sample.Type == "Soil Crust", .keep_all_taxa = FALSE) 
  
La_Gla_SC_ITS %>%
  comp_barplot("Family", n_taxa = 15, merge_other = FALSE, label = "Sample.Name") +
  facet_wrap(vars(Moraine), scales = "free") + # scales = "free" is IMPORTANT!
  coord_flip() +
  ggtitle(
    "Fungal Composition of Soil Crust Samples",
    "Separated by moraine"
  ) +
  theme(axis.ticks.y = element_blank(), strip.text = element_text(face = "bold"))

plot_richness(La_Gla_SC_ITS, x="Moraine", color = "Species", measures=c("Chao1", "Shannon"))

La_Gla_SC_16S %>%
  comp_barplot("Phylum", n_taxa = 15, merge_other = FALSE, label = "Sample.Name") +
  facet_wrap(vars(Moraine), scales = "free") + # scales = "free" is IMPORTANT!
  coord_flip() +
  ggtitle(
    "Bacterial Composition of Soil Crust Samples",
    "Separated by moraine"
  ) +
  theme(axis.ticks.y = element_blank(), strip.text = element_text(face = "bold"))

plot_richness(La_Gla_SC_16S, x="Moraine", color = "Species", measures=c("Chao1", "Shannon"))

#ord_explore(La_Gla_SC_16S)

La_Gla_SC_16S %>%
 tax_transform(rank = "Phylum", trans = "clr") %>%
 # dist_calc(dist = "euclidean") %>%
 ord_calc(
  method = "auto"
 ) %>% 
 ord_plot(
  axes = c(1, 2),
  colour = "Location", fill = "Location",
  shape = "Moraine", alpha = 0.5,
  size = 3,
  plot_taxa = 1:5,
  )+
 scale_shape_girafe_filled()+
  ggtitle(
    "PCA of Bacterial Phyla from Soil Crust",
  )


#ord_explore(La_Gla_SC_ITS)

La_Gla_SC_ITS %>%
 tax_transform(rank = "Phylum", trans = "clr") %>%
 # dist_calc(dist = "euclidean") %>%
 ord_calc(
  method = "auto"
 ) %>% 
 ord_plot(
  axes = c(1, 2),
  colour = "Location", fill = "Location",
  shape = "Moraine", alpha = 0.5,
  size = 3,
  plot_taxa = 1:5,
  )+
 scale_shape_girafe_filled()+
  ggtitle(
    "PCA of Fungal Phyla from Soil Crust",
  )




#remova all taxa with zero reads
La_Gla_SC_ITS_pseudo <- prune_taxa(taxa_sums(La_Gla_SC_ITS) > 0, La_Gla_SC_ITS)

otu_table(La_Gla_SC_ITS_pseudo) <- otu_table(La_Gla_SC_ITS_pseudo) + + 1e-6

La_Gla_SC_16S_pseudo <- prune_taxa(taxa_sums(La_Gla_SC_16S) > 0, La_Gla_SC_16S) 

otu_table(La_Gla_SC_16S_pseudo) <- otu_table(La_Gla_SC_16S_pseudo) + + 1e-6



Fun_otu_pseudo <- as.matrix(otu_table(La_Gla_SC_ITS_pseudo)) 
Fun_otu_pseudo <- as.data.frame(Fun_otu_pseudo)

Bac_otu_pseudo <- as.matrix(otu_table(La_Gla_SC_16S_pseudo)) 
Bac_otu_pseudo <- as.data.frame(Bac_otu_pseudo)

#CLR transformation
Fun_otu_clr <- Fun_otu_pseudo %>%
  microbiome::transform("clr")
Fun_otu_clr <- as.data.frame(Fun_otu_clr)

Bac_otu_clr <- Bac_otu_pseudo %>%
  microbiome::transform("clr")
Bac_otu_clr <- as.data.frame(Bac_otu_clr)

class(Fun_otu_clr)
class(Bac_otu_clr)

fun_meta <- as.matrix(sample_data(La_Gla_SC_ITS_pseudo))
fun_meta <- as.data.frame(fun_meta)

Bac_meta <- as.matrix(sample_data(La_Gla_SC_16S_pseudo))
Bac_meta <- as.data.frame(Bac_meta)

class(fun_meta)

#convert to numeric and factor
fun_meta$Moraine <- as.factor(fun_meta$Moraine)
fun_meta$Location <- as.factor(fun_meta$Location)
fun_meta$Plot <- as.factor(fun_meta$Plot)

Bac_meta$Moraine <- as.factor(Bac_meta$Moraine)
Bac_meta$Location <- as.factor(Bac_meta$Location)
Bac_meta$Plot <- as.factor(Bac_meta$Plot)

result <- adonis2(Fun_otu_clr ~  Moraine + Location, data = fun_meta, method="euclidean" ,perm=999)
print(result)

result <- adonis2(Bac_otu_clr ~  Moraine + Location, data = Bac_meta, method="euclidean" ,perm=999)
print(result)



#remova all taxa with zero reads
La_Gla_SC_ITS_pseudo <- prune_taxa(taxa_sums(La_Gla_SC_ITS) > 0, La_Gla_SC_ITS)

otu_table(La_Gla_SC_ITS_pseudo) <- otu_table(La_Gla_SC_ITS_pseudo) + + 1e-6

La_Gla_SC_16S_pseudo <- prune_taxa(taxa_sums(La_Gla_SC_16S) > 0, La_Gla_SC_16S) 

otu_table(La_Gla_SC_16S_pseudo) <- otu_table(La_Gla_SC_16S_pseudo) + + 1e-6



Fun_otu_pseudo <- as.matrix(otu_table(La_Gla_SC_ITS_pseudo)) 
Fun_otu_pseudo <- as.data.frame(Fun_otu_pseudo)

Bac_otu_pseudo <- as.matrix(otu_table(La_Gla_SC_16S_pseudo)) 
Bac_otu_pseudo <- as.data.frame(Bac_otu_pseudo)

#CLR transformation
Fun_otu_clr <- Fun_otu_pseudo %>%
  microbiome::transform("clr")
Fun_otu_clr <- as.data.frame(Fun_otu_clr)

Bac_otu_clr <- Bac_otu_pseudo %>%
  microbiome::transform("clr")
Bac_otu_clr <- as.data.frame(Bac_otu_clr)

class(Fun_otu_clr)
class(Bac_otu_clr)

fun_meta <- as.matrix(sample_data(La_Gla_SC_ITS_pseudo))
fun_meta <- as.data.frame(fun_meta)

Bac_meta <- as.matrix(sample_data(La_Gla_SC_16S_pseudo))
Bac_meta <- as.data.frame(Bac_meta)

class(fun_meta)

#convert to numeric and factor
fun_meta$Moraine <- as.factor(fun_meta$Moraine)
fun_meta$Location <- as.factor(fun_meta$Location)
fun_meta$Plot <- as.factor(fun_meta$Plot)

Bac_meta$Moraine <- as.factor(Bac_meta$Moraine)
Bac_meta$Location <- as.factor(Bac_meta$Location)
Bac_meta$Plot <- as.factor(Bac_meta$Plot)

result <- adonis2(Fun_otu_clr ~  Moraine * Location, data = fun_meta, method="euclidean" ,perm=999)
print(result)

result <- adonis2(Bac_otu_clr ~  Moraine * Location, data = Bac_meta, method="euclidean" ,perm=999)
print(result)



```

##  Bulk Soil

```{r, echo=FALSE}

La_Gla_BS_ITS <- La_Gla_ITS_genus_renamed %>%
  ps_filter(Sample.Type == "Bulk Soil", .keep_all_taxa = FALSE) 

La_Gla_BS_16S <- La_Gla_16S_genus_renamed %>%
  ps_filter(Sample.Type == "Bulk Soil", .keep_all_taxa = FALSE) 
  
La_Gla_BS_ITS %>%
  comp_barplot("Family", n_taxa = 15, merge_other = FALSE, label = "Sample.Name") +
  facet_wrap(vars(Moraine), scales = "free") + # scales = "free" is IMPORTANT!
  coord_flip() +
  ggtitle(
    "Fungal Composition of Bulk Soil Samples",
    "Separated by moraine"
  ) +
  theme(axis.ticks.y = element_blank(), strip.text = element_text(face = "bold"))

plot_richness(La_Gla_BS_ITS, x="Moraine", color = "Species", measures=c("Chao1", "Shannon"))

La_Gla_BS_16S %>%
  comp_barplot("Phylum", n_taxa = 15, merge_other = FALSE, label = "Sample.Name") +
  facet_wrap(vars(Moraine), scales = "free") + # scales = "free" is IMPORTANT!
  coord_flip() +
  ggtitle(
    "Bacterial Composition of Bulk Soil Samples",
    "Separated by moraine"
  ) +
  theme(axis.ticks.y = element_blank(), strip.text = element_text(face = "bold"))

plot_richness(La_Gla_BS_16S, x="Moraine", color = "Species", measures=c("Chao1", "Shannon"))


#ord_explore(La_Gla_BS_16S)

La_Gla_BS_16S %>%
 tax_transform(rank = "Phylum", trans = "clr") %>%
 # dist_calc(dist = "euclidean") %>%
 ord_calc(
  method = "auto"
 ) %>% 
 ord_plot(
  axes = c(1, 2),
  colour = "Location", fill = "Location",
  shape = "Moraine", alpha = 0.5,
  size = 2,
  plot_taxa = 1:7
 ) + 
 scale_shape_girafe_filled()+
  ggtitle(
    "PCoA of Bacterial Phyla from Bulk Soil",
  )


#ord_explore(La_Gla_BS_ITS)

La_Gla_BS_ITS %>%
 tax_transform(rank = "Phylum", trans = "clr") %>%
 # dist_calc(dist = "euclidean") %>%
 ord_calc(
  method = "auto"
 ) %>% 
 ord_plot(
  axes = c(1, 2),
  colour = "Location", fill = "Location",
  shape = "Moraine", alpha = 0.5,
  size = 3,
  plot_taxa = 1:5,
  )+
 scale_shape_girafe_filled()+
  ggtitle(
    "PCA of Fungal Phyla from Bulk Soil",
  )


#remova all taxa with zero reads
La_Gla_BS_ITS_pseudo <- prune_taxa(taxa_sums(La_Gla_BS_ITS) > 0, La_Gla_BS_ITS)

otu_table(La_Gla_BS_ITS_pseudo) <- otu_table(La_Gla_BS_ITS_pseudo) + + 1e-6

La_Gla_BS_16S_pseudo <- prune_taxa(taxa_sums(La_Gla_BS_16S) > 0, La_Gla_BS_16S) 

otu_table(La_Gla_BS_16S_pseudo) <- otu_table(La_Gla_BS_16S_pseudo) + + 1e-6



Fun_otu_pseudo <- as.matrix(otu_table(La_Gla_BS_ITS_pseudo)) 
Fun_otu_pseudo <- as.data.frame(Fun_otu_pseudo)

Bac_otu_pseudo <- as.matrix(otu_table(La_Gla_BS_16S_pseudo)) 
Bac_otu_pseudo <- as.data.frame(Bac_otu_pseudo)

#CLR transformation
Fun_otu_clr <- Fun_otu_pseudo %>%
  microbiome::transform("clr")
Fun_otu_clr <- as.data.frame(Fun_otu_clr)

Bac_otu_clr <- Bac_otu_pseudo %>%
  microbiome::transform("clr")
Bac_otu_clr <- as.data.frame(Bac_otu_clr)

class(Fun_otu_clr)
class(Bac_otu_clr)

fun_meta <- as.matrix(sample_data(La_Gla_BS_ITS_pseudo))
fun_meta <- as.data.frame(fun_meta)

Bac_meta <- as.matrix(sample_data(La_Gla_BS_16S_pseudo))
Bac_meta <- as.data.frame(Bac_meta)

class(fun_meta)

#convert to numeric and factor
fun_meta$Moraine <- as.factor(fun_meta$Moraine)
fun_meta$Location <- as.factor(fun_meta$Location)
fun_meta$Plot <- as.factor(fun_meta$Plot)

Bac_meta$Moraine <- as.factor(Bac_meta$Moraine)
Bac_meta$Location <- as.factor(Bac_meta$Location)
Bac_meta$Plot <- as.factor(Bac_meta$Plot)

result <- adonis2(Fun_otu_clr ~  Moraine + Location, data = fun_meta, method="euclidean" ,perm=999)
print(result)

result <- adonis2(Bac_otu_clr ~  Moraine + Location, data = Bac_meta, method="euclidean" ,perm=999)
print(result)






```

```{r, network analysis results, echo=FALSE}
# 
# setwd("~/PhD/Proj/Ladakh_2022/Glacier Transects/Molecular")
# NetPar_16S <- read_csv("La_Gla_Net_Summary_16S_Parameters.csv")
# 
# NetPar_16S <- subset(NetPar_16S, Num_Samples > 2)
# 
# summary(NetPar_16S)
# colnames(NetPar_16S)
# 
# cols_to_scale <- c("avDiss1",    "avPath1", "clustCoef1", "modularity1",  "vertConnect1", "edgeConnect1", "natConnect1",  "density1",  "pep1") 
# 
# NetPar_16S_scale <- as.data.frame(NetPar_16S %>%
#   mutate(across(all_of(cols_to_scale), scale)))
# 
# summary(NetPar_16S_scale)
# 
# hist(NetPar_16S$avDiss1)
# hist(NetPar_16S_scale$avDiss1)
# 
# hist(NetPar_16S$avPath1)
# hist(NetPar_16S_scale$avPath1)
# 
# hist(NetPar_16S$clustCoef1)
# hist(NetPar_16S_scale$clustCoef1)
# 
# hist(NetPar_16S$modularity1)
# hist(NetPar_16S_scale$modularity1)
# 
# hist(NetPar_16S$vertConnect1)
# hist(NetPar_16S_scale$vertConnect1)
# 
# hist(NetPar_16S$edgeConnect1)
# hist(NetPar_16S_scale$edgeConnect1)
# 
# hist(NetPar_16S$natConnect1)
# hist(NetPar_16S_scale$natConnect1)
# 
# hist(NetPar_16S$density1)
# hist(NetPar_16S_scale$density1)
# 
# hist(NetPar_16S$pep1)
# hist(NetPar_16S_scale$pep1)
# 
# meta <- as.data.frame(as.matrix(meta))
# class(meta)
# merged <- merge(meta,NetPar_16S_scale, by = "Plot")
# 



```


```{r, rda}

```
##RDA with soil as explanatory variables
```{r, echo= FALSE}
# 
# Arizona_clean_16S_rare_forsoil_RDA <- Arizona_clean_16S_rare %>%
#   ps_filter(!is.na(pH), .keep_all_taxa = TRUE)
#  
# 
# 
# # Agglomerate at Genus level (example)
# Arizona_clean_16S_rare_genus <- Arizona_clean_16S_rare_forsoil_RDA %>%
#   tax_transform(rank = "Genus", trans = "hellinger")
# 
# #tax_fix_interactive(Arizona_clean_16S_rare_genus)
# 
# # Agglomerate at Phylum level (example)
# Arizona_clean_16S_rare_phylum <- Arizona_clean_16S_rare_forsoil_RDA %>%
#   tax_transform(rank = "Phylum", trans = "hellinger")
# 
# # Convert OTU table to matrix
# otu_table_matrix_phylum <- as(otu_table(Arizona_clean_16S_rare_phylum), "matrix")  # Change to Ps_obj_genus if needed
# Phylum_hell <- as.data.frame(decostand (otu_table_matrix_phylum, 'hell'))
# 
# # Convert sample metadata to a data frame
# metadata_df <- as.data.frame((sample_data(Arizona_clean_16S_rare_phylum)))
# 
# # Ensure sample names match
# metadata_df <- metadata_df[rownames(metadata_df) %in% rownames(otu_table_matrix_phylum), ]
# 
# # Select relevant environmental variables
# metadata_selected <- as.data.frame(metadata_df[, c("pH" ,                  "DW_g_g" ,              "OM_g_g" ,              "course_fraction_perc",
#  "log_N.NH4_m_g",        "log_P_PO4_g_g" ,       "log_TN_g_g"  ,         "log_TP_g_g" ,          "log_Mg_mg_g",          "log_K_mg_g" ,         
#  "log_Ca_mg_g")])  # Modify based on dataset
# 
# 
# metadata_scale <- as.data.frame(scale(metadata_selected))
# 
# 
# rda_result_phylum <- rda(Phylum_hell ~ ., data = metadata_scale[c("pH", "DW_g_g", "OM_g_g", 
#                                            "course_fraction_perc", "log_N.NH4_m_g", 
#                                            "log_P_PO4_g_g", "log_TN_g_g", "log_TP_g_g", 
#                                            "log_Mg_mg_g", "log_K_mg_g", "log_Ca_mg_g")])
# 
# # View summary
# #summary(rda_result_phylum)
# anova(rda_result_phylum)
# 
# plot(rda_result_phylum, scaling=2)  # Basic RDA plot
# 
# 
# # Extract RDA site scores (samples)
# rda_sites <- scores(rda_result_phylum, display="sites")
# rda_sites_df <- as.data.frame(rda_sites)
# rda_sites_df$Sample <- rownames(rda_sites_df)
# 
# # Extract RDA explanatory variable scores (arrows)
# rda_env <- scores(rda_result_phylum, display="bp")  # "bp" for biplot scores
# rda_env_df <- as.data.frame(rda_env)
# rda_env_df$Variable <- rownames(rda_env_df)
# 
# # Merge site scores with metadata
# rda_sites_df <- merge(rda_sites_df, metadata_scale, by.x="Sample", by.y="row.names")
# 
# # Plot with explanatory variable arrows
# ggplot() +
#   # Plot RDA samples
#   geom_point(data=rda_sites_df, aes(x=RDA1, y=RDA2, color = metadata_df$Latitude), size=3) +
#   
#   # Add arrows for explanatory variables
#   geom_segment(data=rda_env_df, aes(x=0, y=0, xend=RDA1, yend=RDA2),
#                arrow=arrow(length=unit(0.3, "cm")), color="black") +
#   
#   # Add text labels for explanatory variables
#   geom_text(data=rda_env_df, aes(x=RDA1, y=RDA2, label=Variable), 
#             hjust=0.5, vjust=-0.5, size=3, color="black") +
# 
#   theme_minimal() +
#   labs(title="RDA Ordination of Bacterial Genera with Soil Chemistry as Explanatory Variables", 
#        x="RDA1", y="RDA2")
# 
# 
# 
# print("OVERALL TEST OF RDA OF BACTERIAL PHYLA AND SOIL CHEMISTY")
# anova(rda_result_phylum, permutations=999)  # Overall test
# print("TESTING OF INDIVIDUAL SOIL VARIABLES ")
# anova(rda_result_phylum, by="terms", permutations=999)  # Test individual variables
# 
# #####GENUS NOW
# # Convert OTU table to matrix
# otu_table_matrix_genus <- as(otu_table(Arizona_clean_16S_rare_genus), "matrix")  # Change to Ps_obj_genus if needed
# genus_hell <- as.data.frame(decostand (otu_table_matrix_genus, 'hell'))
# 
# # Convert sample metadata to a data frame
# metadata_df <- as.data.frame((sample_data(Arizona_clean_16S_rare_genus)))
# 
# # Ensure sample names match
# metadata_df <- metadata_df[rownames(metadata_df) %in% rownames(otu_table_matrix_genus), ]
# 
# # Select relevant environmental variables
# metadata_selected <- as.data.frame(metadata_df[, c("pH" ,                  "DW_g_g" ,              "OM_g_g" ,              "course_fraction_perc",
#  "log_N.NH4_m_g",        "log_P_PO4_g_g" ,       "log_TN_g_g"  ,         "log_TP_g_g" ,          "log_Mg_mg_g",          "log_K_mg_g" ,         
#  "log_Ca_mg_g")])  # Modify based on dataset
# 
# 
# metadata_scale <- as.data.frame(scale(metadata_selected))
# 
# 
# rda_result_genus <- rda(genus_hell ~ ., data = metadata_scale[c("pH", "DW_g_g", "OM_g_g", 
#                                            "course_fraction_perc", "log_N.NH4_m_g", 
#                                            "log_P_PO4_g_g", "log_TN_g_g", "log_TP_g_g", 
#                                            "log_Mg_mg_g", "log_K_mg_g", "log_Ca_mg_g")])
# 
# # View summary
# #summary(rda_result_genus)
# anova(rda_result_genus)
# 
# plot(rda_result_genus, scaling=2)  # Basic RDA plot
# 
# 
# # Extract RDA site scores (samples)
# rda_sites <- scores(rda_result_genus, display="sites")
# rda_sites_df <- as.data.frame(rda_sites)
# rda_sites_df$Sample <- rownames(rda_sites_df)
# 
# # Extract RDA explanatory variable scores (arrows)
# rda_env <- scores(rda_result_genus, display="bp")  # "bp" for biplot scores
# rda_env_df <- as.data.frame(rda_env)
# rda_env_df$Variable <- rownames(rda_env_df)
# 
# # Merge site scores with metadata
# rda_sites_df <- merge(rda_sites_df, metadata_scale, by.x="Sample", by.y="row.names")
# 
# # Plot with explanatory variable arrows
# ggplot() +
#   # Plot RDA samples
#   geom_point(data=rda_sites_df, aes(x=RDA1, y=RDA2, color = metadata_df$Latitude), size=3) +
#   
#   # Add arrows for explanatory variables
#   geom_segment(data=rda_env_df, aes(x=0, y=0, xend=RDA1, yend=RDA2),
#                arrow=arrow(length=unit(0.3, "cm")), color="black") +
#   
#   # Add text labels for explanatory variables
#   geom_text(data=rda_env_df, aes(x=RDA1, y=RDA2, label=Variable), 
#             hjust=0.5, vjust=-0.5, size=3, color="black") +
# 
#   theme_minimal() +
#   labs(title="RDA Ordination of Bacterial Genera with Soil as Explanatory Variables", 
#        x="RDA1", y="RDA2")
# 
# 
# 
# print("OVERALL TEST OF RDA OF BACTERIAL PHYLA AND SOIL CHEMISTY")
# anova(rda_result_genus, permutations=999)  # Overall test
# print("TESTING OF INDIVIDUAL SOIL VARIABLES ")
# anova(rda_result_genus, by="terms", permutations=999)  # Test individual variables
# 




```
