---
title: "Moraine-pooled composition bar plots by source (16S Class / ITS Order, REVISION)"
author: "Adam Ruka"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

## Overview

Composition stacked bar plots for each **Sample.Type** (source), with samples **pooled by
moraine age** (A, B, C). Uses the **full filtered phyloseq** objects created in
`Microbe_explore.Rmd`, not the later core/shared-taxa subsets.

| Marker | Taxonomic rank | Plotting method |
|--------|----------------|-----------------|
| 16S    | Class          | `microViz::comp_barplot()` (as in `Microbe_explore.Rmd`) |
| ITS    | Order          | same |

**Workflow (per source × marker):**

1. Load the full filtered phyloseq objects and deduplicate by `Sample.Name`.
2. Remove unclassified taxa at the display rank (16S: Class; ITS: Order).
3. `phyloseq::tax_glom()` to Class (16S) or Order (ITS).
4. `phyloseq::merge_samples()` — sum counts for each source × moraine combination.
5. Call `comp_barplot()` once per marker, so every source uses the same global top-taxa
   selection and colour mapping.
6. Combine the 16S and ITS panels into one figure and export **SVG**.

**Inputs**

- `Molecular/ITS/La_Gla_final_16S.Rds` (or `Molecular/La_Gla_final_16S.Rds`)
- `Molecular/ITS/La_Gla_final_ITS.Rds` (or `Molecular/La_Gla_final_ITS.Rds`)

**Outputs**

- `Molecular/Composition_moraine_pooled_REVISION/Composition_moraine_pooled_all_sources_REVISION.svg`
- `Molecular/Composition_moraine_pooled_REVISION/Composition_moraine_pooled_sample_counts_REVISION.csv`

```{r project-setup}
if (!file.exists("Glacier Transects.Rproj") && file.exists("../Glacier Transects.Rproj")) {
  setwd("..")
}
stopifnot("Open Glacier Transects.Rproj first" = file.exists("Glacier Transects.Rproj"))

project_root <- normalizePath(getwd(), winslash = "/", mustWork = TRUE)
molecular_dir <- normalizePath(file.path(project_root, "Molecular"), winslash = "/", mustWork = TRUE)
out_dir <- file.path(molecular_dir, "Composition_moraine_pooled_REVISION")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

revision_tag <- "REVISION"
n_taxa_display <- 15L
merge_other <- FALSE

source_levels <- c("Lichen", "Soil Crust", "Moss", "Bulk Soil", "Plant")
moraine_levels <- c("A", "B", "C")
moraine_labels <- c(
  A = "Young (A)",
  B = "Intermediate (B)",
  C = "Old (C)"
)

suppressPackageStartupMessages({
  library(phyloseq)
  #library(speedyseq)
  library(microViz)
  library(ggplot2)
  library(dplyr)
  library(tidyr)
  library(readr)
  library(svglite)
  library(patchwork)
  library(knitr)
  library(kableExtra)
})

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

cat("Project root:", project_root, "\n")
cat("Output dir:", out_dir, "\n")
```

```{r helpers}
pick_existing_file <- function(candidates, label = "file") {
  hits <- candidates[file.exists(candidates)]
  if (length(hits) == 0) {
    stop(
      "Could not find ", label, ". Tried:\n",
      paste0("  - ", candidates, collapse = "\n")
    )
  }
  normalizePath(hits[[1]], winslash = "/", mustWork = FALSE)
}

source_to_slug <- function(x) {
  x %>%
    tolower() %>%
    gsub("[^a-z0-9]+", "_", .) %>%
    gsub("^_|_$", "", .)
}

sample_data_df <- function(ps) {
  out <- as.data.frame(
    as.matrix(phyloseq::sample_data(ps)),
    stringsAsFactors = FALSE,
    check.names = FALSE
  )
  rownames(out) <- phyloseq::sample_names(ps)
  out
}

set_sample_type_factor <- function(ps) {
  meta <- sample_data_df(ps)
  meta$Sample.Type <- factor(as.character(meta$Sample.Type), levels = source_levels)
  sample_data(ps) <- sample_data(meta)
  ps
}

dedupe_by_sample_name <- function(ps) {
  meta <- sample_data_df(ps)
  stopifnot("Sample.Name" %in% colnames(meta))
  keep <- !duplicated(meta$Sample.Name)
  prune_samples(rownames(meta)[keep], ps)
}

remove_unclassified_at_rank <- function(ps, taxrank) {
  tt <- as.data.frame(as.matrix(tax_table(ps)), stringsAsFactors = FALSE)
  stopifnot(taxrank %in% colnames(tt))
  bad <- tt[[taxrank]] %in% c("Unclassified", "NA", "") | is.na(tt[[taxrank]])
  prune_taxa(!bad, ps)
}

pool_phyloseq_by_source_moraine <- function(ps) {
  meta <- sample_data_df(ps)
  stopifnot(all(c("Sample.Type", "Moraine") %in% colnames(meta)))

  meta$Sample.Type <- as.character(meta$Sample.Type)
  meta$Moraine <- as.character(meta$Moraine)
  keep <- meta$Sample.Type %in% source_levels & meta$Moraine %in% moraine_levels
  ps <- prune_samples(rownames(meta)[keep], ps)
  meta <- meta[keep, , drop = FALSE]

  meta$Pool_ID <- paste(
    source_to_slug(meta$Sample.Type),
    meta$Moraine,
    sep = "__"
  )
  sample_data(ps) <- sample_data(meta)
  ps_pooled <- merge_samples(ps, "Pool_ID")

  pool_lookup <- meta %>%
    distinct(.data$Pool_ID, .data$Sample.Type, .data$Moraine) %>%
    mutate(
      Sample.Type = factor(.data$Sample.Type, levels = source_levels),
      Moraine = factor(.data$Moraine, levels = moraine_levels),
      Moraine_label = unname(moraine_labels[as.character(.data$Moraine)])
    ) %>%
    arrange(.data$Sample.Type, .data$Moraine)

  pool_meta <- as.data.frame(pool_lookup)
  rownames(pool_meta) <- pool_meta$Pool_ID
  pool_meta <- pool_meta[sample_names(ps_pooled), , drop = FALSE]
  sample_data(ps_pooled) <- sample_data(pool_meta)
  ps_pooled
}

prepare_moraine_pooled_ps <- function(ps, taxrank) {
  ps <- dedupe_by_sample_name(ps)
  ps <- prune_taxa(taxa_sums(ps) > 0, ps)
  ps <- remove_unclassified_at_rank(ps, taxrank)
  ps <- tax_glom(ps, taxrank = taxrank, NArm = TRUE)
  pool_phyloseq_by_source_moraine(ps)
}

comp_barplot_moraine_pooled <- function(
    ps_pooled,
    taxrank,
    title_main
) {
  ps_pooled %>%
    comp_barplot(
      taxrank,
      n_taxa = n_taxa_display,
      merge_other = merge_other,
      label = "Moraine_label"
    ) +
    facet_wrap(vars(Sample.Type), ncol = 1, scales = "free_y") +
    coord_flip() +
    labs(title = title_main, x = NULL, y = "Relative abundance") +
    theme(
      axis.ticks.y = element_blank(),
      strip.text = element_text(face = "bold"),
      legend.position = "right"
    )
}

save_comp_svg <- function(plot_obj, file_path, width = 10, height = 6) {
  ggsave(
    filename = file_path,
    plot = plot_obj,
    width = width,
    height = height,
    device = svglite
  )
  invisible(file_path)
}

count_samples_by_source_moraine <- function(ps) {
  sample_data_df(ps) %>%
    transmute(
      Sample.Type = as.character(.data$Sample.Type),
      Moraine = as.character(.data$Moraine)
    ) %>%
    filter(
      .data$Sample.Type %in% source_levels,
      .data$Moraine %in% moraine_levels
    ) %>%
    count(.data$Sample.Type, .data$Moraine, name = "n_samples") %>%
    complete(
      Sample.Type = source_levels,
      Moraine = moraine_levels,
      fill = list(n_samples = 0L)
    ) %>%
    arrange(.data$Sample.Type, factor(.data$Moraine, levels = moraine_levels))
}
```

```{r load-full-filtered}
ps_16s_path <- pick_existing_file(
  c(
    file.path(molecular_dir, "ITS", "La_Gla_final_16S.Rds"),
    file.path(molecular_dir, "La_Gla_final_16S.Rds")
  ),
  label = "full filtered 16S phyloseq RDS"
)

ps_its_path <- pick_existing_file(
  c(
    file.path(molecular_dir, "ITS", "La_Gla_final_ITS.Rds"),
    file.path(molecular_dir, "La_Gla_final_ITS.Rds")
  ),
  label = "full filtered ITS phyloseq RDS"
)

La_Gla_16S_full <- readRDS(ps_16s_path) %>% set_sample_type_factor()
La_Gla_ITS_full <- readRDS(ps_its_path) %>% set_sample_type_factor()

cat("16S full filtered:", ps_16s_path, "\n")
cat("ITS full filtered:", ps_its_path, "\n")
cat("16S — samples:", nsamples(La_Gla_16S_full), "| taxa:", ntaxa(La_Gla_16S_full), "\n")
cat("ITS — samples:", nsamples(La_Gla_ITS_full), "| taxa:", ntaxa(La_Gla_ITS_full), "\n")
```

## Sample counts before pooling

```{r sample-counts}
sample_counts <- bind_rows(
  count_samples_by_source_moraine(La_Gla_16S_full) %>% mutate(marker = "16S"),
  count_samples_by_source_moraine(La_Gla_ITS_full) %>% mutate(marker = "ITS")
) %>%
  relocate(.data$marker)

write_csv(
  sample_counts,
  file.path(out_dir, sprintf("Composition_moraine_pooled_sample_counts_%s.csv", revision_tag))
)

sample_counts %>%
  kable(caption = "Samples per source × moraine (before pooling)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## Composition plots

Each marker is plotted from one pooled phyloseq object and one `comp_barplot()` call.
Consequently, taxon colours and the top-15 selection are identical across all five
sources within that marker. Bacterial classes and fungal orders have separate legends
because they are different taxonomic sets.

```{r plot-all-sources, fig.width=14, fig.height=28}
ps_bac_pooled <- prepare_moraine_pooled_ps(La_Gla_16S_full, "Class")
ps_fun_pooled <- prepare_moraine_pooled_ps(La_Gla_ITS_full, "Order")

p_bac <- comp_barplot_moraine_pooled(
  ps_bac_pooled,
  "Class",
  title_main = "A. Bacterial composition (Class)"
)

p_fun <- comp_barplot_moraine_pooled(
  ps_fun_pooled,
  "Order",
  title_main = "B. Fungal composition (Order)"
)

p_combined <- p_bac / p_fun +
  plot_annotation(
    title = "Microbial composition across moraine ages by source",
    subtitle = "Counts pooled across all samples within each source × moraine combination"
  )

combined_svg <- file.path(
  out_dir,
  sprintf("Composition_moraine_pooled_all_sources_%s.svg", revision_tag)
)
save_comp_svg(p_combined, combined_svg, width = 14, height = 28)
print(p_combined)

plot_index <- tibble(
  figure = "Combined 16S Class + ITS Order",
  n_pooled_bars_16S = nsamples(ps_bac_pooled),
  n_pooled_bars_ITS = nsamples(ps_fun_pooled),
  svg = combined_svg
)
write_csv(
  plot_index,
  file.path(out_dir, sprintf("Composition_moraine_pooled_plot_index_%s.csv", revision_tag))
)

plot_index %>%
  kable(caption = "Exported composition SVGs") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```
