---
title: "Surface cover change across glacier chronosequences — Ladakh 2022"
author: "Adam Ruka"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

Analyses changes in surface cover types (lichen, moss, soil crust, forb, graminoid,
bare soil, stones) across the moraine-age chronosequence (A = young → B = intermediate
→ C = old) at two glacier locations (Nubra, Tsomoriri).

**Input:** `Glacier_Soil_Topo_merged.xlsx` (project root)

**Open `Glacier Transects.Rproj`**, then Knit.

```{r project-setup, message=FALSE, warning=FALSE}
if (!file.exists("Glacier Transects.Rproj") && file.exists("../Glacier Transects.Rproj")) {
  setwd("..")
}
stopifnot("Open Glacier Transects.Rproj first" = file.exists("Glacier Transects.Rproj"))

project_root  <- normalizePath(getwd(), winslash = "/", mustWork = FALSE)
molecular_dir <- normalizePath(file.path(project_root, "Molecular"),
                               winslash = "/", mustWork = FALSE)

suppressPackageStartupMessages({
  library(tidyverse)
  library(readxl)
  library(multcompView)
  library(knitr)
  library(kableExtra)
  library(ggplot2)
  library(patchwork)
  library(lavaan)
  library(piecewiseSEM)
  library(svglite)
})

# Illustrator-friendly SVG export (plain text/lines; avoids grDevices filter stacks).
save_gg_svg <- function(filename, plot, width, height, bg = "white") {
  ggsave(
    filename = filename,
    plot = plot,
    width = width,
    height = height,
    device = svglite,
    bg = bg
  )
}

cat("Project root: ", project_root,  "\n")
cat("Molecular dir:", molecular_dir, "\n")

# ── Cover variable names (long labels for plotting) ──────────────────────────
cover_vars <- c(
  "Total_lichen_cover",
  "Moss_cover",
  "Crust_cover",
  "Forb_cover",
  "Gaminoid_cover",
  "Bare_Soil_Total",
  "Stones"
)

cover_labels <- c(
  Total_lichen_cover = "Lichen cover (%)",
  Moss_cover         = "Moss cover (%)",
  Crust_cover        = "Soil crust cover (%)",
  Forb_cover         = "Forb cover (%)",
  Gaminoid_cover     = "Graminoid cover (%)",
  Bare_Soil_Total    = "Bare soil (%)",
  Stones             = "Stone cover (%)"
)

moraine_labels <- c(A = "Young (A)", B = "Intermediate (B)", C = "Old (C)")
moraine_colours <- c(
  A = "#8C510A", # young — brown
  B = "#A6D96A", # intermediate — light green
  C = "#1B7837"  # old — dark green
)
location_shapes <- c(
  Nubra = 21,      # circle
  Tsomoriri = 24   # triangle
)
alpha <- 0.05
```

## Load metadata

```{r load-metadata}
meta_file <- file.path(project_root, "Glacier_Soil_Topo_merged.xlsx")
stopifnot("Glacier_Soil_Topo_merged.xlsx not found at project root" = file.exists(meta_file))

plot_meta_raw <- read_excel(meta_file)

cat("Rows:", nrow(plot_meta_raw), "| Columns:", ncol(plot_meta_raw), "\n")
cat("Column names:\n")
print(colnames(plot_meta_raw))
```

## Prepare data

```{r prepare-data}
derive_glacier_id <- function(plot) {
  out <- as.character(plot)
  # Handles IDs such as NUB1-C-1 or Nubra_1_A by dropping moraine/replicate suffixes.
  out <- sub("[-_][ABC][-_][0-9]+$", "", out)
  out <- sub("[-_][ABC]$", "", out)
  out
}

plot_meta <- plot_meta_raw
if (!"GlacierID" %in% names(plot_meta)) {
  plot_meta <- plot_meta %>%
    mutate(GlacierID = derive_glacier_id(Plot))
}

plot_meta <- plot_meta %>%
  mutate(
    Moraine  = factor(Moraine,  levels = c("A", "B", "C"), labels = c("A", "B", "C")),
    Moraine_num = as.integer(Moraine),
    Location = factor(Location, levels = c("Nubra", "Tsomoriri")),
    GlacierID = factor(GlacierID),
    # Glacier × moraine random intercept (matches molecular analyses)
    Glacier_Moraine = factor(Plot),
    Aspect_sin = sin(Aspect * pi / 180),
    Aspect_cos = cos(Aspect * pi / 180)
  ) %>%
  # Ensure all cover columns are numeric
  mutate(across(all_of(cover_vars), as.numeric))

# Quick preview
plot_meta %>%
  select(Plot, Location, Moraine, all_of(cover_vars)) %>%
  arrange(Location, Moraine) %>%
  kable(digits = 1, caption = "Plot metadata — cover variables") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## Descriptive summary

Mean (± SE) cover by moraine age and location.

```{r descriptive-summary}
cover_long <- plot_meta %>%
  select(Plot, Location, Moraine, Glacier_Moraine, all_of(cover_vars)) %>%
  pivot_longer(
    cols      = all_of(cover_vars),
    names_to  = "cover_type",
    values_to = "pct"
  ) %>%
  mutate(
    cover_type = factor(cover_type, levels = cover_vars),
    cover_label = cover_labels[as.character(cover_type)]
  )

cover_summary <- cover_long %>%
  group_by(cover_type, cover_label, Location, Moraine) %>%
  summarise(
    n    = n(),
    mean = mean(pct, na.rm = TRUE),
    se   = sd(pct, na.rm = TRUE) / sqrt(sum(!is.na(pct))),
    .groups = "drop"
  )

cover_summary %>%
  mutate(
    `Mean ± SE` = paste0(round(mean, 1), " ± ", round(se, 1))
  ) %>%
  select(cover_label, Location, Moraine, n, `Mean ± SE`) %>%
  pivot_wider(names_from = Moraine, values_from = c(n, `Mean ± SE`)) %>%
  kable(caption = "Mean (± SE) cover (%) by cover type, location, and moraine age") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## Visualisation — cover across chronosequence

```{r cover-violin-plots, fig.width=10, fig.height=14}
p_list <- lapply(cover_vars, function(cv) {
  d <- cover_long %>%
    filter(cover_type == cv) %>%
    mutate(Moraine = factor(Moraine, levels = c("A", "B", "C")))

  ggplot(d, aes(x = Moraine, y = pct, fill = Moraine)) +
    geom_violin(trim = FALSE, scale = "width", alpha = 0.7, colour = NA) +
    geom_boxplot(width = 0.12, fill = "white", outlier.shape = NA, colour = "grey30") +
    geom_jitter(
      aes(shape = Location),
      width = 0.08, size = 1.8, alpha = 0.75,
      colour = "grey20"
    ) +
    facet_wrap(~ Location) +
    scale_fill_manual(values = moraine_colours) +
    scale_shape_manual(values = location_shapes) +
    scale_x_discrete(labels = moraine_labels) +
    labs(
      title = cover_labels[[cv]],
      x     = "Moraine age",
      y     = "Cover (%)"
    ) +
    theme_bw(base_size = 11) +
    theme(
      legend.position = "none",
      strip.background = element_rect(fill = "grey92"),
      plot.title = element_text(face = "bold", size = 11)
    )
})

wrap_plots(p_list, ncol = 2)
```

## Statistical analysis

Cover data are plot-level measurements (one row per plot), so a random-intercept LMER
is not appropriate. Instead:

- **Moraine effect** — Kruskal-Wallis test (non-parametric; small *n* per group) with
  pairwise Wilcoxon post-hoc and Benjamini–Hochberg FDR across the seven cover types.
- **Location effect** — Wilcoxon rank-sum test (Nubra vs Tsomoriri) with BH-FDR.
- **Environmental drivers** — Spearman *ρ* between each cover type and continuous
  environmental variables; BH-FDR within each cover type.

Environmental variables tested: distance from glacier, elevation, slope, topographic
position index (TPI), aspect (sin/cos), snow cover 1976.

```{r stats-compute}
# ── Environmental variable list ───────────────────────────────────────────────
env_vars <- c(
  "DistanceFromGlacier", "Elevation", "Slope", "TPI",
  "Aspect_sin", "Aspect_cos", "Snow_1976"
)
env_labels <- c(
  DistanceFromGlacier = "Distance from glacier",
  Elevation           = "Elevation (m)",
  Slope               = "Slope (°)",
  TPI                 = "TPI",
  Aspect_sin          = "Aspect (sin)",
  Aspect_cos          = "Aspect (cos)",
  Snow_1976           = "Snow cover 1976"
)

# Coerce env vars to numeric
plot_meta <- plot_meta %>%
  mutate(across(any_of(env_vars), as.numeric))

# ── Helpers ───────────────────────────────────────────────────────────────────
fmt_p <- function(p) {
  ifelse(is.na(p), "—",
    ifelse(p < 0.001, "<0.001",
      ifelse(p < 0.01, formatC(p, digits = 3, format = "f"),
        formatC(p, digits = 3, format = "f"))))
}

sig_stars <- function(p) {
  ifelse(is.na(p), "",
    ifelse(p < 0.001, "***",
      ifelse(p < 0.01, "**",
        ifelse(p < 0.05, "*", ""))))
}

# ── 1. Kruskal-Wallis: Moraine effect ────────────────────────────────────────
kw_results <- lapply(cover_vars, function(cv) {
  x <- plot_meta[[cv]]
  g <- plot_meta$Moraine
  ok <- !is.na(x) & !is.na(g)
  kt <- tryCatch(kruskal.test(x[ok] ~ g[ok]), error = function(e) NULL)
  if (is.null(kt)) return(tibble(cover_type = cv, H = NA_real_, df = NA_real_, p_kw = NA_real_))
  tibble(cover_type = cv, H = kt$statistic, df = kt$parameter, p_kw = kt$p.value)
}) %>% bind_rows() %>%
  mutate(Padj_kw = p.adjust(p_kw, method = "BH"))

# ── 2. Pairwise Wilcoxon: Moraine post-hoc ───────────────────────────────────
pw_moraine_results <- lapply(cover_vars, function(cv) {
  x <- plot_meta[[cv]]
  g <- plot_meta$Moraine
  ok <- !is.na(x) & !is.na(g)
  pw <- tryCatch(
    pairwise.wilcox.test(x[ok], g[ok], p.adjust.method = "none", exact = FALSE),
    error = function(e) NULL
  )
  if (is.null(pw)) return(tibble(cover_type = cv, contrast = NA_character_, p_pw = NA_real_))
  pm <- pw$p.value
  pairs_df <- expand.grid(
    g1 = rownames(pm), g2 = colnames(pm), stringsAsFactors = FALSE
  ) %>%
    filter(!is.na(pm[cbind(g1, g2)])) %>%
    mutate(
      cover_type = cv,
      contrast   = paste(g1, g2, sep = "-"),
      p_pw       = pm[cbind(g1, g2)]
    ) %>%
    select(cover_type, contrast, p_pw)
  pairs_df
}) %>% bind_rows()

# CLD letters from pairwise Wilcoxon (unadjusted p)
cld_moraine <- lapply(cover_vars, function(cv) {
  pw <- pw_moraine_results %>% filter(cover_type == cv, !is.na(p_pw))
  if (!nrow(pw)) return(tibble(cover_type = cv, Moraine = NA_character_, CLD = NA_character_))
  pvals <- setNames(pw$p_pw, pw$contrast)
  ml <- tryCatch(
    multcompView::multcompLetters(pvals, threshold = alpha),
    error = function(e) NULL
  )
  if (is.null(ml)) return(tibble(cover_type = cv, Moraine = NA_character_, CLD = NA_character_))
  tibble(cover_type = cv, Moraine = names(ml$Letters), CLD = trimws(unname(ml$Letters)))
}) %>% bind_rows()

# ── 3. Wilcoxon: Location effect ─────────────────────────────────────────────
loc_results <- lapply(cover_vars, function(cv) {
  x <- plot_meta[[cv]]
  g <- plot_meta$Location
  ok <- !is.na(x) & !is.na(g) & length(unique(g[!is.na(g)])) == 2
  wt <- tryCatch(
    wilcox.test(x[ok] ~ g[ok], exact = FALSE),
    error = function(e) NULL
  )
  if (is.null(wt)) return(tibble(cover_type = cv, W = NA_real_, p_loc = NA_real_))
  tibble(cover_type = cv, W = wt$statistic, p_loc = wt$p.value)
}) %>% bind_rows() %>%
  mutate(Padj_loc = p.adjust(p_loc, method = "BH"))

# ── 4. Spearman correlations: env drivers ────────────────────────────────────
spearman_results <- bind_rows(lapply(cover_vars, function(cv) {
  bind_rows(lapply(env_vars, function(ev) {
    x <- as.numeric(plot_meta[[cv]])
    y <- as.numeric(plot_meta[[ev]])
    ok <- !is.na(x) & !is.na(y)
    if (sum(ok) < 5) return(tibble(cover_type = cv, env_var = ev, rho = NA_real_, p_sp = NA_real_))
    ct <- tryCatch(cor.test(x[ok], y[ok], method = "spearman", exact = FALSE),
                   error = function(e) NULL)
    if (is.null(ct)) return(tibble(cover_type = cv, env_var = ev, rho = NA_real_, p_sp = NA_real_))
    tibble(cover_type = cv, env_var = ev, rho = ct$estimate, p_sp = ct$p.value)
  }))
})) %>%
  group_by(cover_type) %>%
  mutate(Padj_sp = p.adjust(p_sp, method = "BH")) %>%
  ungroup()
```

## Summary table — moraine & location effects

Kruskal-Wallis H (Moraine) and Wilcoxon W (Location), with BH-FDR across seven cover
types. Group means shown for each moraine age; CLD letters (unadjusted pairwise Wilcoxon,
α = 0.05) indicate significant differences.

```{r summary-table-moraine}
# Group means per moraine
moraine_means <- cover_long %>%
  group_by(cover_type, Moraine) %>%
  summarise(mean_pct = round(mean(pct, na.rm = TRUE), 1), .groups = "drop") %>%
  pivot_wider(names_from = Moraine, values_from = mean_pct,
              names_prefix = "Mean_")

# Join CLD letters
cld_wide <- cld_moraine %>%
  filter(!is.na(Moraine)) %>%
  pivot_wider(names_from = Moraine, values_from = CLD, names_prefix = "CLD_")

# Assemble summary
summary_moraine <- kw_results %>%
  left_join(loc_results %>% select(cover_type, W, p_loc, Padj_loc), by = "cover_type") %>%
  left_join(moraine_means, by = "cover_type") %>%
  left_join(cld_wide,      by = "cover_type") %>%
  mutate(
    cover_label = cover_labels[cover_type],
    # Combine mean + letter into single cell
    `Young (A)`        = paste0(Mean_A, if_else(!is.na(CLD_A), paste0(" ", CLD_A), "")),
    `Intermediate (B)` = paste0(Mean_B, if_else(!is.na(CLD_B), paste0(" ", CLD_B), "")),
    `Old (C)`          = paste0(Mean_C, if_else(!is.na(CLD_C), paste0(" ", CLD_C), "")),
    KW_H    = round(H, 2),
    KW_Padj = fmt_p(Padj_kw),
    KW_sig  = sig_stars(Padj_kw),
    Loc_W   = round(W, 1),
    Loc_Padj = fmt_p(Padj_loc),
    Loc_sig  = sig_stars(Padj_loc)
  ) %>%
  select(
    `Cover type` = cover_label,
    `Young (A)`, `Intermediate (B)`, `Old (C)`,
    `KW H` = KW_H, `KW Padj` = KW_Padj, ` ` = KW_sig,
    `W` = Loc_W, `Loc Padj` = Loc_Padj, `  ` = Loc_sig
  )

summary_moraine %>%
  kable(
    caption = paste0(
      "Mean cover (%) by moraine age with compact-letter display (CLD, unadjusted Wilcoxon α = 0.05). ",
      "KW = Kruskal-Wallis H statistic (Moraine effect, BH-FDR). ",
      "W = Wilcoxon rank-sum statistic (Nubra vs Tsomoriri, BH-FDR). ",
      "*** p < 0.001, ** p < 0.01, * p < 0.05."
    ),
    escape = FALSE, align = "lrrrrrrrrrr"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
  add_header_above(c(" " = 1, "Mean cover (%) + CLD" = 3,
                     "Moraine (KW)" = 3, "Location (Wilcoxon)" = 3)) %>%
  column_spec(7, bold = TRUE) %>%
  column_spec(10, bold = TRUE)
```

## Summary table — environmental drivers (Spearman *ρ*)

Spearman correlation between each cover type and continuous environmental variables.
BH-FDR within cover type across seven environmental variables.
Values shown as *ρ* (FDR-adjusted *p*); bold = significant (Padj < 0.05).

```{r summary-table-env}
# Build wide table: one column per env var, cell = "ρ stars (Padj)"
env_wide <- spearman_results %>%
  mutate(
    cover_label = cover_labels[cover_type],
    env_label   = env_labels[env_var],
    cell = case_when(
      is.na(rho) ~ "—",
      TRUE ~ paste0(
        formatC(rho, digits = 2, format = "f"),
        sig_stars(Padj_sp),
        " (", fmt_p(Padj_sp), ")"
      )
    )
  ) %>%
  select(cover_label, env_label, cell) %>%
  pivot_wider(names_from = env_label, values_from = cell)

env_wide %>%
  rename(`Cover type` = cover_label) %>%
  kable(
    caption = paste0(
      "Spearman \u03c1 with BH-FDR adjusted p in parentheses. ",
      "Stars: *** p<0.001, ** p<0.01, * p<0.05."
    ),
    escape = TRUE, align = "lrrrrrrr"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
  row_spec(
    which(env_wide$cover_label %in% cover_labels[
      spearman_results %>%
        filter(Padj_sp < alpha) %>%
        pull(cover_type) %>%
        unique()
    ]),
    background = "#f0f7ff"
  )
```

## Pairwise Wilcoxon — significant moraine contrasts

Full pairwise tables for cover types with significant Kruskal-Wallis (Padj < 0.05).

```{r pairwise-detail, results='asis'}
sig_kw <- kw_results %>% filter(Padj_kw < alpha) %>% pull(cover_type)

for (cv in sig_kw) {
  lbl <- cover_labels[[cv]]
  pw  <- pw_moraine_results %>%
    filter(cover_type == cv, !is.na(p_pw)) %>%
    mutate(Padj = p.adjust(p_pw, method = "BH"),
           Sig  = sig_stars(Padj)) %>%
    select(Contrast = contrast, `p (unadj)` = p_pw, `Padj (BH)` = Padj, ` ` = Sig) %>%
    mutate(across(where(is.numeric), ~ round(.x, 4)))

  cat(
    pw %>%
      kable(format = "html",
            caption = paste0(lbl, " — pairwise Wilcoxon by moraine age"),
            escape = FALSE) %>%
      kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
      as.character(),
    "\n\n"
  )
}
```

## Cover change along the glacier retreat gradient

Cover variables scaled 0–1 (each variable divided by its within-`GlacierID` maximum) and plotted
against distance from glacier. Point shape indicates location (Nubra = circle;
Tsomoriri = triangle), while fill indicates moraine age (young = brown,
intermediate = light green, old = dark green). Trend line = linear fit ± 95% CI
across both locations combined. Spearman *ρ* and FDR-adjusted *p* (from the table
above) are annotated on each panel; the straight line is a visual summary of trend
direction rather than a separate inferential test.

```{r cover-distance-plots, fig.width = 11, fig.height = 14}
# ── Scale each cover variable 0-1 by the within-GlacierID maximum ─────────────
cover_scaled <- plot_meta %>%
  select(Plot, GlacierID, Location, Moraine, Moraine_num, DistanceFromGlacier, all_of(cover_vars)) %>%
  group_by(GlacierID) %>%
  mutate(across(
    all_of(cover_vars),
    ~ {
      mx <- max(.x, na.rm = TRUE)
      if (is.na(mx) || mx == 0) NA_real_ else .x / mx
    }
  )) %>%
  ungroup() %>%
  pivot_longer(
    cols      = all_of(cover_vars),
    names_to  = "cover_type",
    values_to = "scaled_cover"
  ) %>%
  mutate(
    cover_label = cover_labels[cover_type],
    cover_label = factor(cover_label, levels = cover_labels),
    Moraine     = factor(Moraine, levels = c("A", "B", "C")),
    Location    = factor(Location)
  )

# Pull Spearman annotation per cover type
sp_annot <- spearman_results %>%
  filter(env_var == "DistanceFromGlacier") %>%
  mutate(
    cover_label = cover_labels[cover_type],
    cover_label = factor(cover_label, levels = cover_labels),
    label = paste0(
      "\u03c1 = ", formatC(rho, digits = 2, format = "f"),
      sig_stars(Padj_sp),
      "\nPadj = ", fmt_p(Padj_sp)
    )
  )

p_cover_distance <- ggplot(cover_scaled, aes(x = DistanceFromGlacier, y = scaled_cover)) +
  # Linear trend across both locations combined
  geom_smooth(
    method  = "lm", formula = y ~ x,
    colour  = "grey30", fill = "grey80", linewidth = 0.8, alpha = 0.3,
    na.rm   = TRUE
  ) +
  # Points coloured by moraine age, shaped by location
  geom_point(
    aes(fill = Moraine, shape = Location),
    size = 3, colour = "white", stroke = 0.6, na.rm = TRUE
  ) +
  # Spearman annotation (top-left of each facet)
  geom_text(
    data  = sp_annot,
    aes(label = label),
    x     = -Inf, y = Inf,
    hjust = -0.08, vjust = 1.3,
    size  = 3, colour = "grey20", inherit.aes = FALSE
  ) +
  facet_wrap(~ cover_label, ncol = 2, scales = "free_x") +
  scale_fill_manual(
    values = moraine_colours, name = "Moraine age",
    labels = moraine_labels
  ) +
  scale_shape_manual(values = location_shapes, name = "Location") +
  scale_y_continuous(
    limits = c(0, 1.05), breaks = seq(0, 1, 0.25),
    labels = scales::label_percent(accuracy = 1)
  ) +
  guides(
    fill  = guide_legend(override.aes = list(shape = 21, size = 3.5)),
    shape = guide_legend(override.aes = list(fill = "grey60", size = 3.5))
  ) +
  labs(
    x = "Distance from glacier (m)",
    y = "Scaled cover (proportion of GlacierID maximum)"
  ) +
  theme_bw(base_size = 11) +
  theme(
    strip.background  = element_rect(fill = "grey92"),
    strip.text        = element_text(face = "bold", size = 10),
    legend.position   = "bottom",
    legend.box        = "horizontal",
    panel.grid.minor  = element_blank()
  )

p_cover_distance
```

### Cover change along the glacier retreat gradient — raw values

Same distance-gradient plot as above, but using raw cover percentages rather than
within-`GlacierID` scaled values.

```{r cover-distance-raw-plots, fig.width = 11, fig.height = 14}
cover_raw_distance <- plot_meta %>%
  select(Plot, Location, Moraine, DistanceFromGlacier, all_of(cover_vars)) %>%
  pivot_longer(
    cols      = all_of(cover_vars),
    names_to  = "cover_type",
    values_to = "cover_pct"
  ) %>%
  mutate(
    cover_label = cover_labels[cover_type],
    cover_label = factor(cover_label, levels = cover_labels),
    Moraine     = factor(Moraine, levels = c("A", "B", "C")),
    Location    = factor(Location)
  )

p_cover_distance_raw <- ggplot(cover_raw_distance, aes(x = DistanceFromGlacier, y = cover_pct)) +
  geom_smooth(
    method  = "lm", formula = y ~ x,
    colour  = "grey30", fill = "grey80", linewidth = 0.8, alpha = 0.3,
    na.rm   = TRUE
  ) +
  geom_point(
    aes(fill = Moraine, shape = Location),
    size = 3, colour = "white", stroke = 0.6, na.rm = TRUE
  ) +
  geom_text(
    data  = sp_annot,
    aes(label = label),
    x     = -Inf, y = Inf,
    hjust = -0.08, vjust = 1.3,
    size  = 3, colour = "grey20", inherit.aes = FALSE
  ) +
  facet_wrap(~ cover_label, ncol = 2, scales = "free_y") +
  scale_fill_manual(
    values = moraine_colours, name = "Moraine age",
    labels = moraine_labels
  ) +
  scale_shape_manual(values = location_shapes, name = "Location") +
  guides(
    fill  = guide_legend(override.aes = list(shape = 21, size = 3.5)),
    shape = guide_legend(override.aes = list(fill = "grey60", size = 3.5))
  ) +
  labs(
    x = "Distance from glacier (m)",
    y = "Cover (%)"
  ) +
  theme_bw(base_size = 11) +
  theme(
    strip.background  = element_rect(fill = "grey92"),
    strip.text        = element_text(face = "bold", size = 10),
    legend.position   = "bottom",
    legend.box        = "horizontal",
    panel.grid.minor  = element_blank()
  )

p_cover_distance_raw
```

### Cover change along ordinal moraine age

Same scaled cover values as above, but plotted against moraine age treated as an
ordinal numeric gradient (`Young/A = 1`, `Intermediate/B = 2`, `Old/C = 3`).

```{r cover-moraine-age-plots, fig.width = 11, fig.height = 14}
cover_moraine_sp <- cover_scaled %>%
  group_by(cover_type, cover_label) %>%
  group_modify(~ {
    ok <- complete.cases(.x$Moraine_num, .x$scaled_cover)
    if (sum(ok) < 5) {
      return(tibble(rho = NA_real_, p_sp = NA_real_))
    }
    ct <- cor.test(
      .x$Moraine_num[ok],
      .x$scaled_cover[ok],
      method = "spearman",
      exact = FALSE
    )
    tibble(rho = unname(ct$estimate), p_sp = ct$p.value)
  }) %>%
  ungroup() %>%
  mutate(
    Padj_sp = p.adjust(p_sp, method = "BH"),
    label = paste0(
      "\u03c1 = ", formatC(rho, digits = 2, format = "f"),
      sig_stars(Padj_sp),
      "\nPadj = ", fmt_p(Padj_sp)
    )
  )

p_cover_moraine_age <- ggplot(cover_scaled, aes(x = Moraine_num, y = scaled_cover)) +
  geom_smooth(
    method  = "loess", formula = y ~ x, span = 1.5,
    colour  = "grey30", fill = "grey80", linewidth = 0.8, alpha = 0.3,
    na.rm   = TRUE
  ) +
  geom_point(
    aes(fill = Moraine, shape = Location),
    position = position_jitter(width = 0.05, height = 0),
    size = 3, colour = "white", stroke = 0.6, na.rm = TRUE
  ) +
  geom_text(
    data  = cover_moraine_sp,
    aes(label = label),
    x     = -Inf, y = Inf,
    hjust = -0.08, vjust = 1.3,
    size  = 3, colour = "grey20", inherit.aes = FALSE
  ) +
  facet_wrap(~ cover_label, ncol = 2) +
  scale_x_continuous(
    breaks = 1:3,
    labels = c("Young\n(1)", "Intermediate\n(2)", "Old\n(3)")
  ) +
  scale_fill_manual(
    values = moraine_colours, name = "Moraine age",
    labels = moraine_labels
  ) +
  scale_shape_manual(values = location_shapes, name = "Location") +
  scale_y_continuous(
    limits = c(0, 1.05), breaks = seq(0, 1, 0.25),
    labels = scales::label_percent(accuracy = 1)
  ) +
  guides(
    fill  = guide_legend(override.aes = list(shape = 21, size = 3.5)),
    shape = guide_legend(override.aes = list(fill = "grey60", size = 3.5))
  ) +
  labs(
    x = "Moraine age class (Young = 1, Intermediate = 2, Old = 3)",
    y = "Scaled cover (proportion of GlacierID maximum)"
  ) +
  theme_bw(base_size = 11) +
  theme(
    strip.background  = element_rect(fill = "grey92"),
    strip.text        = element_text(face = "bold", size = 10),
    legend.position   = "bottom",
    legend.box        = "horizontal",
    panel.grid.minor  = element_blank()
  )

p_cover_moraine_age
```

## Soil properties along the glacier retreat gradient

Same scaling (0–1 by within-`GlacierID` variable maximum) and plot style as the cover figures above.
Environmental variables: pH, organic matter, texture (>2 mm stones), total N, total P,
and exchangeable cations (Na, Mg, K, Ca).

```{r soil-stats-compute}
soil_vars <- c("pH", "OM%", "texture_>2mm", "TN_%", "TP_%",
               "Na_mg_g", "Mg_mg_g", "K_mg_g", "Ca_mg_g")

soil_labels <- c(
  "pH"           = "pH",
  "OM%"          = "Organic matter (%)",
  "texture_>2mm" = "Coarse texture >2 mm (%)",
  "TN_%"         = "Total N (%)",
  "TP_%"         = "Total P (%)",
  "Na_mg_g"      = "Na (mg/g)",
  "Mg_mg_g"      = "Mg (mg/g)",
  "K_mg_g"       = "K (mg/g)",
  "Ca_mg_g"      = "Ca (mg/g)"
)

# Coerce to numeric (names with special chars can read as character from Excel)
plot_meta <- plot_meta %>%
  mutate(across(any_of(soil_vars), as.numeric))

# ── 1. Kruskal-Wallis: Moraine effect ────────────────────────────────────────
soil_kw <- bind_rows(lapply(soil_vars, function(sv) {
  x  <- plot_meta[[sv]]
  g  <- plot_meta$Moraine
  ok <- !is.na(x) & !is.na(g)
  kt <- tryCatch(kruskal.test(x[ok] ~ g[ok]), error = function(e) NULL)
  if (is.null(kt)) return(tibble(soil_var = sv, H = NA_real_, df = NA_real_, p_kw = NA_real_))
  tibble(soil_var = sv, H = kt$statistic, df = kt$parameter, p_kw = kt$p.value)
})) %>%
  mutate(Padj_kw = p.adjust(p_kw, method = "BH"))

# ── 2. Pairwise Wilcoxon post-hoc (Moraine) ──────────────────────────────────
soil_pw_moraine <- bind_rows(lapply(soil_vars, function(sv) {
  x  <- plot_meta[[sv]]
  g  <- plot_meta$Moraine
  ok <- !is.na(x) & !is.na(g)
  pw <- tryCatch(
    pairwise.wilcox.test(x[ok], g[ok], p.adjust.method = "none", exact = FALSE),
    error = function(e) NULL
  )
  if (is.null(pw)) return(tibble(soil_var = sv, contrast = NA_character_, p_pw = NA_real_))
  pm <- pw$p.value
  expand.grid(g1 = rownames(pm), g2 = colnames(pm), stringsAsFactors = FALSE) %>%
    filter(!is.na(pm[cbind(g1, g2)])) %>%
    mutate(soil_var = sv, contrast = paste(g1, g2, sep = "-"),
           p_pw = pm[cbind(g1, g2)]) %>%
    select(soil_var, contrast, p_pw)
}))

# CLD letters per soil variable
soil_cld_moraine <- bind_rows(lapply(soil_vars, function(sv) {
  pw <- soil_pw_moraine %>% filter(soil_var == sv, !is.na(p_pw))
  if (!nrow(pw)) return(tibble(soil_var = sv, Moraine = NA_character_, CLD = NA_character_))
  pvals <- setNames(pw$p_pw, pw$contrast)
  ml <- tryCatch(
    multcompView::multcompLetters(pvals, threshold = alpha),
    error = function(e) NULL
  )
  if (is.null(ml)) return(tibble(soil_var = sv, Moraine = NA_character_, CLD = NA_character_))
  tibble(soil_var = sv, Moraine = names(ml$Letters), CLD = trimws(unname(ml$Letters)))
}))

# ── 3. Wilcoxon: Location effect ─────────────────────────────────────────────
soil_loc <- bind_rows(lapply(soil_vars, function(sv) {
  x  <- plot_meta[[sv]]
  g  <- plot_meta$Location
  ok <- !is.na(x) & !is.na(g)
  wt <- tryCatch(wilcox.test(x[ok] ~ g[ok], exact = FALSE), error = function(e) NULL)
  if (is.null(wt)) return(tibble(soil_var = sv, W = NA_real_, p_loc = NA_real_))
  tibble(soil_var = sv, W = wt$statistic, p_loc = wt$p.value)
})) %>%
  mutate(Padj_loc = p.adjust(p_loc, method = "BH"))

# ── 4. Spearman: all env vars ─────────────────────────────────────────────────
soil_spearman_all <- bind_rows(lapply(soil_vars, function(sv) {
  bind_rows(lapply(env_vars, function(ev) {
    x  <- as.numeric(plot_meta[[ev]])
    y  <- as.numeric(plot_meta[[sv]])
    ok <- !is.na(x) & !is.na(y)
    if (sum(ok) < 5) return(tibble(soil_var = sv, env_var = ev, rho = NA_real_, p_sp = NA_real_))
    ct <- tryCatch(cor.test(x[ok], y[ok], method = "spearman", exact = FALSE),
                   error = function(e) NULL)
    if (is.null(ct)) return(tibble(soil_var = sv, env_var = ev, rho = NA_real_, p_sp = NA_real_))
    tibble(soil_var = sv, env_var = ev, rho = ct$estimate, p_sp = ct$p.value)
  }))
})) %>%
  group_by(soil_var) %>%
  mutate(Padj_sp = p.adjust(p_sp, method = "BH")) %>%
  ungroup()

# Convenience alias for the distance-from-glacier subset (used by the plot below)
soil_spearman <- soil_spearman_all %>% filter(env_var == "DistanceFromGlacier") %>%
  select(soil_var, rho, p_sp, Padj_sp)
```

### Moraine & location effects — summary table

```{r soil-summary-table}
# Group means per moraine
soil_long <- plot_meta %>%
  select(Plot, Location, Moraine, any_of(soil_vars)) %>%
  pivot_longer(cols = any_of(soil_vars), names_to = "soil_var", values_to = "value") %>%
  mutate(Moraine = factor(Moraine, levels = c("A", "B", "C")))

soil_moraine_means <- soil_long %>%
  group_by(soil_var, Moraine) %>%
  summarise(mean_val = round(mean(value, na.rm = TRUE), 3), .groups = "drop") %>%
  pivot_wider(names_from = Moraine, values_from = mean_val, names_prefix = "Mean_")

soil_cld_wide <- soil_cld_moraine %>%
  filter(!is.na(Moraine)) %>%
  pivot_wider(names_from = Moraine, values_from = CLD, names_prefix = "CLD_")

soil_summary <- soil_kw %>%
  left_join(soil_loc %>% select(soil_var, W, p_loc, Padj_loc), by = "soil_var") %>%
  left_join(soil_moraine_means,  by = "soil_var") %>%
  left_join(soil_cld_wide,       by = "soil_var") %>%
  mutate(
    soil_label = soil_labels[soil_var],
    `Young (A)`        = paste0(Mean_A, if_else(!is.na(CLD_A), paste0(" ", CLD_A), "")),
    `Intermediate (B)` = paste0(Mean_B, if_else(!is.na(CLD_B), paste0(" ", CLD_B), "")),
    `Old (C)`          = paste0(Mean_C, if_else(!is.na(CLD_C), paste0(" ", CLD_C), "")),
    KW_H     = round(H, 2),
    KW_Padj  = fmt_p(Padj_kw),
    KW_sig   = sig_stars(Padj_kw),
    Loc_W    = round(W, 1),
    Loc_Padj = fmt_p(Padj_loc),
    Loc_sig  = sig_stars(Padj_loc)
  ) %>%
  select(
    `Soil property` = soil_label,
    `Young (A)`, `Intermediate (B)`, `Old (C)`,
    `KW H` = KW_H, `KW Padj` = KW_Padj, ` ` = KW_sig,
    `W` = Loc_W, `Loc Padj` = Loc_Padj, `  ` = Loc_sig
  )

soil_summary %>%
  kable(
    caption = paste0(
      "Mean soil property by moraine age with compact-letter display ",
      "(CLD, unadjusted pairwise Wilcoxon, \u03b1 = 0.05). ",
      "KW = Kruskal-Wallis H (Moraine, BH-FDR). ",
      "W = Wilcoxon rank-sum (Nubra vs Tsomoriri, BH-FDR). ",
      "*** p<0.001, ** p<0.01, * p<0.05."
    ),
    escape = FALSE, align = "lrrrrrrrrrr"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
  add_header_above(c(" " = 1, "Mean value + CLD" = 3,
                     "Moraine (KW)" = 3, "Location (Wilcoxon)" = 3)) %>%
  column_spec(7, bold = TRUE) %>%
  column_spec(10, bold = TRUE)
```

### Environmental drivers — Spearman *ρ* table

```{r soil-env-table}
soil_env_wide <- soil_spearman_all %>%
  mutate(
    soil_label = soil_labels[soil_var],
    soil_label = factor(soil_label, levels = soil_labels),
    env_label  = env_labels[env_var],
    cell = case_when(
      is.na(rho) ~ "—",
      TRUE ~ paste0(
        formatC(rho, digits = 2, format = "f"),
        sig_stars(Padj_sp),
        " (", fmt_p(Padj_sp), ")"
      )
    )
  ) %>%
  select(soil_label, env_label, cell) %>%
  pivot_wider(names_from = env_label, values_from = cell)

soil_env_wide %>%
  rename(`Soil property` = soil_label) %>%
  kable(
    caption = paste0(
      "Spearman \u03c1 (BH-FDR adjusted p) — soil properties vs environmental variables. ",
      "Stars: *** p<0.001, ** p<0.01, * p<0.05."
    ),
    escape = TRUE, align = "lrrrrrrr"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

### Pairwise Wilcoxon — significant moraine contrasts

```{r soil-pairwise-detail, results='asis'}
sig_soil_kw <- soil_kw %>% filter(Padj_kw < alpha) %>% pull(soil_var)

for (sv in sig_soil_kw) {
  lbl <- soil_labels[[sv]]
  pw  <- soil_pw_moraine %>%
    filter(soil_var == sv, !is.na(p_pw)) %>%
    mutate(Padj = p.adjust(p_pw, method = "BH"),
           Sig  = sig_stars(Padj)) %>%
    select(Contrast = contrast, `p (unadj)` = p_pw, `Padj (BH)` = Padj, ` ` = Sig) %>%
    mutate(across(where(is.numeric), ~ round(.x, 4)))
  cat(
    pw %>%
      kable(format = "html",
            caption = paste0(lbl, " — pairwise Wilcoxon by moraine age"),
            escape = FALSE) %>%
      kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
      as.character(),
    "\n\n"
  )
}
```

```{r soil-distance-plots, fig.width = 11, fig.height = 16}
# Scale each soil variable 0-1 by the within-GlacierID maximum
soil_scaled <- plot_meta %>%
  select(Plot, GlacierID, Location, Moraine, Moraine_num, DistanceFromGlacier, any_of(soil_vars)) %>%
  group_by(GlacierID) %>%
  mutate(across(
    any_of(soil_vars),
    ~ {
      mx <- max(.x, na.rm = TRUE)
      if (is.na(mx) || mx == 0) NA_real_ else .x / mx
    }
  )) %>%
  ungroup() %>%
  pivot_longer(
    cols      = any_of(soil_vars),
    names_to  = "soil_var",
    values_to = "scaled_val"
  ) %>%
  mutate(
    soil_label = soil_labels[soil_var],
    soil_label = factor(soil_label, levels = soil_labels),
    Moraine    = factor(Moraine, levels = c("A", "B", "C")),
    Location   = factor(Location)
  )

# Spearman annotation per panel
soil_sp_annot <- soil_spearman %>%
  mutate(
    soil_label = soil_labels[soil_var],
    soil_label = factor(soil_label, levels = soil_labels),
    label = paste0(
      "\u03c1 = ", formatC(rho, digits = 2, format = "f"),
      sig_stars(Padj_sp),
      "\nPadj = ", fmt_p(Padj_sp)
    )
  )

p_soil_distance <- ggplot(soil_scaled, aes(x = DistanceFromGlacier, y = scaled_val)) +
  geom_smooth(
    method  = "lm", formula = y ~ x,
    colour  = "grey30", fill = "grey80", linewidth = 0.8, alpha = 0.3,
    na.rm   = TRUE
  ) +
  geom_point(
    aes(fill = Moraine, shape = Location),
    size = 3, colour = "white", stroke = 0.6, na.rm = TRUE
  ) +
  geom_text(
    data  = soil_sp_annot,
    aes(label = label),
    x     = -Inf, y = Inf,
    hjust = -0.08, vjust = 1.3,
    size  = 3, colour = "grey20", inherit.aes = FALSE
  ) +
  facet_wrap(~ soil_label, ncol = 2, scales = "free_x") +
  scale_fill_manual(
    values = moraine_colours, name = "Moraine age",
    labels = moraine_labels
  ) +
  scale_shape_manual(values = location_shapes, name = "Location") +
  scale_y_continuous(
    limits = c(0, 1.05), breaks = seq(0, 1, 0.25),
    labels = scales::label_percent(accuracy = 1)
  ) +
  guides(
    fill  = guide_legend(override.aes = list(shape = 21, size = 3.5)),
    shape = guide_legend(override.aes = list(fill = "grey60", size = 3.5))
  ) +
  labs(
    x = "Distance from glacier (m)",
    y = "Scaled value (proportion of GlacierID maximum)"
  ) +
  theme_bw(base_size = 11) +
  theme(
    strip.background = element_rect(fill = "grey92"),
    strip.text       = element_text(face = "bold", size = 10),
    legend.position  = "bottom",
    legend.box       = "horizontal",
    panel.grid.minor = element_blank()
  )

p_soil_distance
```

### Soil properties along the glacier retreat gradient — raw values

Same distance-gradient plot as above, but using raw soil chemistry values rather
than within-`GlacierID` scaled values.

```{r soil-distance-raw-plots, fig.width = 11, fig.height = 16}
soil_raw_distance <- plot_meta %>%
  select(Plot, Location, Moraine, DistanceFromGlacier, any_of(soil_vars)) %>%
  pivot_longer(
    cols      = any_of(soil_vars),
    names_to  = "soil_var",
    values_to = "soil_value"
  ) %>%
  mutate(
    soil_label = soil_labels[soil_var],
    soil_label = factor(soil_label, levels = soil_labels),
    Moraine    = factor(Moraine, levels = c("A", "B", "C")),
    Location   = factor(Location)
  )

p_soil_distance_raw <- ggplot(soil_raw_distance, aes(x = DistanceFromGlacier, y = soil_value)) +
  geom_smooth(
    method  = "lm", formula = y ~ x,
    colour  = "grey30", fill = "grey80", linewidth = 0.8, alpha = 0.3,
    na.rm   = TRUE
  ) +
  geom_point(
    aes(fill = Moraine, shape = Location),
    size = 3, colour = "white", stroke = 0.6, na.rm = TRUE
  ) +
  geom_text(
    data  = soil_sp_annot,
    aes(label = label),
    x     = -Inf, y = Inf,
    hjust = -0.08, vjust = 1.3,
    size  = 3, colour = "grey20", inherit.aes = FALSE
  ) +
  facet_wrap(~ soil_label, ncol = 2, scales = "free_y") +
  scale_fill_manual(
    values = moraine_colours, name = "Moraine age",
    labels = moraine_labels
  ) +
  scale_shape_manual(values = location_shapes, name = "Location") +
  guides(
    fill  = guide_legend(override.aes = list(shape = 21, size = 3.5)),
    shape = guide_legend(override.aes = list(fill = "grey60", size = 3.5))
  ) +
  labs(
    x = "Distance from glacier (m)",
    y = "Raw soil chemistry value"
  ) +
  theme_bw(base_size = 11) +
  theme(
    strip.background = element_rect(fill = "grey92"),
    strip.text       = element_text(face = "bold", size = 10),
    legend.position  = "bottom",
    legend.box       = "horizontal",
    panel.grid.minor = element_blank()
  )

p_soil_distance_raw
```

### Soil properties along ordinal moraine age

Same scaled soil-property values as above, but plotted against moraine age treated
as an ordinal numeric gradient (`Young/A = 1`, `Intermediate/B = 2`, `Old/C = 3`).

```{r soil-moraine-age-plots, fig.width = 11, fig.height = 16}
soil_moraine_sp <- soil_scaled %>%
  group_by(soil_var, soil_label) %>%
  group_modify(~ {
    ok <- complete.cases(.x$Moraine_num, .x$scaled_val)
    if (sum(ok) < 5) {
      return(tibble(rho = NA_real_, p_sp = NA_real_))
    }
    ct <- cor.test(
      .x$Moraine_num[ok],
      .x$scaled_val[ok],
      method = "spearman",
      exact = FALSE
    )
    tibble(rho = unname(ct$estimate), p_sp = ct$p.value)
  }) %>%
  ungroup() %>%
  mutate(
    Padj_sp = p.adjust(p_sp, method = "BH"),
    label = paste0(
      "\u03c1 = ", formatC(rho, digits = 2, format = "f"),
      sig_stars(Padj_sp),
      "\nPadj = ", fmt_p(Padj_sp)
    )
  )

p_soil_moraine_age <- ggplot(soil_scaled, aes(x = Moraine_num, y = scaled_val)) +
  geom_smooth(
    method  = "loess", formula = y ~ x, span = 1.5,
    colour  = "grey30", fill = "grey80", linewidth = 0.8, alpha = 0.3,
    na.rm   = TRUE
  ) +
  geom_point(
    aes(fill = Moraine, shape = Location),
    position = position_jitter(width = 0.05, height = 0),
    size = 3, colour = "white", stroke = 0.6, na.rm = TRUE
  ) +
  geom_text(
    data  = soil_moraine_sp,
    aes(label = label),
    x     = -Inf, y = Inf,
    hjust = -0.08, vjust = 1.3,
    size  = 3, colour = "grey20", inherit.aes = FALSE
  ) +
  facet_wrap(~ soil_label, ncol = 2) +
  scale_x_continuous(
    breaks = 1:3,
    labels = c("Young\n(1)", "Intermediate\n(2)", "Old\n(3)")
  ) +
  scale_fill_manual(
    values = moraine_colours, name = "Moraine age",
    labels = moraine_labels
  ) +
  scale_shape_manual(values = location_shapes, name = "Location") +
  scale_y_continuous(
    limits = c(0, 1.05), breaks = seq(0, 1, 0.25),
    labels = scales::label_percent(accuracy = 1)
  ) +
  guides(
    fill  = guide_legend(override.aes = list(shape = 21, size = 3.5)),
    shape = guide_legend(override.aes = list(fill = "grey60", size = 3.5))
  ) +
  labs(
    x = "Moraine age class (Young = 1, Intermediate = 2, Old = 3)",
    y = "Scaled value (proportion of GlacierID maximum)"
  ) +
  theme_bw(base_size = 11) +
  theme(
    strip.background = element_rect(fill = "grey92"),
    strip.text       = element_text(face = "bold", size = 10),
    legend.position  = "bottom",
    legend.box       = "horizontal",
    panel.grid.minor = element_blank()
  )

p_soil_moraine_age
```

### Export smoothed cover and soil figures

```{r export-cover-soil-smoothed-figures, message=FALSE, warning=FALSE}
cover_fig_dir <- file.path(project_root, "Cover_figures_REVISION")
dir.create(cover_fig_dir, recursive = TRUE, showWarnings = FALSE)

cover_soil_figures <- list(
  cover_distance = list(
    plot = p_cover_distance,
    width = 11,
    height = 14,
    description = "Scaled cover estimates vs distance from glacier"
  ),
  cover_distance_raw = list(
    plot = p_cover_distance_raw,
    width = 11,
    height = 14,
    description = "Raw cover estimates vs distance from glacier"
  ),
  cover_moraine_age = list(
    plot = p_cover_moraine_age,
    width = 11,
    height = 14,
    description = "Cover estimates vs ordinal moraine age"
  ),
  soil_distance = list(
    plot = p_soil_distance,
    width = 11,
    height = 16,
    description = "Scaled soil chemistry vs distance from glacier"
  ),
  soil_distance_raw = list(
    plot = p_soil_distance_raw,
    width = 11,
    height = 16,
    description = "Raw soil chemistry vs distance from glacier"
  ),
  soil_moraine_age = list(
    plot = p_soil_moraine_age,
    width = 11,
    height = 16,
    description = "Soil chemistry vs ordinal moraine age"
  )
)

export_cover_soil_figure <- function(name, fig) {
  png_file <- file.path(cover_fig_dir, paste0(name, "_REVISION.png"))
  svg_file <- file.path(cover_fig_dir, paste0(name, "_REVISION.svg"))

  ggsave(
    filename = png_file,
    plot = fig$plot,
    width = fig$width,
    height = fig$height,
    dpi = 300,
    bg = "white"
  )

  save_gg_svg(
    filename = svg_file,
    plot = fig$plot,
    width = fig$width,
    height = fig$height
  )

  tibble(
    figure = name,
    description = fig$description,
    png = png_file,
    svg = svg_file
  )
}

cover_soil_export_index <- bind_rows(
  lapply(names(cover_soil_figures), function(nm) {
    export_cover_soil_figure(nm, cover_soil_figures[[nm]])
  })
)

write.csv(
  cover_soil_export_index,
  file.path(cover_fig_dir, "Cover_soil_smoothed_figures_index_REVISION.csv"),
  row.names = FALSE
)

cover_soil_export_index %>%
  kable(caption = "Exported smoothed cover and soil figures") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## Cover × soil relationships

Spearman *ρ* between every cover type and every soil property. BH-FDR correction applied
across all 63 pairs (7 cover × 9 soil). A heatmap shows the full correlation matrix;
scatter plots are produced for all pairs with Padj < 0.05.

```{r cover-soil-correlations}
# ── Full cover × soil Spearman matrix ────────────────────────────────────────
cover_soil_cor <- bind_rows(lapply(cover_vars, function(cv) {
  bind_rows(lapply(soil_vars, function(sv) {
    x  <- as.numeric(plot_meta[[cv]])
    y  <- as.numeric(plot_meta[[sv]])
    ok <- !is.na(x) & !is.na(y)
    if (sum(ok) < 5) {
      return(tibble(cover_var = cv, soil_var = sv, rho = NA_real_, p_sp = NA_real_))
    }
    ct <- tryCatch(
      cor.test(x[ok], y[ok], method = "spearman", exact = FALSE),
      error = function(e) NULL
    )
    if (is.null(ct)) {
      return(tibble(cover_var = cv, soil_var = sv, rho = NA_real_, p_sp = NA_real_))
    }
    tibble(cover_var = cv, soil_var = sv, rho = ct$estimate, p_sp = ct$p.value)
  }))
})) %>%
  mutate(Padj = p.adjust(p_sp, method = "BH"))

# Significant pairs
sig_pairs <- cover_soil_cor %>% filter(!is.na(Padj), Padj < alpha)
cat(nrow(sig_pairs), "significant cover × soil pairs (Padj <", alpha, ")\n")

# Numeric summary table
cover_soil_cor %>%
  mutate(
    cover_label = cover_labels[cover_var],
    soil_label  = soil_labels[soil_var],
    cell = paste0(
      formatC(rho, digits = 2, format = "f"),
      sig_stars(Padj),
      " (", fmt_p(Padj), ")"
    )
  ) %>%
  select(cover_label, soil_label, cell) %>%
  pivot_wider(names_from = soil_label, values_from = cell) %>%
  rename(`Cover type` = cover_label) %>%
  kable(
    caption = paste0(
      "Spearman \u03c1 (BH-FDR p) — cover types vs soil properties ",
      "(FDR across all 63 pairs). Stars: *** p<0.001, ** p<0.01, * p<0.05."
    ),
    escape = TRUE, align = "lrrrrrrrrr"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

```{r cover-soil-heatmap, fig.width = 9, fig.height = 5}
# ── Heatmap of Spearman ρ ─────────────────────────────────────────────────────
hm_df <- cover_soil_cor %>%
  mutate(
    cover_label = factor(cover_labels[cover_var], levels = rev(cover_labels)),
    soil_label  = factor(soil_labels[soil_var],   levels = soil_labels),
    sig_label   = case_when(
      is.na(Padj)   ~ "",
      Padj < 0.001  ~ "***",
      Padj < 0.01   ~ "**",
      Padj < 0.05   ~ "*",
      TRUE          ~ ""
    )
  )

ggplot(hm_df, aes(x = soil_label, y = cover_label, fill = rho)) +
  geom_tile(colour = "white", linewidth = 0.5) +
  geom_text(aes(label = sig_label), size = 4, vjust = 0.8, colour = "grey10") +
  scale_fill_gradient2(
    low      = "#d73027",
    mid      = "white",
    high     = "#1a9850",
    midpoint = 0,
    limits   = c(-1, 1),
    name     = "Spearman \u03c1"
  ) +
  scale_x_discrete(position = "top") +
  labs(
    x       = NULL, y = NULL,
    caption = "Stars indicate BH-FDR significance: *** p<0.001, ** p<0.01, * p<0.05"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x      = element_text(angle = 35, hjust = 0, size = 10),
    axis.text.y      = element_text(size = 10),
    legend.position  = "right",
    panel.grid       = element_blank()
  )
```

```{r cover-soil-scatter, fig.width = 11, fig.height = 3.5 * ceiling(nrow(sig_pairs) / 3), warning = FALSE}
# ── Scatter plots for significant pairs only ──────────────────────────────────
if (nrow(sig_pairs) == 0) {
  cat("No significant cover × soil pairs at Padj < 0.05.\n")
} else {
  scatter_plots <- lapply(seq_len(nrow(sig_pairs)), function(i) {
    cv  <- sig_pairs$cover_var[[i]]
    sv  <- sig_pairs$soil_var[[i]]
    rho <- sig_pairs$rho[[i]]
    pad <- sig_pairs$Padj[[i]]
    lbl <- paste0(
      "\u03c1 = ", formatC(rho, digits = 2, format = "f"),
      sig_stars(pad),
      "\nPadj = ", fmt_p(pad)
    )

    d <- plot_meta %>%
      select(Location, Moraine,
             x = all_of(sv), y = all_of(cv)) %>%
      filter(!is.na(x), !is.na(y)) %>%
      mutate(Moraine = factor(Moraine, levels = c("A", "B", "C")))

    ggplot(d, aes(x = x, y = y)) +
      geom_smooth(
        method = "lm", formula = y ~ x, colour = "grey40",
        fill = "grey80", linewidth = 0.7, alpha = 0.25, na.rm = TRUE
      ) +
      geom_point(
        aes(fill = Moraine, shape = Location),
        size = 2.8, colour = "white", stroke = 0.5
      ) +
      annotate(
        "text", x = -Inf, y = Inf, label = lbl,
        hjust = -0.08, vjust = 1.4, size = 2.9, colour = "grey20"
      ) +
      scale_fill_manual(values = moraine_colours, guide = "none") +
      scale_shape_manual(values = location_shapes, guide = "none") +
      labs(
        x = soil_labels[[sv]],
        y = cover_labels[[cv]]
      ) +
      theme_bw(base_size = 10) +
      theme(panel.grid.minor = element_blank())
  })

  # Shared legend (extract from one plot with legends shown)
  legend_plot <- ggplot(
    plot_meta %>%
      mutate(Moraine = factor(Moraine, levels = c("A", "B", "C"))),
    aes(x = 1, y = 1, fill = Moraine, shape = Location)
  ) +
    geom_point(size = 3, colour = "white", stroke = 0.5) +
    scale_fill_manual(
      values = moraine_colours, name = "Moraine age",
      labels = moraine_labels
    ) +
    scale_shape_manual(values = location_shapes, name = "Location") +
    guides(
      fill  = guide_legend(override.aes = list(shape = 21, size = 3)),
      shape = guide_legend(override.aes = list(fill = "grey60", size = 3))
    ) +
    theme_void()

  wrap_plots(scatter_plots, ncol = 3) +
    plot_layout(guides = "collect") &
    theme(legend.position = "bottom")
}
```

## Spearman correlation heatmaps — cover, soil, and environment

Three cross-block Spearman heatmaps:

1. **Cover × soil**
2. **Cover × topography / environmental** (distance, moraine age, elevation, slope, TPI, aspect, snow)
3. **Soil × topography / environmental**

Benjamini–Hochberg FDR is applied **within each heatmap** (across that block's unique pairs).

```{r cross-block-spearman-setup}
env_corr_vars <- c(
  "DistanceFromGlacier",
  "Moraine_num",
  setdiff(env_vars, "DistanceFromGlacier")
)

env_corr_labels <- c(
  DistanceFromGlacier = env_labels[["DistanceFromGlacier"]],
  Moraine_num         = "Moraine age (Young=1, Old=3)",
  env_labels[setdiff(env_vars, "DistanceFromGlacier")]
)

compute_cross_spearman <- function(df, vars_row, vars_col) {
  bind_rows(lapply(vars_row, function(v_row) {
    bind_rows(lapply(vars_col, function(v_col) {
      x <- as.numeric(df[[v_row]])
      y <- as.numeric(df[[v_col]])
      ok <- complete.cases(x, y)
      if (sum(ok) < 5L) {
        return(tibble(
          var_row = v_row,
          var_col = v_col,
          rho = NA_real_,
          p_sp = NA_real_
        ))
      }
      ct <- tryCatch(
        cor.test(x[ok], y[ok], method = "spearman", exact = FALSE),
        error = function(e) NULL
      )
      if (is.null(ct)) {
        return(tibble(
          var_row = v_row,
          var_col = v_col,
          rho = NA_real_,
          p_sp = NA_real_
        ))
      }
      tibble(
        var_row = v_row,
        var_col = v_col,
        rho = unname(ct$estimate),
        p_sp = ct$p.value
      )
    }))
  })) %>%
    mutate(Padj = p.adjust(p_sp, method = "BH"))
}

annotate_cross_cor <- function(cor_df, vars_row, vars_col, row_labels, col_labels) {
  cor_df %>%
    mutate(
      row_label = factor(
        unname(row_labels[var_row]),
        levels = rev(unname(row_labels[vars_row]))
      ),
      col_label = factor(
        unname(col_labels[var_col]),
        levels = unname(col_labels[vars_col])
      ),
      sig_label = case_when(
        is.na(Padj)  ~ "",
        Padj < 0.001 ~ "***",
        Padj < 0.01  ~ "**",
        Padj < 0.05  ~ "*",
        TRUE         ~ ""
      )
    )
}

make_cross_spearman_heatmap <- function(
    hm_df,
    title,
    subtitle,
    x_lab = NULL,
    y_lab = NULL
) {
  ggplot(hm_df, aes(x = col_label, y = row_label, fill = rho)) +
    geom_tile(colour = "white", linewidth = 0.5) +
    geom_text(aes(label = sig_label), size = 3.5, vjust = 0.8, colour = "grey10") +
    scale_fill_gradient2(
      low = "#d73027",
      mid = "white",
      high = "#1a9850",
      midpoint = 0,
      limits = c(-1, 1),
      name = "Spearman \u03c1",
      na.value = "grey95"
    ) +
    scale_x_discrete(position = "top") +
    labs(
      title = title,
      subtitle = subtitle,
      x = x_lab,
      y = y_lab,
      caption = "Stars: *** Padj < 0.001, ** Padj < 0.01, * Padj < 0.05 (BH-FDR within heatmap)."
    ) +
    theme_minimal(base_size = 11) +
    theme(
      axis.text.x = element_text(angle = 35, hjust = 0, size = 9),
      axis.text.y = element_text(size = 9),
      panel.grid = element_blank(),
      legend.position = "right"
    )
}

cover_soil_cor_block <- compute_cross_spearman(plot_meta, cover_vars, soil_vars) %>%
  annotate_cross_cor(cover_vars, soil_vars, cover_labels, soil_labels)

cover_env_cor_block <- compute_cross_spearman(plot_meta, cover_vars, env_corr_vars) %>%
  annotate_cross_cor(cover_vars, env_corr_vars, cover_labels, env_corr_labels)

soil_env_cor_block <- compute_cross_spearman(plot_meta, soil_vars, env_corr_vars) %>%
  annotate_cross_cor(soil_vars, env_corr_vars, soil_labels, env_corr_labels)

cross_block_cor_all <- bind_rows(
  cover_soil_cor_block %>% mutate(block = "Cover x Soil"),
  cover_env_cor_block %>% mutate(block = "Cover x Environment"),
  soil_env_cor_block %>% mutate(block = "Soil x Environment")
) %>%
  mutate(
    row_label = case_when(
      block == "Cover x Soil" ~ cover_labels[var_row],
      block == "Cover x Environment" ~ cover_labels[var_row],
      block == "Soil x Environment" ~ soil_labels[var_row]
    ),
    col_label = case_when(
      block == "Cover x Soil" ~ soil_labels[var_col],
      block == "Cover x Environment" ~ env_corr_labels[var_col],
      block == "Soil x Environment" ~ env_corr_labels[var_col]
    ),
    sig_label = case_when(
      is.na(Padj)  ~ "",
      Padj < 0.001 ~ "***",
      Padj < 0.01  ~ "**",
      Padj < 0.05  ~ "*",
      TRUE         ~ ""
    )
  )

cat("Cover x soil pairs:", nrow(cover_soil_cor_block),
    "| sig:", sum(cover_soil_cor_block$Padj < alpha, na.rm = TRUE), "\n")
cat("Cover x environment pairs:", nrow(cover_env_cor_block),
    "| sig:", sum(cover_env_cor_block$Padj < alpha, na.rm = TRUE), "\n")
cat("Soil x environment pairs:", nrow(soil_env_cor_block),
    "| sig:", sum(soil_env_cor_block$Padj < alpha, na.rm = TRUE), "\n")
```

### 1. Cover × soil

```{r cover-soil-cross-heatmap, fig.width=10, fig.height=6}
p_cover_soil_cross_heatmap <- make_cross_spearman_heatmap(
  cover_soil_cor_block,
  title = "Cover × soil — Spearman correlations",
  subtitle = paste0(
    "BH-FDR across ", nrow(cover_soil_cor_block), " cover × soil pairs."
  ),
  x_lab = "Soil property",
  y_lab = "Cover type"
)

p_cover_soil_cross_heatmap
```

### 2. Cover × topography / environmental

```{r cover-env-cross-heatmap, fig.width=10, fig.height=6}
p_cover_env_cross_heatmap <- make_cross_spearman_heatmap(
  cover_env_cor_block,
  title = "Cover × topography / environmental — Spearman correlations",
  subtitle = paste0(
    "BH-FDR across ", nrow(cover_env_cor_block),
    " cover × environmental pairs. Moraine age = ordinal (A=1, B=2, C=3)."
  ),
  x_lab = "Topography / environmental variable",
  y_lab = "Cover type"
)

p_cover_env_cross_heatmap
```

### 3. Soil × topography / environmental

```{r soil-env-cross-heatmap, fig.width=10, fig.height=6.5}
p_soil_env_cross_heatmap <- make_cross_spearman_heatmap(
  soil_env_cor_block,
  title = "Soil × topography / environmental — Spearman correlations",
  subtitle = paste0(
    "BH-FDR across ", nrow(soil_env_cor_block),
    " soil × environmental pairs. Moraine age = ordinal (A=1, B=2, C=3)."
  ),
  x_lab = "Topography / environmental variable",
  y_lab = "Soil property"
)

p_soil_env_cross_heatmap
```

```{r export-cross-block-spearman-heatmaps, message=FALSE, warning=FALSE}
cover_fig_dir <- file.path(project_root, "Cover_figures_REVISION")
dir.create(cover_fig_dir, recursive = TRUE, showWarnings = FALSE)

cross_block_corr_csv <- file.path(
  cover_fig_dir,
  "Cross_block_spearman_correlations_REVISION.csv"
)
write.csv(cross_block_cor_all, cross_block_corr_csv, row.names = FALSE)

export_cross_heatmap <- function(plot_obj, slug, width, height) {
  png_file <- file.path(cover_fig_dir, paste0(slug, "_REVISION.png"))
  svg_file <- file.path(cover_fig_dir, paste0(slug, "_REVISION.svg"))
  ggsave(png_file, plot = plot_obj, width = width, height = height, dpi = 300, bg = "white")
  save_gg_svg(svg_file, plot = plot_obj, width = width, height = height)
  tibble(figure = slug, png = png_file, svg = svg_file)
}

cross_block_export_index <- bind_rows(
  export_cross_heatmap(p_cover_soil_cross_heatmap, "Spearman_heatmap_cover_soil", 10, 6),
  export_cross_heatmap(p_cover_env_cross_heatmap, "Spearman_heatmap_cover_environment", 10, 6),
  export_cross_heatmap(p_soil_env_cross_heatmap, "Spearman_heatmap_soil_environment", 10, 6.5)
) %>%
  mutate(csv = cross_block_corr_csv)

cross_block_export_index %>%
  kable(caption = "Exported cross-block Spearman heatmaps") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## Structural equation model — environment → colonization → soil

A path model testing the directional chain:
**environmental gradient → vegetation colonization → soil development**.

With ~20–30 plots, individual-variable SEM is parameter-saturated. Instead:

1. **PCA on cover variables** → PC1 scores as a *vegetation colonisation index*
   (loadings show which cover types load positively).
2. **PCA on soil variables** → PC1 scores as a *soil development index*.
3. **`lavaan` path model** (all variables standardised):

```
Distance, Elevation, Snow → VegPC1 → SoilPC1
                          ↘ direct ↗
```

Standardised path coefficients, *p*-values, and bootstrapped indirect effects are
reported. Fisher's C test (from a `piecewiseSEM` cross-check) assesses model fit.

```{r sem-setup, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(lavaan)
  library(piecewiseSEM)
})

# ── Clean variable names for formula use ─────────────────────────────────────
sem_raw <- plot_meta %>%
  rename(
    OM_pct       = `OM%`,
    TN_pct       = `TN_%`,
    TP_pct       = `TP_%`,
    Texture_2mm  = `texture_>2mm`
  ) %>%
  select(
    Plot, Location, Moraine,
    DistanceFromGlacier, Elevation, Slope, TPI, Aspect_sin, Aspect_cos, Snow_1976,
    all_of(cover_vars),
    OM_pct, pH, TN_pct, TP_pct, Na_mg_g, Mg_mg_g, K_mg_g, Ca_mg_g, Texture_2mm
  ) %>%
  filter(complete.cases(.))

cat("n plots (complete cases):", nrow(sem_raw), "\n")

# ── PCA — cover (vegetation colonisation index) ───────────────────────────────
cover_mat <- sem_raw %>%
  select(all_of(cover_vars)) %>%
  mutate(across(everything(), as.numeric))

pca_cover <- prcomp(cover_mat, center = TRUE, scale. = TRUE)

cat("\nCover PCA — variance explained:\n")
print(round(summary(pca_cover)$importance[, 1:4], 3))
cat("\nCover PC1 loadings:\n")
print(round(pca_cover$rotation[, 1], 3))

# ── PCA — soil (soil development index) ──────────────────────────────────────
soil_mat <- sem_raw %>%
  select(OM_pct, pH, TN_pct, TP_pct, Na_mg_g, Mg_mg_g, K_mg_g, Ca_mg_g, Texture_2mm) %>%
  mutate(across(everything(), as.numeric))

pca_soil <- prcomp(soil_mat, center = TRUE, scale. = TRUE)

cat("\nSoil PCA — variance explained:\n")
print(round(summary(pca_soil)$importance[, 1:4], 3))
cat("\nSoil PC1 loadings:\n")
print(round(pca_soil$rotation[, 1], 3))

# ── PCA — topography (all env except Distance) ────────────────────────────────
# Snow_1976 is binary (0/1) but included as numeric for PCA; interpretable as
# "recently glaciated" axis combined with terrain shape variables.
topo_vars_pca <- c("Elevation", "Slope", "TPI", "Aspect_sin", "Aspect_cos", "Snow_1976")

topo_mat <- sem_raw %>%
  select(all_of(topo_vars_pca)) %>%
  mutate(across(everything(), as.numeric))

pca_topo <- prcomp(topo_mat, center = TRUE, scale. = TRUE)

cat("\nTopography PCA — variance explained:\n")
print(round(summary(pca_topo)$importance[, 1:4], 3))
cat("\nTopography PC1 loadings:\n")
print(round(pca_topo$rotation[, 1], 3))

# ── Microbial genus richness — rarefied, plot-level ───────────────────────────
suppressPackageStartupMessages({
  library(phyloseq)
  library(vegan)
  library(NetCoMi)
})

# ── Helper: resolve RDS path ──────────────────────────────────────────────────
pick_existing_file <- function(candidates, label = "file") {
  hits <- candidates[file.exists(candidates)]
  if (length(hits) == 0)
    stop("Could not find ", label, ". Tried:\n",
         paste0("  - ", candidates, collapse = "\n"))
  normalizePath(hits[[1]], winslash = "/", mustWork = FALSE)
}

# ── Load and glom to genus ────────────────────────────────────────────────────
load_genus_ps <- function(rds_candidates, label) {
  ps_raw <- readRDS(pick_existing_file(rds_candidates, label))
  ps_g   <- phyloseq::tax_glom(ps_raw, taxrank = "Genus")
  ps_g   <- NetCoMi::renameTaxa(
    ps_g,
    pat      = "<name>",
    substPat = "<name>_<subst_name>(<subst_R>)",
    numDupli = "Genus"
  )
  ps_g
}

ps_16s_genus <- load_genus_ps(
  c(file.path(molecular_dir, "ITS", "La_Gla_final_16S_core.Rds"),
    file.path(molecular_dir, "La_Gla_final_16S_core.Rds")),
  "16S core phyloseq"
)

ps_its_genus <- load_genus_ps(
  c(file.path(molecular_dir, "ITS", "La_Gla_final_ITS_core.Rds"),
    file.path(molecular_dir, "La_Gla_final_ITS_core.Rds")),
  "ITS core phyloseq"
)

# ── Pool samples to plot level, then rarefy iteratively ───────────────────────
# n_iter rarefaction draws are averaged to remove stochastic noise from a
# single subsample draw.  Both observed genus richness and Shannon H' are
# averaged across iterations.
pool_and_rarefy <- function(ps, plot_col = "Plot", n_iter = 999, seed = 42) {
  otu  <- as(phyloseq::otu_table(ps), "matrix")
  if (phyloseq::taxa_are_rows(ps)) otu <- t(otu)   # rows = samples
  meta <- as(phyloseq::sample_data(ps), "data.frame")

  stopifnot(plot_col %in% colnames(meta))

  plots     <- as.character(meta[[plot_col]])
  pool      <- rowsum(otu, group = plots)     # sum counts per plot
  depth     <- rowSums(pool)
  min_depth <- min(depth)
  cat("  Pooled reads — min:", min_depth,
      " | median:", round(median(depth)),
      " | n plots:", nrow(pool), "\n")

  pool_ok <- pool[depth >= min_depth, , drop = FALSE]

  set.seed(seed)
  rich_mat <- matrix(NA_real_, nrow = nrow(pool_ok), ncol = n_iter)
  shan_mat <- matrix(NA_real_, nrow = nrow(pool_ok), ncol = n_iter)

  for (i in seq_len(n_iter)) {
    rare_i          <- vegan::rrarefy(pool_ok, sample = min_depth)
    rich_mat[, i]   <- rowSums(rare_i > 0)
    shan_mat[, i]   <- vegan::diversity(rare_i, index = "shannon")
  }

  tibble(
    Plot     = rownames(pool_ok),
    n_genera = rowMeans(rich_mat),        # mean observed genus richness
    shannon  = rowMeans(shan_mat)         # mean Shannon H'
  )
}

cat("Pooling 16S to plot level — iterative rarefaction (999 draws) ...\n")
div_16s <- pool_and_rarefy(ps_16s_genus) %>%
  rename(Rich_16S = n_genera, Shan_16S = shannon)

cat("Pooling ITS to plot level — iterative rarefaction (999 draws) ...\n")
div_its <- pool_and_rarefy(ps_its_genus) %>%
  rename(Rich_ITS = n_genera, Shan_ITS = shannon)

rich_raw <- full_join(div_16s, div_its, by = "Plot")

cat("\nDiversity summary (999-draw average):\n")
cat("16S — median richness:", round(median(rich_raw$Rich_16S, na.rm = TRUE), 1),
    " | median Shannon:", round(median(rich_raw$Shan_16S, na.rm = TRUE), 3), "\n")
cat("ITS — median richness:", round(median(rich_raw$Rich_ITS, na.rm = TRUE), 1),
    " | median Shannon:", round(median(rich_raw$Shan_ITS, na.rm = TRUE), 3), "\n")
print(rich_raw)
```

```{r sem-pca-biplots, fig.width=13, fig.height=4.5}
# ── Generic biplot helper ─────────────────────────────────────────────────────
biplot_df <- function(pca_obj, score_df) {
  scores   <- as.data.frame(pca_obj$x[, 1:2])
  loadings <- as.data.frame(pca_obj$rotation[, 1:2])
  loadings$Variable <- rownames(loadings)
  list(scores = bind_cols(score_df, scores), loadings = loadings)
}

make_biplot <- function(bp, pca_obj, title, var_labels = NULL, scale = 2.5) {
  pct <- round(summary(pca_obj)$importance[2, 1:2] * 100, 1)
  labs_fn <- function(v) if (!is.null(var_labels) && v %in% names(var_labels)) var_labels[v] else v

  ggplot(bp$scores, aes(x = PC1, y = PC2)) +
    geom_hline(yintercept = 0, colour = "grey70", linetype = "dashed") +
    geom_vline(xintercept = 0, colour = "grey70", linetype = "dashed") +
    geom_point(aes(fill = Moraine, shape = Location),
               size = 3, colour = "white", stroke = 0.5) +
    geom_segment(data = bp$loadings,
                 aes(x = 0, y = 0, xend = PC1 * scale, yend = PC2 * scale),
                 arrow = arrow(length = unit(0.18, "cm")),
                 colour = "#c0392b", linewidth = 0.7) +
    geom_text(data = bp$loadings %>%
                mutate(lbl = sapply(Variable, labs_fn)),
              aes(x = PC1 * (scale + 0.3), y = PC2 * (scale + 0.3), label = lbl),
              size = 2.8, colour = "#c0392b", fontface = "italic") +
    scale_fill_manual(
      values = moraine_colours,
      labels = moraine_labels,
      name = "Moraine age"
    ) +
    scale_shape_manual(values = location_shapes, name = "Location") +
    guides(
      fill = guide_legend(override.aes = list(shape = 21)),
      shape = guide_legend(override.aes = list(fill = "grey60"))
    ) +
    labs(title = paste0(title, " PCA"),
         x = paste0("PC1 (", pct[1], "%)"),
         y = paste0("PC2 (", pct[2], "%)")) +
    theme_bw(base_size = 10) +
    theme(legend.position = "right", panel.grid.minor = element_blank())
}

cover_bp <- biplot_df(pca_cover, sem_raw %>% select(Location, Moraine))
soil_bp  <- biplot_df(pca_soil,  sem_raw %>% select(Location, Moraine))
topo_bp  <- biplot_df(pca_topo,  sem_raw %>% select(Location, Moraine))

cover_var_labels <- setNames(
  c("Lichen","Moss","Crust","Forb","Graminoid","Bare soil","Stones"), cover_vars
)
soil_var_labels <- c(
  OM_pct = "OM%", pH = "pH", TN_pct = "TN%", TP_pct = "TP%",
  Na_mg_g = "Na", Mg_mg_g = "Mg", K_mg_g = "K", Ca_mg_g = "Ca",
  Texture_2mm = ">2mm"
)
topo_var_labels <- c(
  Elevation  = "Elevation", Slope   = "Slope", TPI       = "TPI",
  Aspect_sin = "Asp.(sin)", Aspect_cos = "Asp.(cos)", Snow_1976 = "Snow 1976"
)

make_biplot(cover_bp, pca_cover, "Cover",       cover_var_labels) +
make_biplot(soil_bp,  pca_soil,  "Soil",        soil_var_labels) +
make_biplot(topo_bp,  pca_topo,  "Topography",  topo_var_labels) +
plot_layout(guides = "collect", ncol = 3) &
  theme(legend.position = "bottom")
```

### PCA conditioned by location

The same three PCAs repeated separately for **Nubra** and **Tsomoriri** to check
whether the dominant axes of variation are consistent across locations or driven
by location-specific gradients.

```{r sem-pca-biplots-location, fig.width=13, fig.height=9, warning=FALSE}
# ── Run all three PCAs within each location ───────────────────────────────────
locations <- unique(sem_raw$Location)

# For each location, compute PCAs and collect biplots
pca_vars <- list(
  Cover      = list(vars = cover_vars,    labels = cover_var_labels,  scale = 2.5),
  Soil       = list(vars = c("OM_pct","pH","TN_pct","TP_pct",
                              "Na_mg_g","Mg_mg_g","K_mg_g","Ca_mg_g",
                              "Texture_2mm"),
                    labels = soil_var_labels, scale = 2.5),
  Topography = list(vars = topo_vars_pca, labels = topo_var_labels,   scale = 2.0)
)

loc_plots <- lapply(locations, function(loc) {
  sub <- sem_raw %>% filter(Location == loc)

  lapply(names(pca_vars), function(ptype) {
    cfg     <- pca_vars[[ptype]]
    mat     <- sub %>%
      select(all_of(cfg$vars)) %>%
      mutate(across(everything(), as.numeric)) %>%
      # drop columns with zero variance (can't scale them)
      select(where(~ var(., na.rm = TRUE) > 0))

    if (ncol(mat) < 2 || nrow(mat) < 3) return(NULL)

    pca_loc <- prcomp(mat, center = TRUE, scale. = TRUE)
    bp_loc  <- biplot_df(pca_loc, sub %>% select(Location, Moraine))

    make_biplot(bp_loc, pca_loc,
                title      = paste0(ptype, "\n(", loc, ")"),
                var_labels = cfg$labels,
                scale      = cfg$scale) +
      theme(plot.title = element_text(size = 9, face = "bold"))
  })
})

# Flatten into a list, removing NULLs
plot_list <- Filter(Negate(is.null), unlist(loc_plots, recursive = FALSE))

wrap_plots(plot_list, ncol = 3) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

### Partial PCA — variation independent of location

The between-location centroid (Nubra vs Tsomoriri mean) is removed as a
conditioning variable before decomposition: `rda(mat ~ Condition(Location))`.
The resulting axes (`PC1`, `PC2`) capture only the *within-location* covariance
structure — i.e. variation shared by both glacier systems after the location
effect is partialled out. Points are coloured by moraine age; shapes still
distinguish locations so any residual separation can be seen.

```{r sem-pca-partial, fig.width=13, fig.height=4.5, warning=FALSE, message=FALSE}
suppressPackageStartupMessages(library(vegan))

# ── Partial RDA variance summary helper ───────────────────────────────────────
summarise_partial_rda_variance <- function(mat, meta, input_name) {
  ok_cols <- apply(mat, 2, var, na.rm = TRUE) > 0
  mat <- mat[, ok_cols, drop = FALSE]

  rda_obj <- vegan::rda(
    mat ~ Condition(Location),
    data = meta,
    scale = TRUE
  )

  total_inertia <- rda_obj$tot.chi
  conditioned_inertia <- if (is.null(rda_obj$pCCA)) {
    0
  } else {
    rda_obj$pCCA$tot.chi
  }
  residual_inertia <- rda_obj$CA$tot.chi

  residual_eigenvalues <- vegan::eigenvals(
    rda_obj,
    model = "unconstrained"
  )
  residual_axis_pct <- residual_eigenvalues /
    sum(residual_eigenvalues) * 100

  tibble(
    PCA_input = input_name,
    n_variables = ncol(mat),
    total_inertia = total_inertia,
    conditioned_inertia = conditioned_inertia,
    variation_conditioned_by_location_pct =
      conditioned_inertia / total_inertia * 100,
    residual_inertia = residual_inertia,
    residual_variation_pct = residual_inertia / total_inertia * 100,
    PC1_pct_of_residual_variation = residual_axis_pct[1],
    PC2_pct_of_residual_variation = residual_axis_pct[2]
  )
}

# ── Partial RDA biplot helper ─────────────────────────────────────────────────
make_partial_biplot <- function(mat, meta, title, var_labels = NULL,
                                scale = 2.0) {

  # Drop zero-variance columns
  ok_cols <- apply(mat, 2, var, na.rm = TRUE) > 0
  mat     <- mat[, ok_cols, drop = FALSE]

  # Partial RDA: condition on Location
  rda_obj <- vegan::rda(mat ~ Condition(Location), data = meta,
                         scale = TRUE)

  # Variance explained by each residual PC
  ev      <- eigenvals(rda_obj, model = "unconstrained")
  pct_ev  <- round(ev / sum(ev) * 100, 1)

  # Site scores (one row per plot)
  sites   <- as.data.frame(scores(rda_obj, display = "sites",
                                   choices = 1:2, scaling = 2))
  sites$Location <- meta$Location
  sites$Moraine  <- meta$Moraine

  # Species loadings
  sp      <- as.data.frame(scores(rda_obj, display = "species",
                                   choices = 1:2, scaling = 2))
  sp$Variable <- rownames(sp)

  labs_fn <- function(v)
    if (!is.null(var_labels) && v %in% names(var_labels)) var_labels[[v]] else v

  ggplot(sites, aes(x = PC1, y = PC2)) +
    geom_hline(yintercept = 0, colour = "grey70", linetype = "dashed") +
    geom_vline(xintercept = 0, colour = "grey70", linetype = "dashed") +
    geom_point(aes(fill = Moraine, shape = Location),
               size = 3.5, colour = "white", stroke = 0.5) +
    geom_segment(data = sp,
                 aes(x = 0, y = 0, xend = PC1 * scale, yend = PC2 * scale),
                 arrow = arrow(length = unit(0.18, "cm")),
                 colour = "#c0392b", linewidth = 0.7) +
    geom_text(data = sp %>%
                mutate(lbl = sapply(Variable, labs_fn)),
              aes(x = PC1 * (scale + 0.3), y = PC2 * (scale + 0.3),
                  label = lbl),
              size = 2.8, colour = "#c0392b", fontface = "italic") +
    scale_fill_manual(
      values = moraine_colours,
      labels = moraine_labels,
      name   = "Moraine age"
    ) +
    scale_shape_manual(values = location_shapes, name = "Location") +
    guides(
      fill = guide_legend(override.aes = list(shape = 21)),
      shape = guide_legend(override.aes = list(fill = "grey60"))
    ) +
    labs(title = paste0(title, "\n(partial PCA | Location conditioned out)"),
         x = paste0("PC1 (", pct_ev[1], "% of residual var.)"),
         y = paste0("PC2 (", pct_ev[2], "% of residual var.)")) +
    theme_bw(base_size = 10) +
    theme(legend.position = "right", panel.grid.minor = element_blank(),
          plot.title = element_text(size = 9, face = "bold"))
}

# ── Prepare numeric matrices (same rows as sem_raw) ───────────────────────────
cover_mat_all <- sem_raw %>%
  select(all_of(cover_vars)) %>%
  mutate(across(everything(), as.numeric))

soil_mat_all  <- sem_raw %>%
  select(OM_pct, pH, TN_pct, TP_pct,
         Na_mg_g, Mg_mg_g, K_mg_g, Ca_mg_g, Texture_2mm) %>%
  mutate(across(everything(), as.numeric))

topo_mat_all  <- sem_raw %>%
  select(all_of(topo_vars_pca)) %>%
  mutate(across(everything(), as.numeric))

meta_all <- sem_raw %>% select(Location, Moraine)

partial_pca_variance_summary <- bind_rows(
  summarise_partial_rda_variance(
    cover_mat_all, meta_all, "Vegetation cover"
  ),
  summarise_partial_rda_variance(
    soil_mat_all, meta_all, "Soil"
  ),
  summarise_partial_rda_variance(
    topo_mat_all, meta_all, "Topography"
  )
)

partial_pca_variance_summary %>%
  mutate(
    across(
      c(
        total_inertia,
        conditioned_inertia,
        residual_inertia
      ),
      ~ round(.x, 3)
    ),
    across(
      c(
        variation_conditioned_by_location_pct,
        residual_variation_pct,
        PC1_pct_of_residual_variation,
        PC2_pct_of_residual_variation
      ),
      ~ round(.x, 1)
    )
  ) %>%
  rename(
    `PCA input` = PCA_input,
    `Variables (n)` = n_variables,
    `Total inertia` = total_inertia,
    `Conditioned inertia` = conditioned_inertia,
    `Conditioned by Location (%)` =
      variation_conditioned_by_location_pct,
    `Residual inertia` = residual_inertia,
    `Residual variation (%)` = residual_variation_pct,
    `PC1 (% residual)` = PC1_pct_of_residual_variation,
    `PC2 (% residual)` = PC2_pct_of_residual_variation
  ) %>%
  kable(
    caption = paste(
      "Location-conditioned partial RDA variance partitioning.",
      "PC1 and PC2 percentages are relative to residual variation"
    )
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = FALSE
  )

p_partial_pca_biplots <- make_partial_biplot(cover_mat_all, meta_all, "Cover",      cover_var_labels) +
  make_partial_biplot(soil_mat_all,  meta_all, "Soil",       soil_var_labels)  +
  make_partial_biplot(topo_mat_all,  meta_all, "Topography", topo_var_labels)  +
  plot_layout(guides = "collect", ncol = 3) &
  theme(legend.position = "bottom")

p_partial_pca_biplots
```

```{r export-partial-pca-biplots, message=FALSE, warning=FALSE}
partial_pca_fig_dir <- file.path(molecular_dir, "SEM_figures_REVISION")
dir.create(partial_pca_fig_dir, recursive = TRUE, showWarnings = FALSE)

partial_pca_png <- file.path(partial_pca_fig_dir, "partial_pca_biplots_REVISION.png")
partial_pca_svg <- file.path(partial_pca_fig_dir, "partial_pca_biplots_REVISION.svg")

ggsave(
  filename = partial_pca_png,
  plot = p_partial_pca_biplots,
  width = 13,
  height = 4.5,
  dpi = 300,
  bg = "white"
)

save_gg_svg(
  filename = partial_pca_svg,
  plot = p_partial_pca_biplots,
  width = 13,
  height = 4.5
)

partial_pca_export_index <- tibble(
  figure = "partial_pca_biplots",
  description = "Location-conditioned partial PCA biplots for cover, soil, and topography SEM inputs",
  png = partial_pca_png,
  svg = partial_pca_svg
)

write.csv(
  partial_pca_export_index,
  file.path(partial_pca_fig_dir, "Partial_PCA_biplots_index_REVISION.csv"),
  row.names = FALSE
)

partial_pca_export_index %>%
  kable(caption = "Exported partial PCA biplots") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

```{r sem-fit, message=FALSE, warning=FALSE}
# ── Assemble core SEM data ────────────────────────────────────────────────────
# TopoPC1 = first PC of Elevation, Slope, TPI, Aspect_sin/cos, Snow_1976.
# All composites and continuous env vars are z-scored for comparability.
sem_data <- sem_raw %>%
  mutate(
    VegPC1   = as.numeric(scale(pca_cover$x[, 1])),
    SoilPC1  = as.numeric(scale(pca_soil$x[, 1])),
    TopoPC1  = as.numeric(scale(pca_topo$x[, 1])),
    Dist_z   = as.numeric(scale(DistanceFromGlacier))
  )

# ── Core lavaan path model: Distance + TopoPC1 -> Veg -> Soil ────────────────
sem_model <- '
  # Tier 1: env -> vegetation colonisation
  VegPC1  ~ a1*Dist_z + a2*TopoPC1

  # Tier 2: veg -> soil + direct env effects on soil
  SoilPC1 ~ b*VegPC1 + c1*Dist_z + c2*TopoPC1

  # Indirect effects (env -> veg -> soil)
  ind_dist := a1 * b
  ind_topo := a2 * b

  # Total effects on soil
  tot_dist := a1 * b + c1
  tot_topo := a2 * b + c2
'

sem_fit <- sem(
  sem_model,
  data      = sem_data,
  estimator = "ML",
  se        = "bootstrap",
  bootstrap = 1000
)

fit_idx <- fitMeasures(sem_fit, c("chisq", "df", "pvalue", "cfi", "rmsea", "srmr"))
cat("Core model fit:\n"); print(round(fit_idx, 3))
cat("\nR-squared:\n"); print(lavInspect(sem_fit, "r2"))
```

```{r sem-tables}
# ── Shared helper: format parameterEstimates into a clean tibble ──────────────
fmt_params <- function(fit) {
  parameterEstimates(fit, standardized = TRUE, boot.ci.type = "bca.simple") %>%
    as_tibble() %>%
    filter(op %in% c("~", ":=")) %>%
    mutate(
      Type     = if_else(op == ":=", "Indirect/Total", "Direct"),
      std_beta = round(std.all, 3),
      SE       = round(se,      3),
      CI_lo    = round(ci.lower, 3),
      CI_hi    = round(ci.upper, 3),
      p_val    = round(pvalue,  4),
      Sig      = sig_stars(pvalue)
    )
}

core_labels <- c(
  Dist_z  = "Distance from glacier (z)",
  TopoPC1 = "Topo. PCA1 (z)",
  VegPC1  = "Veg. colonisation PC1",
  SoilPC1 = "Soil development PC1"
)
ind_labels_core <- c(
  ind_dist = "Distance -> Veg -> Soil (indirect)",
  ind_topo = "TopoPC1  -> Veg -> Soil (indirect)",
  tot_dist = "Distance -> Soil (total)",
  tot_topo = "TopoPC1  -> Soil (total)"
)

param_core <- fmt_params(sem_fit)

# Direct paths
direct_core <- param_core %>%
  filter(Type == "Direct") %>%
  mutate(
    Response  = dplyr::recode(rhs, !!!core_labels),
    Predictor = dplyr::recode(lhs, !!!core_labels)
  ) %>%
  select(Response, Predictor, `Beta (std)` = std_beta, SE,
         `95% CI lo` = CI_lo, `95% CI hi` = CI_hi, p = p_val, ` ` = Sig)

nv <- sum(direct_core$Response == core_labels["VegPC1"])
ns <- sum(direct_core$Response == core_labels["SoilPC1"])

direct_core %>%
  kable(caption = paste0(
    "Core SEM direct paths (std. beta; 1000-rep bootstrap BCA CI). ",
    "TopoPC1 = PC1 of Elevation, Slope, TPI, Aspect sin/cos, Snow 1976 (all z-scored)."
  )) %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE) %>%
  pack_rows("-> Veg. colonisation PC1", 1,      nv) %>%
  pack_rows("-> Soil development PC1",  nv + 1, nv + ns)

# Indirect & total effects
param_core %>%
  filter(Type == "Indirect/Total") %>%
  mutate(Effect = dplyr::recode(lhs, !!!ind_labels_core)) %>%
  select(Effect, `Beta (std)` = std_beta, SE,
         `95% CI lo` = CI_lo, `95% CI hi` = CI_hi, p = p_val, ` ` = Sig) %>%
  kable(caption = "Core SEM indirect and total effects") %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE) %>%
  pack_rows("Indirect (env -> veg -> soil)", 1, 2) %>%
  pack_rows("Total (direct + indirect)",     3, 4)
```

```{r sem-path-diagram, fig.width = 8, fig.height = 4.5}
paths <- parameterEstimates(sem_fit, standardized = TRUE) %>%
  filter(op == "~") %>%
  mutate(std_beta = round(std.all, 2), sig = pvalue < alpha)

# 4-node layout: 2 env inputs | VegPC1 | SoilPC1
nodes <- tribble(
  ~node,      ~x,    ~y,    ~label,                              ~type,
  "Dist_z",    0.00,  0.72,  "Distance\nfrom glacier (z)",        "env",
  "TopoPC1",   0.00,  0.28,  "Topography\nPC1 (z)",              "env",
  "VegPC1",    0.50,  0.50,  "Vegetation\ncolonisation\n(PC1)",   "composite",
  "SoilPC1",   1.00,  0.50,  "Soil\ndevelopment\n(PC1)",          "composite"
)

draw_sem_diagram <- function(paths_df, nodes_df, node_w = 0.30, node_h = 0.12,
                             txt_size = 3.2, edge_frac = 0.55) {
  if (!"show_label" %in% names(paths_df)) {
    paths_df$show_label <- paths_df$sig
  }
  if (!"label_sig" %in% names(paths_df)) {
    paths_df$label_sig <- paths_df$sig
  }

  edges_df <- paths_df %>%
    left_join(nodes_df %>% select(lhs = node, x0 = x, y0 = y), by = "lhs") %>%
    left_join(nodes_df %>% select(rhs = node, x1 = x, y1 = y), by = "rhs") %>%
    filter(!is.na(x0), !is.na(x1)) %>%
    mutate(
      alpha = if_else(sig, 1.0, 0.22),
      lwd   = pmax(abs(std_beta) * 2.8 + 0.3, 0.4),
      lbl_x = x0 + (x1 - x0) * edge_frac,
      lbl_y = y0 + (y1 - y0) * edge_frac,
      lbl   = if_else(show_label, paste0(std_beta, if_else(label_sig, "*", "")), "")
    )

  nfill <- c(env = "#dbeafe", composite = "#fef3c7", outcome = "#d1fae5")
  nbord <- c(env = "#2563eb", composite = "#d97706", outcome = "#059669")
  nodes_c <- nodes_df %>%
    mutate(fill_c = nfill[type], bord_c = nbord[type])

  ggplot() +
    geom_segment(
      data = edges_df %>% filter(std_beta >= 0),
      aes(x = x0, y = y0, xend = x1, yend = y1, linewidth = lwd, alpha = alpha),
      colour = "#1a9850",
      arrow = arrow(length = unit(0.2, "cm"), type = "closed"),
      show.legend = FALSE
    ) +
    geom_segment(
      data = edges_df %>% filter(std_beta < 0),
      aes(x = x0, y = y0, xend = x1, yend = y1, linewidth = lwd, alpha = alpha),
      colour = "#d73027",
      arrow = arrow(length = unit(0.2, "cm"), type = "closed"),
      show.legend = FALSE
    ) +
    geom_label(
      data = edges_df %>% filter(show_label),
      aes(x = lbl_x, y = lbl_y, label = lbl),
      size = txt_size - 0.5, label.padding = unit(0.1, "lines"),
      label.size = 0, fill = "white", colour = "grey15"
    ) +
    geom_tile(
      data = nodes_c,
      aes(x = x, y = y, fill = fill_c, colour = bord_c),
      width = node_w, height = node_h, linewidth = 0.8, show.legend = FALSE
    ) +
    geom_text(
      data = nodes_c,
      aes(x = x, y = y, label = label),
      size = txt_size, lineheight = 0.85
    ) +
    scale_fill_identity() +
    scale_colour_identity() +
    scale_linewidth_identity() +
    scale_alpha_identity() +
    labs(caption = paste0(
      "Blue = env inputs; amber = PCA composites; green = microbial outcomes.\n",
      "Green arrow = positive, red = negative path. ",
      "Faded = ns (p >= 0.05). Values = std. beta (* p < 0.05)."
    )) +
    theme_void(base_size = 11) +
    theme(plot.caption = element_text(hjust = 0.5, size = 8, colour = "grey40",
                                      lineheight = 1.3))
}

draw_sem_diagram(paths, nodes,
                 node_w = 0.32, node_h = 0.13) +
  coord_cartesian(xlim = c(-0.22, 1.22), ylim = c(0.05, 0.95), clip = "off")
```

## Extended SEM — Microbial richness as terminal outcomes

Vegetation and soil development are hypothesised to drive microbial community
richness (measured as the number of co-occurrence network nodes per plot; source:
`network_n_taxa_by_plot_REVISION.csv`, `Treatment == "all"`). The core model is
extended by adding Bacterial (16S) and Fungal (ITS) richness as terminal outcomes,
allowing both mediated (env → veg → soil → richness) and direct paths.

```{r sem-richness-setup}
# ── Join iteratively-rarefied diversity metrics to sem_data ───────────────────
# rich_raw contains: Plot, Rich_16S, Shan_16S, Rich_ITS, Shan_ITS
# All four metrics are z-scored for SEM comparability.
sem_data_rich <- sem_data %>%
  left_join(rich_raw, by = "Plot") %>%
  mutate(
    Rich16S = as.numeric(scale(Rich_16S)),
    RichITS = as.numeric(scale(Rich_ITS)),
    Shan16S = as.numeric(scale(Shan_16S)),
    ShanITS = as.numeric(scale(Shan_ITS))
  ) %>%
  filter(!is.na(Rich16S), !is.na(RichITS))

cat("n plots with diversity data:", nrow(sem_data_rich), "\n\n")
cat("Observed genus richness (999-draw mean):\n")
cat("  16S — range:", round(range(sem_data_rich$Rich_16S, na.rm = TRUE), 1),
    " | median:", round(median(sem_data_rich$Rich_16S, na.rm = TRUE), 1), "\n")
cat("  ITS — range:", round(range(sem_data_rich$Rich_ITS, na.rm = TRUE), 1),
    " | median:", round(median(sem_data_rich$Rich_ITS, na.rm = TRUE), 1), "\n\n")
cat("Shannon H' (999-draw mean):\n")
cat("  16S — range:", round(range(sem_data_rich$Shan_16S, na.rm = TRUE), 3),
    " | median:", round(median(sem_data_rich$Shan_16S, na.rm = TRUE), 3), "\n")
cat("  ITS — range:", round(range(sem_data_rich$Shan_ITS, na.rm = TRUE), 3),
    " | median:", round(median(sem_data_rich$Shan_ITS, na.rm = TRUE), 3), "\n")
```

### Microbial richness and Shannon diversity along the glacier retreat gradient — raw values

Raw plot-level microbial diversity metrics (999-rarefaction mean) plotted against
distance from glacier for bacteria (16S) and fungi (ITS). Straight linear trend
lines with 95% confidence intervals summarize direction; panel annotations report
Spearman correlations and FDR-adjusted *p*-values.

```{r microbial-distance-raw-plots, fig.width = 11, fig.height = 8}
microbial_diversity_labels <- c(
  Rich_16S = "Bacterial richness",
  Rich_ITS = "Fungal richness",
  Shan_16S = "Bacterial Shannon H'",
  Shan_ITS = "Fungal Shannon H'"
)

microbial_distance_raw <- plot_meta %>%
  select(Plot, Location, Moraine, DistanceFromGlacier) %>%
  left_join(rich_raw, by = "Plot") %>%
  pivot_longer(
    cols      = all_of(names(microbial_diversity_labels)),
    names_to  = "diversity_metric",
    values_to = "diversity_value"
  ) %>%
  mutate(
    diversity_label = microbial_diversity_labels[diversity_metric],
    diversity_label = factor(diversity_label, levels = microbial_diversity_labels),
    Moraine = factor(Moraine, levels = c("A", "B", "C")),
    Location = factor(Location)
  )

microbial_distance_sp <- microbial_distance_raw %>%
  group_by(diversity_metric, diversity_label) %>%
  group_modify(~ {
    ok <- complete.cases(.x$DistanceFromGlacier, .x$diversity_value)
    if (sum(ok) < 5) {
      return(tibble(rho = NA_real_, p_sp = NA_real_))
    }
    ct <- cor.test(
      .x$DistanceFromGlacier[ok],
      .x$diversity_value[ok],
      method = "spearman",
      exact = FALSE
    )
    tibble(rho = unname(ct$estimate), p_sp = ct$p.value)
  }) %>%
  ungroup() %>%
  mutate(
    Padj_sp = p.adjust(p_sp, method = "BH"),
    label = paste0(
      "\u03c1 = ", formatC(rho, digits = 2, format = "f"),
      sig_stars(Padj_sp),
      "\nPadj = ", fmt_p(Padj_sp)
    )
  )

p_microbial_distance_raw <- ggplot(
  microbial_distance_raw,
  aes(x = DistanceFromGlacier, y = diversity_value)
) +
  geom_smooth(
    method = "lm", formula = y ~ x,
    colour = "grey30", fill = "grey80", linewidth = 0.8, alpha = 0.3,
    na.rm = TRUE
  ) +
  geom_point(
    aes(fill = Moraine, shape = Location),
    size = 3, colour = "white", stroke = 0.6, na.rm = TRUE
  ) +
  geom_text(
    data = microbial_distance_sp,
    aes(label = label),
    x = -Inf, y = Inf,
    hjust = -0.08, vjust = 1.3,
    size = 3, colour = "grey20", inherit.aes = FALSE
  ) +
  facet_wrap(~ diversity_label, ncol = 2, scales = "free_y") +
  scale_fill_manual(
    values = moraine_colours, name = "Moraine age",
    labels = moraine_labels
  ) +
  scale_shape_manual(values = location_shapes, name = "Location") +
  guides(
    fill  = guide_legend(override.aes = list(shape = 21, size = 3.5)),
    shape = guide_legend(override.aes = list(fill = "grey60", size = 3.5))
  ) +
  labs(
    x = "Distance from glacier (m)",
    y = "Raw microbial diversity value"
  ) +
  theme_bw(base_size = 11) +
  theme(
    strip.background = element_rect(fill = "grey92"),
    strip.text       = element_text(face = "bold", size = 10),
    legend.position  = "bottom",
    legend.box       = "horizontal",
    panel.grid.minor = element_blank()
  )

p_microbial_distance_raw
```

```{r export-microbial-diversity-distance-figure, message=FALSE, warning=FALSE}
cover_fig_dir <- file.path(project_root, "Cover_figures_REVISION")
dir.create(cover_fig_dir, recursive = TRUE, showWarnings = FALSE)

microbial_distance_png <- file.path(
  cover_fig_dir,
  "microbial_diversity_distance_raw_REVISION.png"
)
microbial_distance_svg <- file.path(
  cover_fig_dir,
  "microbial_diversity_distance_raw_REVISION.svg"
)

ggsave(
  filename = microbial_distance_png,
  plot = p_microbial_distance_raw,
  width = 11,
  height = 8,
  dpi = 300,
  bg = "white"
)

save_gg_svg(
  filename = microbial_distance_svg,
  plot = p_microbial_distance_raw,
  width = 11,
  height = 8
)

microbial_export_index <- tibble(
  figure = "microbial_diversity_distance_raw",
  description = "Raw bacterial/fungal richness and Shannon diversity vs distance from glacier",
  png = microbial_distance_png,
  svg = microbial_distance_svg
)

write.csv(
  microbial_export_index,
  file.path(cover_fig_dir, "Microbial_diversity_smoothed_figures_index_REVISION.csv"),
  row.names = FALSE
)

if (exists("cover_soil_export_index")) {
  bind_rows(cover_soil_export_index, microbial_export_index) %>%
    write.csv(
      file.path(cover_fig_dir, "Cover_soil_microbe_smoothed_figures_index_REVISION.csv"),
      row.names = FALSE
    )
}

microbial_export_index %>%
  kable(caption = "Exported smoothed microbial diversity figure") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

```{r sem-richness-fit, message=FALSE, warning=FALSE}
sem_model_rich <- '
  # Tier 1: env -> vegetation colonisation
  VegPC1  ~ a1*Dist_z + a2*TopoPC1

  # Tier 2: veg -> soil + direct env effects on soil
  SoilPC1 ~ b*VegPC1 + c1*Dist_z + c2*TopoPC1

  # Tier 3: microbial richness as terminal outcomes
  # Parsimonious model: environment affects richness through vegetation/soil.
  Rich16S ~ d1*VegPC1 + d2*SoilPC1
  RichITS ~ f1*VegPC1 + f2*SoilPC1

  # Correlated residuals — 16S and ITS share unmeasured environmental drivers
  Rich16S ~~ RichITS

  # ── Key indirect effects ──────────────────────────────────────────────────
  # Env -> Soil via Veg
  ind_dist_veg_soil  := a1 * b
  ind_topo_veg_soil  := a2 * b

  # Veg -> richness (direct)  vs via Soil
  ind_veg_soil_16s   := b  * d2
  ind_veg_soil_its   := b  * f2

  # Env -> 16S via Veg and soil-mediated pathways
  ind_dist_veg_16s   := a1 * d1
  ind_dist_soil_16s  := (a1*b + c1) * d2
  ind_topo_veg_16s   := a2 * d1
  ind_topo_soil_16s  := (a2*b + c2) * d2

  # Env -> ITS via Veg and soil-mediated pathways
  ind_dist_veg_its   := a1 * f1
  ind_dist_soil_its  := (a1*b + c1) * f2
  ind_topo_veg_its   := a2 * f1
  ind_topo_soil_its  := (a2*b + c2) * f2
'

sem_rich <- sem(
  sem_model_rich,
  data      = sem_data_rich,
  estimator = "ML",
  se        = "bootstrap",
  bootstrap = 1000
)

fit_rich <- fitMeasures(sem_rich, c("chisq", "df", "pvalue", "cfi", "rmsea", "srmr"))
cat("Extended model fit:\n"); print(round(fit_rich, 3))
cat("\nR-squared:\n"); print(lavInspect(sem_rich, "r2"))
```

```{r sem-richness-tables}
rich_node_labels <- c(
  Dist_z   = "Distance (z)",
  TopoPC1  = "TopoPC1 (z)",
  VegPC1   = "Veg. colonisation PC1",
  SoilPC1  = "Soil development PC1",
  Rich16S  = "Bacterial richness (z)",
  RichITS  = "Fungal richness (z)"
)

param_rich <- fmt_params(sem_rich)

# Direct paths table
direct_rich <- param_rich %>%
  filter(Type == "Direct") %>%
  mutate(
    Response  = dplyr::recode(rhs, !!!rich_node_labels),
    Predictor = dplyr::recode(lhs, !!!rich_node_labels)
  ) %>%
  select(Response, Predictor, `Beta (std)` = std_beta, SE,
         `95% CI lo` = CI_lo, `95% CI hi` = CI_hi, p = p_val, ` ` = Sig)

row_groups <- direct_rich %>%
  mutate(.rn = row_number()) %>%
  group_by(Response) %>%
  summarise(start = min(.rn), end = max(.rn), .groups = "drop") %>%
  arrange(start)

tbl <- direct_rich %>%
  kable(caption = paste0(
    "Extended SEM direct paths. Bacterial/Fungal richness = network node count ",
    "(z-scored). Bootstrap SE & 95% BCA CI (1000 reps)."
  )) %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE)

for (i in seq_len(nrow(row_groups))) {
  tbl <- tbl %>% pack_rows(
    paste0("-> ", row_groups$Response[i]),
    row_groups$start[i], row_groups$end[i]
  )
}
tbl

# Indirect effects table
ind_labels_rich <- c(
  ind_dist_veg_soil  = "Distance -> Veg -> Soil",
  ind_topo_veg_soil  = "TopoPC1  -> Veg -> Soil",
  ind_veg_soil_16s   = "Veg -> Soil -> Bacteria",
  ind_veg_soil_its   = "Veg -> Soil -> Fungi",
  ind_dist_veg_16s   = "Distance -> Veg -> Bacteria",
  ind_dist_soil_16s  = "Distance -> Soil -> Bacteria",
  ind_topo_veg_16s   = "TopoPC1  -> Veg -> Bacteria",
  ind_topo_soil_16s  = "TopoPC1  -> Soil -> Bacteria",
  ind_dist_veg_its   = "Distance -> Veg -> Fungi",
  ind_dist_soil_its  = "Distance -> Soil -> Fungi",
  ind_topo_veg_its   = "TopoPC1  -> Veg -> Fungi",
  ind_topo_soil_its  = "TopoPC1  -> Soil -> Fungi"
)

param_rich %>%
  filter(Type == "Indirect/Total") %>%
  mutate(Effect = dplyr::recode(lhs, !!!ind_labels_rich)) %>%
  select(Effect, `Beta (std)` = std_beta, SE,
         `95% CI lo` = CI_lo, `95% CI hi` = CI_hi, p = p_val, ` ` = Sig) %>%
  kable(caption = "Extended SEM: key indirect effects") %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE) %>%
  pack_rows("Env -> Soil via Veg",          1,  2) %>%
  pack_rows("Veg -> Richness via Soil",      3,  4) %>%
  pack_rows("Env -> Bacteria (mediated)",    5,  8) %>%
  pack_rows("Env -> Fungi (mediated)",       9, 12)
```

```{r sem-richness-diagram, fig.width = 10, fig.height = 5.5}
paths_rich <- parameterEstimates(sem_rich, standardized = TRUE) %>%
  filter(op == "~") %>%
  mutate(std_beta = round(std.all, 2), sig = pvalue < alpha)

# 6-node layout: env left | composites centre | richness right
nodes_rich <- tribble(
  ~node,     ~x,    ~y,    ~label,                             ~type,
  "Dist_z",   0.00,  0.72,  "Distance\nfrom glacier (z)",       "env",
  "TopoPC1",  0.00,  0.28,  "Topography\nPC1 (z)",             "env",
  "VegPC1",   0.40,  0.72,  "Vegetation\ncolonisation\n(PC1)", "composite",
  "SoilPC1",  0.40,  0.28,  "Soil\ndevelopment\n(PC1)",        "composite",
  "Rich16S",  0.85,  0.72,  "Bacterial\nrichness",             "outcome",
  "RichITS",  0.85,  0.28,  "Fungal\nrichness",                "outcome"
)

draw_sem_diagram(paths_rich, nodes_rich,
                 node_w = 0.30, node_h = 0.12, txt_size = 3.0, edge_frac = 0.52) +
  coord_cartesian(xlim = c(-0.22, 1.08), ylim = c(0.05, 0.95), clip = "off")
```

```{r sem-richness-bacterial-trimmed-diagram, fig.width = 9, fig.height = 4.8}
# Main trimmed microbial SEM: bacterial richness as the terminal outcome.
# This focuses the figure on the hypothesised environment -> vegetation -> soil
# -> bacterial richness pathway, while retaining direct env -> soil paths.
paths_rich_bact <- paths_rich %>%
  filter(.data$lhs %in% c("VegPC1", "SoilPC1", "Rich16S"))

nodes_rich_bact <- tribble(
  ~node,     ~x,    ~y,    ~label,                             ~type,
  "Dist_z",   0.00,  0.72,  "Distance\nfrom glacier (z)",       "env",
  "TopoPC1",  0.00,  0.28,  "Topography\nPC1 (z)",             "env",
  "VegPC1",   0.40,  0.72,  "Vegetation\ncolonisation\n(PC1)", "composite",
  "SoilPC1",  0.40,  0.28,  "Soil\ndevelopment\n(PC1)",        "composite",
  "Rich16S",  0.85,  0.50,  "Bacterial\nrichness",             "outcome"
)

draw_sem_diagram(paths_rich_bact, nodes_rich_bact,
                 node_w = 0.30, node_h = 0.12,
                 txt_size = 3.0, edge_frac = 0.52) +
  coord_cartesian(xlim = c(-0.22, 1.08), ylim = c(0.05, 0.95), clip = "off") +
  labs(subtitle = "Trimmed SEM — observed bacterial genus richness")
```

## Extended SEM — Shannon diversity as terminal outcomes

The same model structure as above but using **Shannon H'** (evenness-weighted
diversity) in place of observed genus richness.  Shannon responds differently
from raw richness when dominant genera shift along the succession — compare the
two to distinguish richness gain from evenness restructuring.

```{r sem-shannon-fit, message=FALSE, warning=FALSE}
sem_model_shan <- '
  # Tier 1: env -> vegetation colonisation
  VegPC1  ~ a1*Dist_z + a2*TopoPC1

  # Tier 2: veg -> soil + direct env effects
  SoilPC1 ~ b*VegPC1 + c1*Dist_z + c2*TopoPC1

  # Tier 3: Shannon diversity as terminal outcomes
  # Parsimonious model: environment affects diversity through vegetation/soil.
  Shan16S ~ d1*VegPC1 + d2*SoilPC1
  ShanITS ~ f1*VegPC1 + f2*SoilPC1

  # Correlated residuals
  Shan16S ~~ ShanITS

  # Indirect effects
  ind_dist_veg_soil  := a1 * b
  ind_topo_veg_soil  := a2 * b
  ind_veg_soil_16s   := b  * d2
  ind_veg_soil_its   := b  * f2
  ind_dist_veg_16s   := a1 * d1
  ind_dist_soil_16s  := (a1*b + c1) * d2
  ind_topo_veg_16s   := a2 * d1
  ind_topo_soil_16s  := (a2*b + c2) * d2
  ind_dist_veg_its   := a1 * f1
  ind_dist_soil_its  := (a1*b + c1) * f2
  ind_topo_veg_its   := a2 * f1
  ind_topo_soil_its  := (a2*b + c2) * f2
'

sem_shan <- sem(
  sem_model_shan,
  data      = sem_data_rich,
  estimator = "ML",
  se        = "bootstrap",
  bootstrap = 1000
)

fit_shan <- fitMeasures(sem_shan, c("chisq", "df", "pvalue", "cfi", "rmsea", "srmr"))
cat("Shannon model fit:\n"); print(round(fit_shan, 3))
cat("\nR-squared:\n"); print(lavInspect(sem_shan, "r2"))
```

```{r sem-shannon-tables}
shan_node_labels <- c(
  Dist_z  = "Distance (z)",
  TopoPC1 = "TopoPC1 (z)",
  VegPC1  = "Veg. colonisation PC1",
  SoilPC1 = "Soil development PC1",
  Shan16S = "Bacterial Shannon H' (z)",
  ShanITS = "Fungal Shannon H' (z)"
)

ind_labels_shan <- c(
  ind_dist_veg_soil  = "Distance -> Veg -> Soil",
  ind_topo_veg_soil  = "TopoPC1  -> Veg -> Soil",
  ind_veg_soil_16s   = "Veg -> Soil -> Bacteria H'",
  ind_veg_soil_its   = "Veg -> Soil -> Fungi H'",
  ind_dist_veg_16s   = "Distance -> Veg -> Bacteria H'",
  ind_dist_soil_16s  = "Distance -> Soil -> Bacteria H'",
  ind_topo_veg_16s   = "TopoPC1  -> Veg -> Bacteria H'",
  ind_topo_soil_16s  = "TopoPC1  -> Soil -> Bacteria H'",
  ind_dist_veg_its   = "Distance -> Veg -> Fungi H'",
  ind_dist_soil_its  = "Distance -> Soil -> Fungi H'",
  ind_topo_veg_its   = "TopoPC1  -> Veg -> Fungi H'",
  ind_topo_soil_its  = "TopoPC1  -> Soil -> Fungi H'"
)

param_shan <- fmt_params(sem_shan)

# Direct paths
direct_shan <- param_shan %>%
  filter(Type == "Direct") %>%
  mutate(
    Response  = dplyr::recode(rhs, !!!shan_node_labels),
    Predictor = dplyr::recode(lhs, !!!shan_node_labels)
  ) %>%
  select(Response, Predictor, `Beta (std)` = std_beta, SE,
         `95% CI lo` = CI_lo, `95% CI hi` = CI_hi, p = p_val, ` ` = Sig)

rg_shan <- direct_shan %>%
  mutate(.rn = row_number()) %>%
  group_by(Response) %>%
  summarise(start = min(.rn), end = max(.rn), .groups = "drop") %>%
  arrange(start)

tbl_shan <- direct_shan %>%
  kable(caption = paste0(
    "Shannon H' SEM direct paths (std. beta; 1000-rep bootstrap BCA CI). ",
    "Diversity metrics averaged over 999 rarefaction draws."
  )) %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE)

for (i in seq_len(nrow(rg_shan))) {
  tbl_shan <- tbl_shan %>%
    pack_rows(paste0("-> ", rg_shan$Response[i]),
              rg_shan$start[i], rg_shan$end[i])
}
tbl_shan

# Indirect effects
param_shan %>%
  filter(Type == "Indirect/Total") %>%
  mutate(Effect = dplyr::recode(lhs, !!!ind_labels_shan)) %>%
  select(Effect, `Beta (std)` = std_beta, SE,
         `95% CI lo` = CI_lo, `95% CI hi` = CI_hi, p = p_val, ` ` = Sig) %>%
  kable(caption = "Shannon H' SEM: indirect effects") %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE) %>%
  pack_rows("Env -> Soil via Veg",        1,  2) %>%
  pack_rows("Veg -> Diversity via Soil",  3,  4) %>%
  pack_rows("Env -> Bacteria H'",         5,  8) %>%
  pack_rows("Env -> Fungi H'",            9, 12)
```

```{r sem-shannon-diagram, fig.width = 10, fig.height = 5.5}
paths_shan <- parameterEstimates(sem_shan, standardized = TRUE) %>%
  filter(op == "~") %>%
  mutate(std_beta = round(std.all, 2), sig = pvalue < alpha)

nodes_shan <- tribble(
  ~node,     ~x,    ~y,    ~label,                             ~type,
  "Dist_z",   0.00,  0.72,  "Distance\nfrom glacier (z)",       "env",
  "TopoPC1",  0.00,  0.28,  "Topography\nPC1 (z)",             "env",
  "VegPC1",   0.40,  0.72,  "Vegetation\ncolonisation\n(PC1)", "composite",
  "SoilPC1",  0.40,  0.28,  "Soil\ndevelopment\n(PC1)",        "composite",
  "Shan16S",  0.85,  0.72,  "Bacterial\nShannon H'",           "outcome",
  "ShanITS",  0.85,  0.28,  "Fungal\nShannon H'",              "outcome"
)

draw_sem_diagram(paths_shan, nodes_shan,
                 node_w = 0.30, node_h = 0.12, txt_size = 3.0, edge_frac = 0.52) +
  coord_cartesian(xlim = c(-0.22, 1.08), ylim = c(0.05, 0.95), clip = "off")
```

---

## SEM sensitivity analyses

The core and richness/Shannon SEMs above use **global PCA scores** (all plots
together), so the composite axes absorb between-location variance.
Two alternative specifications are tested here:

1. **Partial PCA scores** — location centroid removed from all three composites
   before SEM; focuses on within-location succession gradients.
2. **Global PCA + Location** — global scores retained but Location is added as
   an explicit binary predictor in every equation, letting the model estimate
   how much location *per se* explains after accounting for Distance and TopoPC1.

Core model only (4-node: Dist + Topo → Veg → Soil) is used for the sensitivity
analyses to keep degrees of freedom manageable.

### Spec 1 — Partial PCA scores

```{r sem-sens-partial-setup, message=FALSE, warning=FALSE}
suppressPackageStartupMessages(library(vegan))

# ── Partial RDA site scores (residuals after removing Location means) ─────────
# scaling = 1 returns distance-biplot scores (site scores in units of
# sqrt(eigenvalue)); we z-score afterwards for comparability.
run_partial_rda <- function(mat, meta) {
  ok_cols <- apply(mat, 2, var, na.rm = TRUE) > 0
  mat     <- mat[, ok_cols, drop = FALSE]
  vegan::rda(mat ~ Condition(Location), data = meta, scale = TRUE)
}

meta_sem <- sem_raw %>% select(Location, Moraine)

rda_cover <- run_partial_rda(
  sem_raw %>% select(all_of(cover_vars)) %>% mutate(across(everything(), as.numeric)),
  meta_sem
)
rda_soil  <- run_partial_rda(
  sem_raw %>% select(OM_pct, pH, TN_pct, TP_pct,
                     Na_mg_g, Mg_mg_g, K_mg_g, Ca_mg_g, Texture_2mm) %>%
    mutate(across(everything(), as.numeric)),
  meta_sem
)
rda_topo  <- run_partial_rda(
  sem_raw %>% select(all_of(topo_vars_pca)) %>% mutate(across(everything(), as.numeric)),
  meta_sem
)

# Extract PC1 site scores; align sign with global PC1 (flip if r < 0)
align_sign <- function(partial_scores, global_scores) {
  if (cor(partial_scores, global_scores, use = "complete.obs") < 0)
    -partial_scores else partial_scores
}

pc1_cover_partial <- scores(rda_cover, display = "sites", choices = 1, scaling = 1)[, 1]
pc1_soil_partial  <- scores(rda_soil,  display = "sites", choices = 1, scaling = 1)[, 1]
pc1_topo_partial  <- scores(rda_topo,  display = "sites", choices = 1, scaling = 1)[, 1]

sem_data_partial <- sem_raw %>%
  mutate(
    VegPC1  = as.numeric(scale(align_sign(pc1_cover_partial, pca_cover$x[, 1]))),
    SoilPC1 = as.numeric(scale(align_sign(pc1_soil_partial,  pca_soil$x[, 1]))),
    TopoPC1 = as.numeric(scale(align_sign(pc1_topo_partial,  pca_topo$x[, 1]))),
    Dist_z  = as.numeric(scale(DistanceFromGlacier))
  )

cat("Partial RDA variance explained by Location:\n")
cat("  Cover:", round(RsquareAdj(rda_cover)$r.squared * 100, 1), "%\n")
cat("  Soil: ", round(RsquareAdj(rda_soil)$r.squared  * 100, 1), "%\n")
cat("  Topo: ", round(RsquareAdj(rda_topo)$r.squared  * 100, 1), "%\n")
```

```{r sem-sens-partial-fit, message=FALSE, warning=FALSE}
sem_partial <- sem(
  sem_model,          # same 4-node model string defined in sem-fit
  data      = sem_data_partial,
  estimator = "ML",
  se        = "bootstrap",
  bootstrap = 1000
)

fit_partial <- fitMeasures(sem_partial, c("chisq","df","pvalue","cfi","rmsea","srmr"))
cat("Partial PCA SEM fit:\n"); print(round(fit_partial, 3))
cat("\nR-squared:\n"); print(lavInspect(sem_partial, "r2"))
```

```{r sem-sens-partial-tables}
param_part <- fmt_params(sem_partial)

direct_part <- param_part %>%
  filter(Type == "Direct") %>%
  mutate(
    Response  = dplyr::recode(rhs, !!!core_labels),
    Predictor = dplyr::recode(lhs, !!!core_labels)
  ) %>%
  select(Response, Predictor, `Beta (std)` = std_beta, SE,
         `95% CI lo` = CI_lo, `95% CI hi` = CI_hi, p = p_val, ` ` = Sig)

nv_p <- sum(direct_part$Response == core_labels["VegPC1"])
ns_p <- sum(direct_part$Response == core_labels["SoilPC1"])

direct_part %>%
  kable(caption = paste0(
    "Spec 1 (partial PCA) — direct paths. ",
    "Composites are PC1 residuals after removing Location means (partial RDA). ",
    "Signs aligned with global PC1 direction."
  )) %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE) %>%
  pack_rows("-> Veg. colonisation PC1 (partial)", 1,         nv_p) %>%
  pack_rows("-> Soil development PC1 (partial)",  nv_p + 1,  nv_p + ns_p)

param_part %>%
  filter(Type == "Indirect/Total") %>%
  mutate(Effect = dplyr::recode(lhs, !!!ind_labels_core)) %>%
  select(Effect, `Beta (std)` = std_beta, SE,
         `95% CI lo` = CI_lo, `95% CI hi` = CI_hi, p = p_val, ` ` = Sig) %>%
  kable(caption = "Spec 1 (partial PCA) — indirect and total effects") %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE) %>%
  pack_rows("Indirect (env -> veg -> soil)", 1, 2) %>%
  pack_rows("Total (direct + indirect)",     3, 4)
```

```{r sem-sens-partial-diagram, fig.width=8, fig.height=4.5}
paths_part  <- parameterEstimates(sem_partial, standardized = TRUE) %>%
  filter(op == "~") %>%
  mutate(std_beta = round(std.all, 2), sig = pvalue < alpha)

nodes_part <- tribble(
  ~node,     ~x,    ~y,    ~label,                                    ~type,
  "Dist_z",   0.00,  0.72,  "Distance\nfrom glacier (z)",              "env",
  "TopoPC1",  0.00,  0.28,  "Topography\nPC1 — partial (z)",          "env",
  "VegPC1",   0.50,  0.72,  "Vegetation\ncolonisation\nPC1 (partial)", "composite",
  "SoilPC1",  1.00,  0.50,  "Soil\ndevelopment\nPC1 (partial)",        "composite"
)

draw_sem_diagram(paths_part, nodes_part,
                 node_w = 0.32, node_h = 0.13) +
  coord_cartesian(xlim = c(-0.22, 1.22), ylim = c(0.12, 0.88), clip = "off") +
  labs(subtitle = "Spec 1: partial PCA scores (location centroid removed)")
```

#### Spec 1 — Microbial richness (partial PCA)

```{r sem-sens-partial-rich-setup}
sem_data_partial_rich <- sem_data_partial %>%
  left_join(rich_raw, by = "Plot") %>%
  mutate(
    Rich16S = as.numeric(scale(Rich_16S)),
    RichITS = as.numeric(scale(Rich_ITS)),
    Shan16S = as.numeric(scale(Shan_16S)),
    ShanITS = as.numeric(scale(Shan_ITS))
  ) %>%
  filter(!is.na(Rich16S), !is.na(RichITS))

cat("n plots (partial PCA + richness):", nrow(sem_data_partial_rich), "\n")
```

```{r sem-sens-partial-rich-fit, message=FALSE, warning=FALSE}
# Reuse sem_model_rich — same structure, partial PCA scores as composites
sem_partial_rich <- sem(
  sem_model_rich,
  data      = sem_data_partial_rich,
  estimator = "ML",
  se        = "bootstrap",
  bootstrap = 1000
)

fit_pr <- fitMeasures(sem_partial_rich, c("chisq","df","pvalue","cfi","rmsea","srmr"))
cat("Spec 1 richness model fit:\n"); print(round(fit_pr, 3))
cat("\nR-squared:\n"); print(lavInspect(sem_partial_rich, "r2"))
```

```{r sem-sens-partial-rich-trim-fit, message=FALSE, warning=FALSE}
# Final microbial richness SEM:
# start from Spec 1 (partial PCA) and retain only direct pathways with p <= 0.10.
trim_alpha <- 0.10

partial_rich_direct_full <- parameterEstimates(sem_partial_rich, standardized = TRUE) %>%
  as_tibble() %>%
  filter(.data$op == "~") %>%
  mutate(
    keep_path = !is.na(.data$pvalue) & .data$pvalue <= trim_alpha,
    label_use = if_else(!is.na(.data$label) & nzchar(.data$label), .data$label, "")
  )

partial_rich_direct_keep <- partial_rich_direct_full %>%
  filter(.data$keep_path)

if (!nrow(partial_rich_direct_keep)) {
  stop("No Spec 1 richness direct pathways retained at p <= ", trim_alpha)
}

trimmed_direct_lines <- partial_rich_direct_keep %>%
  mutate(
    path_line = if_else(
      nzchar(.data$label_use),
      paste(.data$lhs, "~", paste0(.data$label_use, "*", .data$rhs)),
      paste(.data$lhs, "~", .data$rhs)
    )
  ) %>%
  pull(path_line)

sem_model_partial_rich_trim <- paste(
  "# Direct paths retained from full Spec 1 richness SEM at p <= 0.10",
  paste(trimmed_direct_lines, collapse = "\n"),
  "",
  "# Retain shared marker-level residual covariance",
  "Rich16S ~~ RichITS",
  sep = "\n"
)

cat("Trimmed Spec 1 richness model syntax:\n")
cat(sem_model_partial_rich_trim, "\n")

sem_partial_rich_trim <- sem(
  sem_model_partial_rich_trim,
  data      = sem_data_partial_rich,
  estimator = "ML",
  se        = "bootstrap",
  bootstrap = 1000
)

fit_pr_trim <- fitMeasures(
  sem_partial_rich_trim,
  c("chisq", "df", "pvalue", "cfi", "rmsea", "srmr")
)
cat("Trimmed Spec 1 richness model fit:\n"); print(round(fit_pr_trim, 3))
cat("\nR-squared:\n"); print(lavInspect(sem_partial_rich_trim, "r2"))
```

```{r sem-sens-partial-rich-trim-compare}
sem_compare_row <- function(fit, model_name, n_paths) {
  fm <- fitMeasures(fit, c("chisq", "df", "pvalue", "cfi", "rmsea", "srmr"))
  r2_vals <- lavInspect(fit, "r2")
  tibble(
    Model = model_name,
    `Direct paths` = n_paths,
    chisq = unname(fm["chisq"]),
    df = unname(fm["df"]),
    p = unname(fm["pvalue"]),
    CFI = unname(fm["cfi"]),
    RMSEA = unname(fm["rmsea"]),
    SRMR = unname(fm["srmr"]),
    AIC = AIC(fit),
    BIC = BIC(fit),
    `R2 bacterial richness` = unname(r2_vals["Rich16S"]),
    `R2 fungal richness` = unname(r2_vals["RichITS"])
  )
}

partial_rich_model_compare <- bind_rows(
  sem_compare_row(
    sem_partial_rich,
    "Full Spec 1 richness SEM",
    sum(partial_rich_direct_full$op == "~")
  ),
  sem_compare_row(
    sem_partial_rich_trim,
    paste0("Trimmed Spec 1 richness SEM (p <= ", trim_alpha, ")"),
    nrow(partial_rich_direct_keep)
  )
) %>%
  mutate(
    across(c(chisq, p, CFI, RMSEA, SRMR, AIC, BIC,
             `R2 bacterial richness`, `R2 fungal richness`), ~ round(.x, 3))
  )

partial_rich_model_compare %>%
  kable(
    caption = paste0(
      "Full vs trimmed Spec 1 microbial richness SEM. ",
      "Trimmed model retains direct pathways with full-model p <= ", trim_alpha, "."
    )
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

partial_rich_direct_full %>%
  transmute(
    Response = dplyr::recode(.data$lhs, !!!rich_node_labels),
    Predictor = dplyr::recode(.data$rhs, !!!rich_node_labels),
    `Full-model p` = round(.data$pvalue, 4),
    `Full-model std. beta` = round(.data$std.all, 3),
    `Retained in trimmed model` = if_else(.data$keep_path, "yes", "no")
  ) %>%
  arrange(Response, `Full-model p`) %>%
  kable(
    caption = paste0(
      "Path-retention audit for trimmed Spec 1 richness SEM ",
      "(threshold: full-model p <= ", trim_alpha, ")."
    )
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

```{r sem-sens-partial-rich-trim-tables}
param_pr_trim <- parameterEstimates(
  sem_partial_rich_trim,
  standardized = TRUE,
  boot.ci.type = "bca.simple"
) %>%
  as_tibble() %>%
  filter(.data$op == "~") %>%
  mutate(
    Response  = dplyr::recode(.data$lhs, !!!rich_node_labels),
    Predictor = dplyr::recode(.data$rhs, !!!rich_node_labels),
    `Estimate (raw)` = round(.data$est, 3),
    `Beta (std)` = round(.data$std.all, 3),
    SE = round(.data$se, 3),
    `95% CI lo` = round(.data$ci.lower, 3),
    `95% CI hi` = round(.data$ci.upper, 3),
    p = round(.data$pvalue, 4),
    ` ` = sig_stars(.data$pvalue)
  ) %>%
  select(
    Response, Predictor, `Estimate (raw)`, `Beta (std)`, SE,
    `95% CI lo`, `95% CI hi`, p, ` `
  ) %>%
  arrange(Response, Predictor)

rg_pr_trim <- param_pr_trim %>%
  mutate(.rn = row_number()) %>%
  group_by(Response) %>%
  summarise(start = min(.rn), end = max(.rn), .groups = "drop") %>%
  arrange(start)

tbl_pr_trim <- param_pr_trim %>%
  kable(
    caption = paste0(
      "Final trimmed Spec 1 richness SEM direct paths. ",
      "Figure labels are `Beta (std)` rounded to two decimals; * indicates p < 0.05."
    )
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

for (i in seq_len(nrow(rg_pr_trim))) {
  tbl_pr_trim <- tbl_pr_trim %>%
    pack_rows(paste0("-> ", rg_pr_trim$Response[i]),
              rg_pr_trim$start[i], rg_pr_trim$end[i])
}
tbl_pr_trim
```

```{r sem-sens-partial-rich-tables}
param_pr <- fmt_params(sem_partial_rich)

direct_pr <- param_pr %>%
  filter(Type == "Direct") %>%
  mutate(
    Response  = dplyr::recode(rhs, !!!rich_node_labels),
    Predictor = dplyr::recode(lhs, !!!rich_node_labels)
  ) %>%
  select(Response, Predictor, `Beta (std)` = std_beta, SE,
         `95% CI lo` = CI_lo, `95% CI hi` = CI_hi, p = p_val, ` ` = Sig)

rg_pr <- direct_pr %>%
  mutate(.rn = row_number()) %>%
  group_by(Response) %>%
  summarise(start = min(.rn), end = max(.rn), .groups = "drop") %>%
  arrange(start)

tbl_pr <- direct_pr %>%
  kable(caption = paste0(
    "Spec 1 (partial PCA) — richness SEM direct paths. ",
    "Composites are location-conditioned PC1 scores."
  )) %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE)

for (i in seq_len(nrow(rg_pr)))
  tbl_pr <- tbl_pr %>% pack_rows(paste0("-> ", rg_pr$Response[i]),
                                  rg_pr$start[i], rg_pr$end[i])
tbl_pr

param_pr %>%
  filter(Type == "Indirect/Total") %>%
  mutate(Effect = dplyr::recode(lhs, !!!ind_labels_rich)) %>%
  select(Effect, `Beta (std)` = std_beta, SE,
         `95% CI lo` = CI_lo, `95% CI hi` = CI_hi, p = p_val, ` ` = Sig) %>%
  kable(caption = "Spec 1 richness SEM — indirect effects") %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE) %>%
  pack_rows("Env -> Soil via Veg",        1,  2) %>%
  pack_rows("Veg -> Richness via Soil",   3,  4) %>%
  pack_rows("Env -> Bacteria (mediated)", 5,  8) %>%
  pack_rows("Env -> Fungi (mediated)",    9, 12)
```

```{r sem-sens-partial-rich-diagram, fig.width=10, fig.height=5.5}
paths_pr <- parameterEstimates(sem_partial_rich, standardized = TRUE) %>%
  filter(op == "~") %>%
  mutate(std_beta = round(std.all, 2), sig = pvalue < alpha)

nodes_pr <- tribble(
  ~node,     ~x,    ~y,    ~label,                              ~type,
  "Dist_z",   0.00,  0.72,  "Distance\nfrom glacier (z)",        "env",
  "TopoPC1",  0.00,  0.28,  "Topography\nPC1 — partial (z)",    "env",
  "VegPC1",   0.40,  0.72,  "Vegetation\ncolonisation\n(PC1)",  "composite",
  "SoilPC1",  0.40,  0.28,  "Soil\ndevelopment\n(PC1)",         "composite",
  "Rich16S",  0.85,  0.72,  "Bacterial\nrichness",              "outcome",
  "RichITS",  0.85,  0.28,  "Fungal\nrichness",                 "outcome"
)

draw_sem_diagram(paths_pr, nodes_pr,
                 node_w = 0.30, node_h = 0.12, txt_size = 3.0, edge_frac = 0.52) +
  coord_cartesian(xlim = c(-0.22, 1.08), ylim = c(0.05, 0.95), clip = "off") +
  labs(subtitle = "Spec 1 (partial PCA) — observed genus richness")
```

```{r sem-sens-partial-rich-trim-diagram, fig.width=10, fig.height=5.5}
paths_pr_trim <- parameterEstimates(sem_partial_rich_trim, standardized = TRUE) %>%
  as_tibble() %>%
  filter(.data$op == "~") %>%
  mutate(
    std_beta = round(.data$std.all, 2),
    sig = .data$pvalue < alpha,
    show_label = TRUE,
    label_sig = .data$pvalue < alpha
  )

nodes_pr_trim <- nodes_pr

draw_sem_diagram(paths_pr_trim, nodes_pr_trim,
                 node_w = 0.30, node_h = 0.12, txt_size = 3.0, edge_frac = 0.52) +
  coord_cartesian(xlim = c(-0.22, 1.08), ylim = c(0.05, 0.95), clip = "off") +
  labs(subtitle = paste0(
    "Final trimmed Spec 1 richness SEM — retained full-model paths p <= ",
    trim_alpha
  ))
```

#### Spec 1 — Shannon diversity (partial PCA)

```{r sem-sens-partial-shan-fit, message=FALSE, warning=FALSE}
sem_partial_shan <- sem(
  sem_model_shan,
  data      = sem_data_partial_rich,
  estimator = "ML",
  se        = "bootstrap",
  bootstrap = 1000
)

fit_ps <- fitMeasures(sem_partial_shan, c("chisq","df","pvalue","cfi","rmsea","srmr"))
cat("Spec 1 Shannon model fit:\n"); print(round(fit_ps, 3))
cat("\nR-squared:\n"); print(lavInspect(sem_partial_shan, "r2"))
```

```{r sem-sens-partial-shan-tables}
param_ps <- fmt_params(sem_partial_shan)

direct_ps <- param_ps %>%
  filter(Type == "Direct") %>%
  mutate(
    Response  = dplyr::recode(rhs, !!!shan_node_labels),
    Predictor = dplyr::recode(lhs, !!!shan_node_labels)
  ) %>%
  select(Response, Predictor, `Beta (std)` = std_beta, SE,
         `95% CI lo` = CI_lo, `95% CI hi` = CI_hi, p = p_val, ` ` = Sig)

rg_ps <- direct_ps %>%
  mutate(.rn = row_number()) %>%
  group_by(Response) %>%
  summarise(start = min(.rn), end = max(.rn), .groups = "drop") %>%
  arrange(start)

tbl_ps <- direct_ps %>%
  kable(caption = paste0(
    "Spec 1 (partial PCA) — Shannon H' SEM direct paths. ",
    "Composites are location-conditioned PC1 scores."
  )) %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE)

for (i in seq_len(nrow(rg_ps)))
  tbl_ps <- tbl_ps %>% pack_rows(paste0("-> ", rg_ps$Response[i]),
                                  rg_ps$start[i], rg_ps$end[i])
tbl_ps

param_ps %>%
  filter(Type == "Indirect/Total") %>%
  mutate(Effect = dplyr::recode(lhs, !!!ind_labels_shan)) %>%
  select(Effect, `Beta (std)` = std_beta, SE,
         `95% CI lo` = CI_lo, `95% CI hi` = CI_hi, p = p_val, ` ` = Sig) %>%
  kable(caption = "Spec 1 Shannon SEM — indirect effects") %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE) %>%
  pack_rows("Env -> Soil via Veg",          1,  2) %>%
  pack_rows("Veg -> Diversity via Soil",    3,  4) %>%
  pack_rows("Env -> Bacteria H'",           5,  8) %>%
  pack_rows("Env -> Fungi H'",              9, 12)
```

```{r sem-sens-partial-shan-diagram, fig.width=10, fig.height=5.5}
paths_ps <- parameterEstimates(sem_partial_shan, standardized = TRUE) %>%
  filter(op == "~") %>%
  mutate(std_beta = round(std.all, 2), sig = pvalue < alpha)

nodes_ps <- tribble(
  ~node,     ~x,    ~y,    ~label,                              ~type,
  "Dist_z",   0.00,  0.72,  "Distance\nfrom glacier (z)",        "env",
  "TopoPC1",  0.00,  0.28,  "Topography\nPC1 — partial (z)",    "env",
  "VegPC1",   0.40,  0.72,  "Vegetation\ncolonisation\n(PC1)",  "composite",
  "SoilPC1",  0.40,  0.28,  "Soil\ndevelopment\n(PC1)",         "composite",
  "Shan16S",  0.85,  0.72,  "Bacterial\nShannon H'",            "outcome",
  "ShanITS",  0.85,  0.28,  "Fungal\nShannon H'",               "outcome"
)

draw_sem_diagram(paths_ps, nodes_ps,
                 node_w = 0.30, node_h = 0.12, txt_size = 3.0, edge_frac = 0.52) +
  coord_cartesian(xlim = c(-0.22, 1.08), ylim = c(0.05, 0.95), clip = "off") +
  labs(subtitle = "Spec 1 (partial PCA) — Shannon H' diversity")
```

### Spec 2 — Global PCA + Location as predictor

```{r sem-sens-loc-setup}
# Encode Location as a binary dummy: Nubra = 1, Tsomoriri = 0
sem_data_loc <- sem_data %>%
  mutate(Loc_bin = as.numeric(Location == "Nubra"))

cat("Location dummy: Nubra =", sum(sem_data_loc$Loc_bin),
    "| Tsomoriri =", sum(sem_data_loc$Loc_bin == 0), "\n")
```

```{r sem-sens-loc-fit, message=FALSE, warning=FALSE}
sem_model_loc <- '
  # Env + Location -> vegetation
  VegPC1  ~ a1*Dist_z + a2*TopoPC1 + a3*Loc_bin

  # Veg + Env + Location -> soil
  SoilPC1 ~ b*VegPC1 + c1*Dist_z + c2*TopoPC1 + c3*Loc_bin

  # Indirect effects
  ind_dist := a1 * b
  ind_topo := a2 * b
  ind_loc  := a3 * b

  # Total effects on soil
  tot_dist := a1 * b + c1
  tot_topo := a2 * b + c2
  tot_loc  := a3 * b + c3
'

sem_loc <- sem(
  sem_model_loc,
  data      = sem_data_loc,
  estimator = "ML",
  se        = "bootstrap",
  bootstrap = 1000
)

fit_loc <- fitMeasures(sem_loc, c("chisq","df","pvalue","cfi","rmsea","srmr"))
cat("Global PCA + Location SEM fit:\n"); print(round(fit_loc, 3))
cat("\nR-squared:\n"); print(lavInspect(sem_loc, "r2"))
```

```{r sem-sens-loc-tables}
loc_labels <- c(
  Dist_z  = "Distance (z)",
  TopoPC1 = "TopoPC1 (z)",
  Loc_bin = "Location (Nubra=1)",
  VegPC1  = "Veg. colonisation PC1",
  SoilPC1 = "Soil development PC1"
)
ind_labels_loc <- c(
  ind_dist = "Distance -> Veg -> Soil (indirect)",
  ind_topo = "TopoPC1  -> Veg -> Soil (indirect)",
  ind_loc  = "Location -> Veg -> Soil (indirect)",
  tot_dist = "Distance -> Soil (total)",
  tot_topo = "TopoPC1  -> Soil (total)",
  tot_loc  = "Location -> Soil (total)"
)

param_loc <- fmt_params(sem_loc)

direct_loc <- param_loc %>%
  filter(Type == "Direct") %>%
  mutate(
    Response  = dplyr::recode(rhs, !!!loc_labels),
    Predictor = dplyr::recode(lhs, !!!loc_labels)
  ) %>%
  select(Response, Predictor, `Beta (std)` = std_beta, SE,
         `95% CI lo` = CI_lo, `95% CI hi` = CI_hi, p = p_val, ` ` = Sig)

nv_l <- sum(direct_loc$Response == loc_labels["VegPC1"])
ns_l <- sum(direct_loc$Response == loc_labels["SoilPC1"])

direct_loc %>%
  kable(caption = paste0(
    "Spec 2 (global PCA + Location) — direct paths. ",
    "Global PCA scores; Location (Nubra=1, Tsomoriri=0) as an exogenous dummy."
  )) %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE) %>%
  pack_rows("-> Veg. colonisation PC1", 1,         nv_l) %>%
  pack_rows("-> Soil development PC1",  nv_l + 1,  nv_l + ns_l)

param_loc %>%
  filter(Type == "Indirect/Total") %>%
  mutate(Effect = dplyr::recode(lhs, !!!ind_labels_loc)) %>%
  select(Effect, `Beta (std)` = std_beta, SE,
         `95% CI lo` = CI_lo, `95% CI hi` = CI_hi, p = p_val, ` ` = Sig) %>%
  kable(caption = "Spec 2 (global PCA + Location) — indirect and total effects") %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE) %>%
  pack_rows("Indirect (env/loc -> veg -> soil)", 1, 3) %>%
  pack_rows("Total (direct + indirect)",         4, 6)
```

```{r sem-sens-loc-diagram, fig.width=8, fig.height=5}
paths_loc <- parameterEstimates(sem_loc, standardized = TRUE) %>%
  filter(op == "~") %>%
  mutate(std_beta = round(std.all, 2), sig = pvalue < alpha)

nodes_loc <- tribble(
  ~node,     ~x,    ~y,    ~label,                            ~type,
  "Dist_z",   0.00,  0.85,  "Distance\nfrom glacier (z)",      "env",
  "TopoPC1",  0.00,  0.50,  "Topography\nPC1 (z)",            "env",
  "Loc_bin",  0.00,  0.15,  "Location\n(Nubra = 1)",          "env",
  "VegPC1",   0.50,  0.68,  "Vegetation\ncolonisation\n(PC1)","composite",
  "SoilPC1",  1.00,  0.50,  "Soil\ndevelopment\n(PC1)",       "composite"
)

draw_sem_diagram(paths_loc, nodes_loc,
                 node_w = 0.30, node_h = 0.12) +
  coord_cartesian(xlim = c(-0.20, 1.20), ylim = c(0.02, 0.98), clip = "off") +
  labs(subtitle = "Spec 2: global PCA + Location as explicit predictor")
```

## Export SEM figures

```{r export-sem-figures, message=FALSE, warning=FALSE}
sem_fig_dir <- file.path(molecular_dir, "SEM_figures_REVISION")
dir.create(sem_fig_dir, recursive = TRUE, showWarnings = FALSE)

sem_figures <- list(
  core_global = list(
    plot = draw_sem_diagram(paths, nodes, node_w = 0.32, node_h = 0.13) +
      coord_cartesian(xlim = c(-0.22, 1.22), ylim = c(0.05, 0.95), clip = "off"),
    width = 8, height = 4.5
  ),
  richness_global = list(
    plot = draw_sem_diagram(paths_rich, nodes_rich,
                            node_w = 0.30, node_h = 0.12,
                            txt_size = 3.0, edge_frac = 0.52) +
      coord_cartesian(xlim = c(-0.22, 1.08), ylim = c(0.05, 0.95), clip = "off"),
    width = 10, height = 5.5
  ),
  richness_bacterial_trimmed = list(
    plot = draw_sem_diagram(paths_rich_bact, nodes_rich_bact,
                            node_w = 0.30, node_h = 0.12,
                            txt_size = 3.0, edge_frac = 0.52) +
      coord_cartesian(xlim = c(-0.22, 1.08), ylim = c(0.05, 0.95), clip = "off") +
      labs(subtitle = "Trimmed SEM — observed bacterial genus richness"),
    width = 9, height = 4.8
  ),
  shannon_global = list(
    plot = draw_sem_diagram(paths_shan, nodes_shan,
                            node_w = 0.30, node_h = 0.12,
                            txt_size = 3.0, edge_frac = 0.52) +
      coord_cartesian(xlim = c(-0.22, 1.08), ylim = c(0.05, 0.95), clip = "off"),
    width = 10, height = 5.5
  ),
  core_partial_pca = list(
    plot = draw_sem_diagram(paths_part, nodes_part,
                            node_w = 0.32, node_h = 0.13) +
      coord_cartesian(xlim = c(-0.22, 1.22), ylim = c(0.12, 0.88), clip = "off") +
      labs(subtitle = "Spec 1: partial PCA scores (location centroid removed)"),
    width = 8, height = 4.5
  ),
  richness_partial_pca = list(
    plot = draw_sem_diagram(paths_pr, nodes_pr,
                            node_w = 0.30, node_h = 0.12,
                            txt_size = 3.0, edge_frac = 0.52) +
      coord_cartesian(xlim = c(-0.22, 1.08), ylim = c(0.05, 0.95), clip = "off") +
      labs(subtitle = "Spec 1 (partial PCA) — observed genus richness"),
    width = 10, height = 5.5
  ),
  richness_partial_pca_trimmed = list(
    plot = draw_sem_diagram(paths_pr_trim, nodes_pr_trim,
                            node_w = 0.30, node_h = 0.12,
                            txt_size = 3.0, edge_frac = 0.52) +
      coord_cartesian(xlim = c(-0.22, 1.08), ylim = c(0.05, 0.95), clip = "off") +
      labs(subtitle = paste0(
        "Final trimmed Spec 1 richness SEM — retained full-model paths p <= ",
        trim_alpha
      )),
    width = 10, height = 5.5
  ),
  shannon_partial_pca = list(
    plot = draw_sem_diagram(paths_ps, nodes_ps,
                            node_w = 0.30, node_h = 0.12,
                            txt_size = 3.0, edge_frac = 0.52) +
      coord_cartesian(xlim = c(-0.22, 1.08), ylim = c(0.05, 0.95), clip = "off") +
      labs(subtitle = "Spec 1 (partial PCA) — Shannon H' diversity"),
    width = 10, height = 5.5
  ),
  core_global_location = list(
    plot = draw_sem_diagram(paths_loc, nodes_loc,
                            node_w = 0.30, node_h = 0.12) +
      coord_cartesian(xlim = c(-0.20, 1.20), ylim = c(0.02, 0.98), clip = "off") +
      labs(subtitle = "Spec 2: global PCA + Location as explicit predictor"),
    width = 8, height = 5
  )
)

export_sem_figure <- function(name, fig) {
  png_file <- file.path(sem_fig_dir, paste0(name, "_REVISION.png"))
  svg_file <- file.path(sem_fig_dir, paste0(name, "_REVISION.svg"))

  ggsave(
    filename = png_file,
    plot = fig$plot,
    width = fig$width,
    height = fig$height,
    dpi = 300,
    bg = "white"
  )

  save_gg_svg(
    filename = svg_file,
    plot = fig$plot,
    width = fig$width,
    height = fig$height
  )

  tibble(figure = name, png = png_file, svg = svg_file)
}

sem_export_index <- bind_rows(
  lapply(names(sem_figures), function(nm) export_sem_figure(nm, sem_figures[[nm]]))
)

write.csv(
  sem_export_index,
  file.path(sem_fig_dir, "SEM_figures_index_REVISION.csv"),
  row.names = FALSE
)

sem_export_index %>%
  kable(caption = "Exported SEM figures") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

---

## Results text

```{r results-helpers, include=FALSE}
# ── Inline-code helpers ───────────────────────────────────────────────────────
# Extract a standardised path coefficient from a lavaan fit
std_beta <- function(fit, lhs_v, rhs_v)
  round(parameterEstimates(fit, standardized = TRUE) %>%
          filter(lhs == lhs_v, rhs == rhs_v, op == "~") %>%
          pull(std.all), 2)

# Extract a p-value
pval <- function(fit, lhs_v, rhs_v) {
  p <- parameterEstimates(fit, standardized = TRUE) %>%
    filter(lhs == lhs_v, rhs == rhs_v, op == "~") %>%
    pull(pvalue)
  if (p < 0.001) "< 0.001" else paste0("= ", round(p, 3))
}

# Extract a defined (indirect/total) effect
def_est <- function(fit, label_v)
  round(parameterEstimates(fit, standardized = TRUE) %>%
          filter(lhs == label_v, op == ":=") %>%
          pull(std.all), 2)

def_p <- function(fit, label_v) {
  p <- parameterEstimates(fit, standardized = TRUE) %>%
    filter(lhs == label_v, op == ":=") %>%
    pull(pvalue)
  if (p < 0.001) "< 0.001" else paste0("= ", round(p, 3))
}

# R-squared for a node
r2 <- function(fit, node)
  round(unname(lavInspect(fit, "r2")[node]), 2)

# Fit index — NA-safe (just-identified models return NA for pvalue/chisq)
fi <- function(fit, index) {
  val <- suppressWarnings(unname(fitMeasures(fit, index)))
  if (length(val) == 0 || is.na(val)) return(NA_real_)
  round(val, 3)
}

# Display chi-square test safely
chisq_str <- function(fit) {
  p <- fi(fit, "pvalue")
  df_val <- fi(fit, "df")
  chisq_val <- fi(fit, "chisq")
  p_str <- if (is.na(p)) "n/a (saturated model, df = 0)" else
    if (p < 0.001) "< 0.001" else paste0("= ", p)
  paste0("chisq = ", chisq_val, ", df = ", df_val, ", p ", p_str)
}

# Fit quality label based on CFI/RMSEA (robust to NA pvalue)
fit_quality <- function(fit) {
  df_val <- fi(fit, "df")
  if (!is.na(df_val) && df_val == 0) {
    return("saturated")
  }
  cfi   <- fi(fit, "cfi")
  rmsea <- fi(fit, "rmsea")
  p     <- fi(fit, "pvalue")
  ok_p  <- is.na(p) || p >= 0.05
  ok_cfi   <- !is.na(cfi)   && cfi   >= 0.95
  ok_rmsea <- !is.na(rmsea) && rmsea <= 0.08
  if (ok_p && ok_cfi && ok_rmsea) "good"
  else if (ok_p || (ok_cfi && ok_rmsea)) "acceptable"
  else "marginal"
}

# PCA % variance for PC1
pct_var <- function(pca_obj)
  round(summary(pca_obj)$importance[2, 1] * 100, 1)

# Median rarefied diversity
med_div <- function(col) round(median(col, na.rm = TRUE), 1)
```

### SEM results

#### Composite axes

The first principal component of cover variables explained
`r pct_var(pca_cover)`% of total cover variance (VegPC1), with
positive loadings on lichen and moss cover and negative loadings on bare soil
and stone cover, representing a gradient from bare, recently-deglaciated
substrate to vegetated surfaces.
The soil development axis (SoilPC1) captured `r pct_var(pca_soil)`% of soil
variance, with positive loadings on organic matter, nitrogen, and phosphorus.
The topographic position axis (TopoPC1) explained `r pct_var(pca_topo)`% of
topographic variance.

Median rarefied bacterial genus richness across plots was
`r med_div(sem_data_rich$Rich_16S)` genera (bacterial Shannon H':
`r round(median(sem_data_rich$Shan_16S * sd(rich_raw$Shan_16S, na.rm=TRUE) + mean(rich_raw$Shan_16S, na.rm=TRUE), na.rm=TRUE), 2)`),
and median fungal genus richness was `r med_div(sem_data_rich$Rich_ITS)` genera
(richness and Shannon H' averaged across 999 rarefaction draws).

#### Core SEM — environmental drivers of vegetation and soil development

The core path model provided `r fit_quality(sem_fit)` fit to the data
(`r chisq_str(sem_fit)`;
CFI = `r fi(sem_fit,"cfi")`;
RMSEA = `r fi(sem_fit,"rmsea")`;
SRMR = `r fi(sem_fit,"srmr")`).

Distance from the glacier terminus was a significant predictor of vegetation
colonisation (VegPC1: β = `r std_beta(sem_fit,"VegPC1","Dist_z")`,
*p* `r pval(sem_fit,"VegPC1","Dist_z")`), indicating that plots further from
the active glacier front had greater vegetation cover.
The topographic position axis (TopoPC1) also contributed to vegetation
development (β = `r std_beta(sem_fit,"VegPC1","TopoPC1")`,
*p* `r pval(sem_fit,"VegPC1","TopoPC1")`).
Together, these environmental predictors explained
R² = `r r2(sem_fit,"VegPC1")` of variation in VegPC1.

Vegetation colonisation was the strongest predictor of soil development
(SoilPC1: β = `r std_beta(sem_fit,"SoilPC1","VegPC1")`,
*p* `r pval(sem_fit,"SoilPC1","VegPC1")`), and the model explained
R² = `r r2(sem_fit,"SoilPC1")` of SoilPC1 variance.
The indirect effect of distance on soil development via vegetation was
β = `r def_est(sem_fit,"ind_dist")` (*p* `r def_p(sem_fit,"ind_dist")`),
and the total effect of distance on soil was
β = `r def_est(sem_fit,"tot_dist")` (*p* `r def_p(sem_fit,"tot_dist")`).

#### Microbial richness SEM — global-score reference

The parsimonious microbial richness model showed `r fit_quality(sem_rich)` fit
(`r chisq_str(sem_rich)`;
CFI = `r fi(sem_rich,"cfi")`;
RMSEA = `r fi(sem_rich,"rmsea")`).
Direct environmental paths to richness were omitted a priori, so distance and
topography affect microbial richness only through vegetation colonisation and
soil development.

Bacterial genus richness was significantly predicted by soil development
(β = `r std_beta(sem_rich,"Rich16S","SoilPC1")`,
*p* `r pval(sem_rich,"Rich16S","SoilPC1")`), but not by vegetation colonisation
(β = `r std_beta(sem_rich,"Rich16S","VegPC1")`,
*p* `r pval(sem_rich,"Rich16S","VegPC1")`).
The model explained R² = `r r2(sem_rich,"Rich16S")` of bacterial richness
variance.
In contrast, fungal genus richness was not significantly predicted by any
direct vegetation or soil variable in the model (both *p* > 0.05;
VegPC1: β = `r std_beta(sem_rich,"RichITS","VegPC1")`,
*p* `r pval(sem_rich,"RichITS","VegPC1")`;
SoilPC1: β = `r std_beta(sem_rich,"RichITS","SoilPC1")`,
*p* `r pval(sem_rich,"RichITS","SoilPC1")`),
with R² = `r r2(sem_rich,"RichITS")`.
The indirect effect of distance on bacterial richness through vegetation and
soil was β = `r def_est(sem_rich,"ind_dist_soil_16s")`
(*p* `r def_p(sem_rich,"ind_dist_soil_16s")`).

#### Final microbial richness SEM — partial PCA, trimmed paths

The final microbial richness SEM used the partial-PCA composite scores
(location centroid removed from VegPC1, SoilPC1, and TopoPC1) and then removed
direct pathways with *p* > `r trim_alpha` from the full Spec 1 richness model.
This retained `r nrow(partial_rich_direct_keep)` of
`r nrow(partial_rich_direct_full)` direct pathways.
The trimmed model showed `r fit_quality(sem_partial_rich_trim)` fit
(`r chisq_str(sem_partial_rich_trim)`;
CFI = `r fi(sem_partial_rich_trim,"cfi")`;
RMSEA = `r fi(sem_partial_rich_trim,"rmsea")`;
SRMR = `r fi(sem_partial_rich_trim,"srmr")`).
It explained R² = `r r2(sem_partial_rich_trim,"Rich16S")` of bacterial richness
and R² = `r r2(sem_partial_rich_trim,"RichITS")` of fungal richness.

#### Shannon diversity SEM

The parsimonious Shannon diversity model showed `r fit_quality(sem_shan)` fit
(`r chisq_str(sem_shan)`;
CFI = `r fi(sem_shan,"cfi")`;
RMSEA = `r fi(sem_shan,"rmsea")`).
As in the richness model, direct environmental paths to Shannon diversity were
excluded so that topographic and distance effects were evaluated as mediated
pathways through vegetation and soil.
Bacterial Shannon H' was predicted by soil development
(β = `r std_beta(sem_shan,"Shan16S","SoilPC1")`,
*p* `r pval(sem_shan,"Shan16S","SoilPC1")`),
consistent with the richness model.
Fungal Shannon H' was evaluated through direct vegetation and soil paths, plus
the bootstrapped indirect pathways from distance and TopoPC1.
The model explained R² = `r r2(sem_shan,"Shan16S")` and
R² = `r r2(sem_shan,"ShanITS")` of bacterial and fungal Shannon H' variance,
respectively.

#### Sensitivity analyses

Model specification had little effect on the core succession pathway.
In the partial PCA specification (location centroid removed from all composites),
the vegetation → soil relationship remained
(β = `r std_beta(sem_partial,"SoilPC1","VegPC1")`,
*p* `r pval(sem_partial,"SoilPC1","VegPC1")`),
and bacterial richness was again significantly predicted by soil development
(β = `r std_beta(sem_partial_rich,"Rich16S","SoilPC1")`,
*p* `r pval(sem_partial_rich,"Rich16S","SoilPC1")`).
When location was included as an explicit binary predictor (Spec 2), it
explained additional variance in both VegPC1
(β = `r std_beta(sem_loc,"VegPC1","Loc_bin")`,
*p* `r pval(sem_loc,"VegPC1","Loc_bin")`)
and SoilPC1
(β = `r std_beta(sem_loc,"SoilPC1","Loc_bin")`,
*p* `r pval(sem_loc,"SoilPC1","Loc_bin")`),
indicating that the two glacier systems differ in absolute levels of vegetation
and soil development beyond what is captured by distance and topographic
position alone.

---

## Methods text

### Main text (for manuscript)

Cover types (lichen, moss, biological soil crust, forbs, graminoids, bare soil,
stones), soil properties (organic matter, pH, total nitrogen, total phosphorus,
Na, Mg, K, Ca, coarse fraction >2 mm), and topographic attributes (elevation,
slope, TPI, aspect as sin/cos components, snow cover in 1976 as a binary
proxy for recent deglaciation) were each reduced to a single axis via principal
component analysis (PCA; `prcomp`, centered and scaled).
The first PC of cover variables (VegPC1), soil variables (SoilPC1), and
topographic variables (TopoPC1) served as composite indices of vegetation
colonisation, soil development, and topographic position, respectively.

Microbial genus richness and Shannon diversity (H') were computed separately
for bacteria (16S rRNA gene amplicons) and fungi (ITS amplicons).
Amplicon sequence variants were agglomerated to genus level (`phyloseq::tax_glom`)
and genus names standardised (`NetCoMi::renameTaxa`).
Counts from all microhabitat samples within each plot were pooled by summation,
then rarefied to the minimum pooled read depth observed across plots.
To minimise stochastic noise inherent in a single rarefaction draw, subsampling
was repeated 999 times and both richness (observed genera) and Shannon H' were
averaged across iterations (`vegan::rrarefy`, `vegan::diversity`).

Directional relationships among environmental gradients, vegetation colonisation,
soil development, and microbial diversity were examined using path analysis
(structural equation modelling, SEM) implemented in `lavaan` v0.6
[@Rosseel2012].
The hypothesised succession pathway was:
environmental predictors (Distance from glacier terminus; TopoPC1) →
vegetation colonisation (VegPC1) →
soil development (SoilPC1) →
microbial diversity (bacterial and fungal richness; bacterial and fungal Shannon H').
Direct environmental effects on soil were retained because distance and
topographic position may influence soil development independently of vegetation.
Direct environmental effects on microbial richness and Shannon diversity were
omitted in the primary microbial SEMs, making microbial responses terminal
outcomes of the vegetation-soil pathway.
All continuous predictors were z-scored prior to modelling. The final microbial
richness SEM used partial-PCA composite scores, in which between-location
centroid shifts were removed before extracting PC1 scores, and was trimmed by
removing direct pathways with *p* > 0.10 from the full partial-PCA richness SEM.
Path coefficients were standardised; standard errors and 95% confidence intervals
were obtained by parametric bootstrap (1,000 resamples, BCa intervals).
Global model fit was evaluated using the chi-squared statistic, Comparative Fit
Index (CFI ≥ 0.95), Root Mean Square Error of Approximation (RMSEA ≤ 0.08),
and Standardised Root Mean Square Residual (SRMR ≤ 0.08).

---

### Supplementary methods

#### S1. Principal component analyses

PCA was applied separately to the cover, soil, and topography variable blocks.
All variables were centred and scaled to unit variance before decomposition
(`prcomp(scale. = TRUE)`).
Snow cover in 1976 is a binary variable (0/1 indicating presence of glacier
ice on a 1976 Landsat image); it was treated as numeric for PCA, consistent
with its role as an ordinal terrain descriptor.
The proportion of total variance explained by PC1 and the variable loadings are
reported for each block (see biplots, above).

To assess whether the dominant PCA axes were consistent across the two glacier
systems (Nubra and Tsomoriri), we conducted two supplementary ordination
analyses: (i) **stratified PCAs** — independent PCAs within each location;
(ii) **partial PCAs** — `vegan::rda(mat ~ Condition(Location))`, which removes
the between-location centroid from the variable matrix before decomposition.
The proportion of total variance attributable to Location is reported as
R² from the partial RDA.
All three sets of PC1 scores (global, stratified, partial) were visualised as
biplots and compared. Global scores were retained for the core vegetation-soil
SEM, whereas the final microbial richness SEM used partial-PCA scores to focus
on within-location chronosequence structure.

#### S2. Rarefaction and diversity estimation

Read counts were pooled across microhabitat samples (Bulk Soil, Soil Crust,
Moss, Lichen, Plant) belonging to the same plot by summation of raw counts.
The minimum pooled read depth across all plots set the rarefaction target
(reported in the `sem-setup` chunk output).
Genus-level observed richness (number of genera with count > 0) and Shannon H'
were computed on each of 999 subsampled count matrices drawn without replacement
(`vegan::rrarefy`, seed 42), then averaged across iterations to give a
stable per-plot estimate.

#### S3. SEM model equations

The core model (4 nodes) estimated the following regression equations:

$$
\text{VegPC1} \sim \beta_{a1} \cdot \text{Dist}_z + \beta_{a2} \cdot \text{TopoPC1}
$$
$$
\text{SoilPC1} \sim \beta_b \cdot \text{VegPC1} + \beta_{c1} \cdot \text{Dist}_z + \beta_{c2} \cdot \text{TopoPC1}
$$

The extended model added microbial diversity as terminal outcomes:

$$
\text{Richness}_{16S} \sim \beta_{d1} \cdot \text{VegPC1} + \beta_{d2} \cdot \text{SoilPC1}
$$

with an analogous equation for ITS richness and for Shannon H'.
Direct paths from Distance or TopoPC1 to microbial outcomes were excluded in the
primary microbial SEMs to avoid a saturated specification and to match the
a priori chronosequence hypothesis that environmental effects on microbial
diversity are mediated through vegetation colonisation and soil development.
For the final microbial richness model, the full partial-PCA richness SEM was
further trimmed by removing direct pathways with *p* > 0.10, and the trimmed
model was compared with the full partial-PCA model using AIC, BIC, global fit
indices, and outcome R².
Residuals of the two marker-specific outcomes (bacteria and fungi) were allowed
to covary freely (`~~`), reflecting unmeasured shared environmental drivers.

Indirect effects were computed as products of labelled path coefficients
(e.g., Distance → Vegetation → Soil = $\hat\beta_{a1} \times \hat\beta_b$)
and bootstrapped alongside direct effects.

#### S4. Sensitivity analyses and model specification

Two additional model specifications were fitted to assess robustness to the
treatment of between-location variance:

- **Spec 1 (partial PCA)**: PC1 scores for VegPC1, SoilPC1, and TopoPC1 were
  extracted from partial RDAs conditioning on Location, removing between-location
  centroid shifts from all composites. Signs were aligned with the corresponding
  global PC1 (negated if Pearson r < 0).
- **Spec 2 (global PCA + Location)**: Global PC1 scores were retained and
  Location (binary dummy: Nubra = 1, Tsomoriri = 0) was added as an explicit
  exogenous predictor in all regression equations.

#### S5. Parsimony testing

```{r parsimony-test, message=FALSE, warning=FALSE}
# ── AIC comparison across all core SEM specifications ────────────────────────
aic_tbl <- tibble(
  Specification = c(
    "Global PCA (core)",
    "Partial PCA (core)",
    "Global PCA + Location (core)"
  ),
  AIC  = c(AIC(sem_fit), AIC(sem_partial), AIC(sem_loc)),
  BIC  = c(BIC(sem_fit), BIC(sem_partial), BIC(sem_loc))
) %>%
  mutate(
    delta_AIC = round(AIC - min(AIC), 2),
    delta_BIC = round(BIC - min(BIC), 2),
    AIC = round(AIC, 2),
    BIC = round(BIC, 2)
  ) %>%
  arrange(AIC)

aic_tbl %>%
  kable(caption = paste0(
    "AIC and BIC for the three core SEM specifications (4-node model). ",
    "delta = difference from best-fitting model. ",
    "Models within delta AIC < 2 are considered equivalent."
  )) %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE)

# ── Likelihood-ratio test: is Location a significant addition? ────────────────
# Compare global PCA (nested in global+Location via constraint a3=c3=0)
# Note: both must be fitted by ML (estimator="ML"), which they are.
cat("\nLRT: Global PCA vs Global PCA + Location\n")
lrt_loc <- lavTestLRT(sem_fit, sem_loc)
print(lrt_loc)

# ── Trimmed model: drop non-significant direct env paths from core SEM ────────
# A parsimonious model retains only paths with p < 0.10 in the full model.
sig_paths <- parameterEstimates(sem_fit) %>%
  filter(op == "~", pvalue < 0.10) %>%
  mutate(path_str = paste0(lhs, " ~ ", rhs))

cat("\nPaths retained in trimmed model (p < 0.10):\n")
print(sig_paths$path_str)

sem_model_trim <- paste(
  "VegPC1  ~",  paste(
    sig_paths %>% filter(lhs == "VegPC1")  %>% pull(rhs), collapse = " + "
  ),
  "\nSoilPC1 ~", paste(
    sig_paths %>% filter(lhs == "SoilPC1") %>% pull(rhs), collapse = " + "
  )
)

cat("\nTrimmed model:\n", sem_model_trim, "\n")

# Guard: only fit if both equations have at least one predictor
if (all(c("VegPC1", "SoilPC1") %in% sig_paths$lhs)) {
  sem_trim <- sem(
    sem_model_trim,
    data      = sem_data,
    estimator = "ML",
    se        = "bootstrap",
    bootstrap = 1000
  )

  cat("\nTrimmed model fit:\n")
  print(round(fitMeasures(sem_trim, c("chisq","df","pvalue","cfi","rmsea","srmr")), 3))

  cat("\nLRT: Full core SEM vs trimmed model\n")
  # Trimmed is nested within full if only paths are removed (no new constraints)
  tryCatch(
    print(lavTestLRT(sem_trim, sem_fit)),
    error = function(e) cat("LRT not applicable (models not nested):", conditionMessage(e), "\n")
  )

  cat("\nAIC comparison (full vs trimmed):\n")
  print(tibble(
    Model = c("Full (core)", "Trimmed"),
    AIC   = round(c(AIC(sem_fit), AIC(sem_trim)), 2),
    BIC   = round(c(BIC(sem_fit), BIC(sem_trim)), 2)
  ))
} else {
  cat("One or both SoilPC1/VegPC1 equations have no retained predictors — trimmed model skipped.\n")
}
```
